---
title: "Full Prediction Maps"
author: "Jeff Prancevic"
date: today
format:
  html:
    code-fold: true
execute:
  warning: false
editor: visual
---

```{r setup, include=FALSE}
source("./functions.R")
```

# Overview

This notebook runs the full-island prediction pipeline for both Wrangell and Mitkof using model fits selected in earlier notebooks. All outputs are written to `predictions/temp/` first. Inspect each step visually and with AUC before calling `promote_to_stable()`.

The pipeline for each island:

1.  **Initiation probability** — `init_fit` applied to every masked pixel
2.  **Width map** — `width_fit` applied to full rstack, back-transformed from log scale, focal-smoothed
3.  **Runout probability** — initiation probabilities routed downslope using `stop_fit` / `stop_h0`; runout and width jointly smoothed along flow paths
4.  **Inundation** — runout probabilities spread laterally using the smoothed width map and log-normal decay kernel

# Load island state

```{r load-state}
mitkof <- readRDS(
  "/Users/Jeff/Dropbox/Documents/TealWaters/Data/Tongass/mitkof_rasters/mitkof_state.rds")
wrangell <- readRDS(
  "/Users/Jeff/Dropbox/Documents/TealWaters/Data/Tongass/wrangell_rasters/wrangell_state.rds")

mitkof   <- load_rstack(mitkof)
wrangell <- load_rstack(wrangell)
```

# Temporary model assignment

These will be populated automatically once the runout and inundation notebooks are complete and saving into the island state. For now, assign manually from objects fit in those notebooks.

```{r manual-model-assignment}
# --- Mitkof ---
# Cox GAM stopping model and baseline hazard from runout_survival.qmd
mitkof$stop_fit <- cox_gam_model       # coxph object
mitkof$stop_h0  <- compute_h0(mitkof$stop_fit)

# Width linear model and residual SE from inundation_width.qmd
mitkof$width_fit <- width_model        # lm object
mitkof$width_se  <- sigma(width_model)

# Mean observed channel width (m) — used to set focal smoothing window
mitkof$width_mean <- mean(nodes_df$Width_m, na.rm = TRUE)

# --- Wrangell ---
# Update once Wrangell runout/inundation models are fit.
# For now, transfer Mitkof models as a starting point.
wrangell$stop_fit  <- mitkof$stop_fit
wrangell$stop_h0   <- mitkof$stop_h0
wrangell$width_fit <- mitkof$width_fit
wrangell$width_se  <- mitkof$width_se
wrangell$width_mean <- mitkof$width_mean
```

# Prediction pipeline

## Mitkof

### Step 1 — Initiation probability

```{r mitkof-initiation}
cat("=== Mitkof: Initiation probability ===\n")
mitkof <- full_pred_map(mitkof, mitkof$init_fit)
```

Visually inspect the temp output before promoting.

```{r mitkof-initiation-check}
init_temp <- rast(temp_path(mitkof, "initiation_prop"))

plot(mitkof$rstack$hs, col = gray.colors(256), legend = FALSE,
     main = "Mitkof — Initiation proportion (temp)")
plot(init_temp, col = rev(hcl.colors(256, "RdYlGn")),
     add = TRUE, alpha = 0.4, axes = FALSE)
plot(as.polygons(mitkof$rstack$init_zone, dissolve = TRUE),
     add = TRUE, col = NA, border = "black")

cat("Initiation AUC:", round(mitkof$full_auc, 4), "\n")
```

```{r mitkof-initiation-promote, eval=FALSE}
# Run when satisfied with the temp output
promote_to_stable(mitkof, "initiation")
promote_to_stable(mitkof, "initiation_prop")
```

### Step 2 — Width map

```{r mitkof-width}
cat("=== Mitkof: Width map ===\n")

# Back-transform from log10 scale; cap at 100 m to avoid extreme kernel sizes
width_map <- 10^predict(mitkof$rstack, mitkof$width_fit)
width_map  <- ifel(width_map > 100, 100, width_map)
width_map  <- as.int(width_map)

# Focal smoothing: window sized to mean observed channel width
res_m <- res(width_map)[1]
w_pix <- 2 * round((mitkof$width_mean - 1) / (2 * res_m)) + 1
cat("  Focal window:", w_pix, "pixels (", w_pix * res_m, "m )\n")
width_smooth <- focal(width_map, w = w_pix, fun = "max",
                      na.policy = "omit", na.rm = TRUE)

p_width <- temp_path(mitkof, "width")
writeRaster(width_smooth, p_width, datatype = "INT4S", overwrite = TRUE)
write_raster_meta(p_width, mitkof, "width", mitkof$width_fit)
cat("  Width written to temp:", basename(p_width), "\n")
rm(width_map, width_smooth); gc()
```

### Step 3 — Runout probability

Load stable initiation or fall back to temp if not yet promoted.

```{r mitkof-runout}
cat("=== Mitkof: Runout probability ===\n")

init_pred <- tryCatch(
  load_stable(mitkof, "initiation"),
  error = function(e) {
    cat("  No stable initiation found — using temp\n")
    rast(temp_path(mitkof, "initiation"))
  }
)

mitkof_runout <- runout(init_pred, mitkof, mitkof$rstack, mitkof$stop_fit)
rm(init_pred); gc()
```

### Step 4 — Joint flow path smoothing

Smooths both runout probability and width together along D8 flow paths, reducing noise while preserving the valley-following structure.

```{r mitkof-smooth}
cat("=== Mitkof: Flow path smoothing ===\n")

width_temp <- rast(temp_path(mitkof, "width"))

smoothed_m <- smooth_flowpath(
  runout     = mitkof_runout,
  width      = width_temp,
  dem        = mitkof$rstack$dem,
  flowdir    = mitkof$rstack$flowdir,
  island     = mitkof,
  half_window = 17,
  threshold   = 0.3
)
rm(mitkof_runout, width_temp); gc()
```

### Step 5 — Inundation

```{r mitkof-inundation}
cat("=== Mitkof: Inundation ===\n")

mitkof_inun <- inundate(
  runout = smoothed_m$runout,
  width  = smoothed_m$width,
  se     = mitkof$width_se,
  island = mitkof
)
rm(smoothed_m); gc()
```

### Step 6 — Assess and promote

```{r mitkof-assess}
# Compute proportions against observed landslide polygons
inun_temp <- rast(temp_path(mitkof, "inundation"))
inun_prop <- suscept_prop(mitkof$rstack$ls_poly, inun_temp)

p_inun_prop <- temp_path(mitkof, "inundation_prop")
writeRaster(inun_prop, p_inun_prop, overwrite = TRUE)

auc_inun <- full_auc(mitkof$rstack$ls_poly, inun_prop)
cat("Mitkof inundation AUC:", round(auc_inun, 4), "\n")

plot(mitkof$rstack$hs, col = gray.colors(256), legend = FALSE,
     main = "Mitkof — Inundation proportion (temp)")
plot(inun_prop, col = rev(hcl.colors(256, "RdYlGn")),
     add = TRUE, alpha = 0.4, axes = FALSE)
plot(as.polygons(mitkof$rstack$ls_poly, dissolve = TRUE),
     add = TRUE, col = NA, border = "black")

rm(inun_temp, inun_prop); gc()
```

```{r mitkof-promote, eval=FALSE}
# Run when satisfied
promote_to_stable(mitkof, "runout")
promote_to_stable(mitkof, "width")
promote_to_stable(mitkof, "inundation")
promote_to_stable(mitkof, "inundation_prop")
```

------------------------------------------------------------------------

## Wrangell

### Step 1 — Initiation probability

```{r wrangell-initiation}
cat("=== Wrangell: Initiation probability ===\n")
wrangell <- full_pred_map(wrangell, wrangell$init_fit)
```

```{r wrangell-initiation-check}
init_temp_w <- rast(temp_path(wrangell, "initiation_prop"))

plot(wrangell$rstack$hs, col = gray.colors(256), legend = FALSE,
     main = "Wrangell — Initiation proportion (temp)")
plot(init_temp_w, col = rev(hcl.colors(256, "RdYlGn")),
     add = TRUE, alpha = 0.4, axes = FALSE)
plot(as.polygons(wrangell$rstack$init_zone, dissolve = TRUE),
     add = TRUE, col = NA, border = "black")

cat("Wrangell initiation AUC:", round(wrangell$full_auc, 4), "\n")
rm(init_temp_w); gc()
```

```{r wrangell-initiation-promote, eval=FALSE}
promote_to_stable(wrangell, "initiation")
promote_to_stable(wrangell, "initiation_prop")
```

### Step 2 — Width map

```{r wrangell-width}
cat("=== Wrangell: Width map ===\n")

width_map_w <- 10^predict(wrangell$rstack, wrangell$width_fit)
width_map_w  <- ifel(width_map_w > 100, 100, width_map_w)
width_map_w  <- as.int(width_map_w)

res_m_w <- res(width_map_w)[1]
w_pix_w <- 2 * round((wrangell$width_mean - 1) / (2 * res_m_w)) + 1
cat("  Focal window:", w_pix_w, "pixels (", w_pix_w * res_m_w, "m )\n")
width_smooth_w <- focal(width_map_w, w = w_pix_w, fun = "max",
                        na.policy = "omit", na.rm = TRUE)

p_width_w <- temp_path(wrangell, "width")
writeRaster(width_smooth_w, p_width_w, datatype = "INT4S", overwrite = TRUE)
write_raster_meta(p_width_w, wrangell, "width", wrangell$width_fit)
cat("  Width written to temp:", basename(p_width_w), "\n")
rm(width_map_w, width_smooth_w); gc()
```

### Step 3 — Runout probability

```{r wrangell-runout}
cat("=== Wrangell: Runout probability ===\n")

init_pred_w <- tryCatch(
  load_stable(wrangell, "initiation"),
  error = function(e) {
    cat("  No stable initiation found — using temp\n")
    rast(temp_path(wrangell, "initiation"))
  }
)

wrangell_runout <- runout(init_pred_w, wrangell, wrangell$rstack, wrangell$stop_fit)
rm(init_pred_w); gc()
```

### Step 4 — Joint flow path smoothing

```{r wrangell-smooth}
cat("=== Wrangell: Flow path smoothing ===\n")

width_temp_w <- rast(temp_path(wrangell, "width"))

smoothed_w <- smooth_flowpath(
  runout      = wrangell_runout,
  width       = width_temp_w,
  dem         = wrangell$rstack$dem,
  flowdir     = wrangell$rstack$flowdir,
  island      = wrangell,
  half_window = 17,
  threshold   = 0.3
)
rm(wrangell_runout, width_temp_w); gc()
```

### Step 5 — Inundation

```{r wrangell-inundation}
cat("=== Wrangell: Inundation ===\n")

wrangell_inun <- inundate(
  runout = smoothed_w$runout,
  width  = smoothed_w$width,
  se     = wrangell$width_se,
  island = wrangell
)
rm(smoothed_w); gc()
```

### Step 6 — Assess and promote

```{r wrangell-assess}
inun_temp_w <- rast(temp_path(wrangell, "inundation"))
inun_prop_w <- suscept_prop(wrangell$rstack$ls_poly, inun_temp_w)

p_inun_prop_w <- temp_path(wrangell, "inundation_prop")
writeRaster(inun_prop_w, p_inun_prop_w, overwrite = TRUE)

auc_inun_w <- full_auc(wrangell$rstack$ls_poly, inun_prop_w)
cat("Wrangell inundation AUC:", round(auc_inun_w, 4), "\n")

plot(wrangell$rstack$hs, col = gray.colors(256), legend = FALSE,
     main = "Wrangell — Inundation proportion (temp)")
plot(inun_prop_w, col = rev(hcl.colors(256, "RdYlGn")),
     add = TRUE, alpha = 0.4, axes = FALSE)
plot(as.polygons(wrangell$rstack$ls_poly, dissolve = TRUE),
     add = TRUE, col = NA, border = "black")

rm(inun_temp_w, inun_prop_w); gc()
```

```{r wrangell-promote, eval=FALSE}
promote_to_stable(wrangell, "runout")
promote_to_stable(wrangell, "width")
promote_to_stable(wrangell, "inundation")
promote_to_stable(wrangell, "inundation_prop")
```

------------------------------------------------------------------------

# Save updated state

Saves `stop_h0`, `width_se`, `width_mean`, `full_auc`, and any other new fields added during this session.

```{r save-state}
save_island(mitkof)
save_island(wrangell)
```
