---
title: "Full Prediction Maps"
author: "Jeff Prancevic"
date: today
format:
  html:
    code-fold: true
execute:
  warning: false
editor: visual
---

```{r setup, include=FALSE}
source("./functions.R")
```

# Overview

This notebook runs the full-island prediction pipeline for both Wrangell and Mitkof using model fits selected in earlier notebooks. All outputs are written to `predictions/temp/` first. Inspect each step visually and with AUC before calling `promote_to_stable()`.

The pipeline for each island:

1.  **Initiation probability** — `init_fit` applied to every masked pixel
2.  **Width map** — `width_fit` applied to full rstack, back-transformed from log scale, focal-smoothed
3.  **Runout probability** — initiation probabilities routed downslope using `stop_fit` / `stop_h0`; runout and width jointly smoothed along flow paths
4.  **Inundation** — runout probabilities spread laterally using the smoothed width map and log-normal decay kernel

# Load island state

```{r load-state}
mitkof <- readRDS(
  "/Users/Jeff/Dropbox/Documents/TealWaters/Data/Tongass/mitkof_rasters/mitkof_state.rds")
wrangell <- readRDS(
  "/Users/Jeff/Dropbox/Documents/TealWaters/Data/Tongass/wrangell_rasters/wrangell_state.rds")

mitkof   <- load_rstack(mitkof)
wrangell <- load_rstack(wrangell)
```

# Prediction pipeline

## Mitkof

### Step 1 — Initiation probability

```{r mitkof-initiation}
cat("=== Mitkof: Initiation probability ===\n")
mitkof <- full_pred_map(mitkof, mitkof$init_fit)
```

Visually inspect the temp output before promoting.

```{r mitkof-initiation-check}
init_temp <- rast(temp_path(mitkof, "initiation_prop"))

plot(mitkof$rstack$hs, col = gray.colors(256), legend = FALSE,
     main = "Mitkof — Initiation proportion (temp)")
plot(init_temp, col = rev(hcl.colors(256, "RdYlGn")),
     add = TRUE, alpha = 0.4, axes = FALSE)
plot(as.polygons(mitkof$rstack$init_zone, dissolve = TRUE),
     add = TRUE, col = NA, border = "black")

cat("Initiation AUC:", round(mitkof$full_auc, 4), "\n")
```

```{r mitkof-initiation-promote, eval=FALSE}
# Run when satisfied with the temp output
promote_to_stable(mitkof, "initiation")
promote_to_stable(mitkof, "initiation_prop")
```

### Step 2 — Width map

```{r mitkof-width}
cat("=== Mitkof: Width map ===\n")

# Back-transform from log10 scale; cap at 100 m to avoid extreme kernel sizes
width_map <- 10^predict(mitkof$rstack, mitkof$width_fit)
width_map  <- ifel(width_map > 100, 100, width_map)
width_map  <- as.int(width_map)

p_width <- temp_path(mitkof, "width")
writeRaster(width_map, p_width, datatype = "INT4S", overwrite = TRUE)
write_raster_meta(p_width, mitkof, "width", mitkof$width_fit)
cat("  Width written to temp:", basename(p_width), "\n")
rm(width_map); gc()
```

### Step 3 — Runout probability

Load stable initiation or fall back to temp if not yet promoted.

```{r mitkof-runout}
cat("=== Mitkof: Runout probability ===\n")

init_pred <- tryCatch(
  load_stable(mitkof, "initiation"),
  error = function(e) {
    cat("  No stable initiation found — using temp\n")
    rast(temp_path(mitkof, "initiation_buffer"))
  }
)


mitkof_runout <- runout(init_pred_buffer, mitkof, mitkof$rstack, mitkof$stop_fit)
rm(init_pred,init_pred_buffer); gc()
```

### Step 4 — Joint flow path smoothing

Smooths both runout probability and width together along D8 flow paths, reducing noise while preserving the valley-following structure.

```{r mitkof-smooth}
cat("=== Mitkof: Flow path smoothing ===\n")

width_temp <- rast(temp_path(mitkof, "width"))

smoothed_m <- smooth_flowpath(
  runout     = mitkof_runout,
  width      = width_temp,
  dem        = mitkof$rstack$dem,
  flowdir    = mitkof$rstack$flowdir,
  island     = mitkof,
  half_window = 17,
  threshold   = 0.3,
  w = TRUE
)
rm(mitkof_runout, width_temp); gc()
```

### Step 5 — Inundation

```{r mitkof-inundation}
cat("=== Mitkof: Inundation ===\n")

mitkof_inun <- inundate(
  runout = smoothed_m$runout,
  width  = smoothed_m$width,
  se     = mitkof$width_se,
  island = mitkof,
  w = TRUE
)
rm(smoothed_m); gc()
```

### Step 6 — Assess and promote

```{r mitkof-assess}
# Compute proportions against observed landslide polygons
inun_temp <- rast(temp_path(mitkof, "inundation"))
inun_prop <- suscept_prop(mitkof$rstack$ls_poly, inun_temp)

p_inun_prop <- temp_path(mitkof, "inundation_prop")
writeRaster(inun_prop, p_inun_prop, overwrite = TRUE)

auc_inun <- full_auc(mitkof$rstack$ls_poly, inun_prop)
cat("Mitkof inundation AUC:", round(auc_inun, 4), "\n")

plot(mitkof$rstack$hs, col = gray.colors(256), legend = FALSE,
     main = "Mitkof — Inundation proportion (temp)")
plot(inun_prop, col = rev(hcl.colors(256, "RdYlGn")),
     add = TRUE, alpha = 0.4, axes = FALSE)
plot(as.polygons(mitkof$rstack$ls_poly, dissolve = TRUE),
     add = TRUE, col = NA, border = "black")

rm(inun_temp, inun_prop); gc()
```

```{r mitkof-promote, eval=FALSE}
# Run when satisfied
promote_to_stable(mitkof, "runout")
promote_to_stable(mitkof, "width")
promote_to_stable(mitkof, "inundation")
promote_to_stable(mitkof, "inundation_prop")
```

------------------------------------------------------------------------

## Wrangell

### Step 1 — Initiation probability

```{r wrangell-initiation}
cat("=== Wrangell: Initiation probability ===\n")
wrangell <- full_pred_map(wrangell, wrangell$init_fit)
```

```{r wrangell-initiation-check}
init_temp_w <- rast(temp_path(wrangell, "initiation_prop"))

plot(wrangell$rstack$hs, col = gray.colors(256), legend = FALSE,
     main = "Wrangell — Initiation proportion (temp)")
plot(init_temp_w, col = rev(hcl.colors(256, "RdYlGn")),
     add = TRUE, alpha = 0.4, axes = FALSE)
plot(as.polygons(wrangell$rstack$init_zone, dissolve = TRUE),
     add = TRUE, col = NA, border = "black")

cat("Wrangell initiation AUC:", round(wrangell$full_auc, 4), "\n")
rm(init_temp_w); gc()
```

```{r wrangell-initiation-promote, eval=FALSE}
promote_to_stable(wrangell, "initiation")
promote_to_stable(wrangell, "initiation_prop")
```

### Step 2 — Width map

```{r wrangell-width}
cat("=== Wrangell: Width map ===\n")

width_map_w <- 10^predict(wrangell$rstack, wrangell$width_fit)
width_map_w  <- ifel(width_map_w > 100, 100, width_map_w)
width_map_w  <- as.int(width_map_w)


p_width_w <- temp_path(wrangell, "width")
writeRaster(width_map_w, p_width_w, datatype = "INT4S", overwrite = TRUE)
write_raster_meta(p_width_w, wrangell, "width", wrangell$width_fit)
cat("  Width written to temp:", basename(p_width_w), "\n")
rm(width_map_w); gc()
```

### Step 3 — Runout probability

```{r wrangell-runout}
cat("=== Wrangell: Runout probability ===\n")

init_pred_w <- tryCatch(
  load_stable(wrangell, "initiation"),
  error = function(e) {
    cat("  No stable initiation found — using temp\n")
    rast(temp_path(wrangell, "initiation"))
  }
)

init_pred_buffer <- buffer_streams(init_pred_w,wrangell$rstack$logaccum)

wrangell_runout <- runout(init_pred_buffer, wrangell, wrangell$rstack, wrangell$stop_fit)
rm(init_pred_w); gc()
```

### Step 4 — Joint flow path smoothing

```{r wrangell-smooth}
cat("=== Wrangell: Flow path smoothing ===\n")

width_temp_w <- rast(temp_path(wrangell, "width"))

smoothed_w <- smooth_flowpath(
  runout      = wrangell_runout,
  width       = width_temp_w,
  dem         = wrangell$rstack$dem,
  flowdir     = wrangell$rstack$flowdir,
  island      = wrangell,
  half_window = 17,
  threshold   = 0.3,
  w = TRUE
)
rm(wrangell_runout, width_temp_w); gc()
```

### Step 5 — Inundation

```{r wrangell-inundation}
cat("=== Wrangell: Inundation ===\n")

wrangell_inun <- inundate(
  runout = smoothed_w$runout,
  width  = smoothed_w$width,
  se     = wrangell$width_se,
  island = wrangell,
  w = TRUE
)
rm(smoothed_w); gc()
```

### Step 6 — Assess and promote

```{r wrangell-assess}
inun_temp_w <- rast(temp_path(wrangell, "inundation"))
inun_prop_w <- suscept_prop(wrangell$rstack$ls_poly, inun_temp_w)

p_inun_prop_w <- temp_path(wrangell, "inundation_prop_glm")
writeRaster(inun_prop_w, p_inun_prop_w, overwrite = TRUE)

auc_inun_w <- full_auc(wrangell$rstack$ls_poly, inun_prop_w)
cat("Wrangell inundation AUC:", round(auc_inun_w, 4), "\n")

plot(wrangell$rstack$hs, col = gray.colors(256), legend = FALSE,
     main = "Wrangell — Inundation proportion (temp)")
plot(inun_prop_w, col = rev(hcl.colors(256, "RdYlGn")),
     add = TRUE, alpha = 0.4, axes = FALSE)
plot(as.polygons(wrangell$rstack$ls_poly, dissolve = TRUE),
     add = TRUE, col = NA, border = "black")

rm(inun_temp_w, inun_prop_w); gc()
```

```{r wrangell-promote, eval=FALSE}
promote_to_stable(wrangell, "runout")
promote_to_stable(wrangell, "width")
promote_to_stable(wrangell, "inundation")
promote_to_stable(wrangell, "inundation_prop")
```

------------------------------------------------------------------------

# Save updated state

Saves `stop_h0`, `width_se`, `width_mean`, `full_auc`, and any other new fields added during this session.

```{r save-state}
save_island(mitkof)
save_island(wrangell)
```
