---
title: "Landslide Initiation Modeling"
author: "Jeff Prancevic"
date: today
format:
  html:
    code-fold: true
execute:
  warning: false
editor: visual
bibliography: references.bib
---

# Load training data

```{r}
source("./functions.R")

mitkof <- readRDS(
  "/Users/Jeff/Dropbox/Documents/TealWaters/Data/Tongass/mitkof_rasters/mitkof_state.rds")
wrangell <- readRDS(
  "/Users/Jeff/Dropbox/Documents/TealWaters/Data/Tongass/wrangell_rasters/wrangell_state.rds")

mitkof   <- load_rstack(mitkof)
wrangell <- load_rstack(wrangell)
```

# Model evaluation

## Model types

We evaluate three model types for landslide initiation mapping:

-   **Logistic regression (GLM)** fits a logistic curve to the binary outcome as a linear function of predictors. It assumes monotonic relationships between each predictor and landslide probability, which limits its ability to capture non-linear effects like the U-shaped relationship between aspect and stability. It has no hyperparameters beyond predictor selection.

-   **Generalized Additive Model (GAM)** extends logistic regression by fitting smooth splines to each predictor, allowing non-monotonic relationships. The number of basis functions `k` controls flexibility and can be tuned per predictor.

-   **Random Forest (RF)** uses an ensemble of decision trees and can capture complex interactions between predictors. Hyperparameters include the number of trees (`num.trees`) and minimum node size (`min.node.size`).

The figure below illustrates a logistic curve fitted to topographic gradient, providing a reference for how the GLM maps probabilities to a single predictor.

```{r glm-illustration, fig.width=5, fig.height=4}
set.seed(42)
df_ex <- wrangell$df[sample(nrow(wrangell$df), 100), ]
df_ex$label <- as.numeric(as.character(df_ex$label))
df_ex <- df_ex[order(df_ex$gradient), ]
prob  <- predi(glm_fit(label ~ gradient, df_ex), df_ex)
plot(df_ex$gradient, df_ex$label, pch = 19, col = "gray40",
     xlab = "Topographic gradient", ylab = "P(Landslide)")
lines(df_ex$gradient, prob, col = "steelblue", lwd = 2)
rm(prob, df_ex)
```

## Should we include categorical predictors?

Before evaluating predictor combinations, we assess whether polygon-based categorical predictors (`lith` and `forest`) improve model performance. Soil polygons are excluded a priori — their classifications are explicitly tied to topographic slope and add little beyond what `gradient` and `fos_*` already capture.

```{r categorical-assessment, fig.width=12, fig.height=11}
#| fig-cap: "GLM test-slice predictions with (right) and without (left)
#|   categorical predictors, for Wrangell (top) and Mitkof (bottom)."

formula_cat  <- label ~ fos_72h + northness + norm_curv + logaccum + lith + forest
formula_base <- label ~ fos_72h + northness + norm_curv + logaccum

par(mfrow = c(2, 2))
# map_test_pred(glm_fit(formula_base, wrangell$df), wrangell)
# map_test_pred(glm_fit(formula_cat,  wrangell$df), wrangell)
map_test_pred(glm_fit(formula_base, mitkof$df),   mitkof)
map_test_pred(glm_fit(formula_cat,  mitkof$df),   mitkof)
```

In **Wrangell**, adding `lith` and `forest` improves both quantitative performance (AUC) and map quality without introducing sharp polygon artifacts. In **Mitkof**, the categorical predictors create unnatural discontinuities and reduce AUC. For all remaining analyses, `lith` and `forest` are included for Wrangell but excluded for Mitkof.

## Covariate diagnostics

With 13 quantitative covariates grouped into slope, area, curvature, and aspect families, using all simultaneously creates collinearity. We assess inter-predictor correlations, individual predictive strength, and variance inflation factors to guide predictor selection within each group.

### Mitkof

```{r cov-check-mitkof, fig.width=15, fig.height=6}
cov_check(mitkof)
```

In **Mitkof**, slope and curvature covariates are strongly correlated, leading to high VIFs when used together. We retain one predictor per group. `fos_6h` and `tangential_curv` perform best individually but all within-group predictors are considered. `area_6h` is excluded — it correlates better with slope metrics than with area. `area_72h` is a much stronger predictor than `logaccum` with an acceptably low VIF, so `logaccum` is excluded. `northness` has the highest individual coefficient and lowest VIF in the aspect group; `eastness` is excluded. Both `aspect` and `northness` are retained for non-linear models.

### Wrangell

```{r cov-check-wrangell, fig.width=15, fig.height=6}
cov_check(wrangell)
```

The Wrangell analysis mirrors Mitkof: slope and curvature covariates are correlated and `area_6h` is excluded. Unlike Mitkof, `area_72h` has a concerningly high VIF despite its predictive strength, so both `area_72h` and `logaccum` are evaluated. Raw `aspect` is a stronger predictor than on Mitkof, and all three aspect predictors (`aspect`, `northness`, `eastness`) are retained for evaluation.

## Grid search

For each island we evaluate all combinations of retained predictors (one per group), using 4-fold spatial cross-validation to measure mean AUC. For RF models we also vary `num.trees` (300, 500) and `min.node.size` (5, 10). The top three models by AUC are plotted on the test slice for qualitative review.

**Note:** the narrative descriptions below were written for a specific run of the training data sampler. Results may shift slightly between runs due to random background point selection.

### Mitkof

```{r mitkof-pred-list}
groups_m <- list(
  Slope     = c("gradient", "fos_72h", "fos_6h"),
  Area      = "area_72h",
  Curvature = c("mean_curv", "tangential_curv", "norm_curv", "profile_curv"),
  Aspect    = c("aspect", "northness")
)
pred_list_m <- expand.grid(groups_m, stringsAsFactors = FALSE) |>
  pmap(~ list(c(..1, ..2, ..3, ..4))) |>
  list_flatten()
```

#### GLM

```{r mitkof-glm, cache=TRUE}
mitkof$grid_glm <- grid_search_cv(mitkof, pred_list_m, "glm")
print(head(mitkof$grid_glm))
```

```{r mitkof-glm-maps, fig.width=15, fig.height=6}
par(mfrow = c(1, 3))
for (i in 1:3) {
  formula <- as.formula(paste("label ~",
    paste(strsplit(mitkof$grid_glm$predictors[i], ", ")[[1]], collapse = " + ")))
  map_test_pred(glm_fit(formula, mitkof$df), mitkof)
}
```

The top GLM models differ only in slope metric. Both factor-of-safety metrics (`fos_6h`, `fos_72h`) produce nearly identical maps; gradient yields a slightly different spatial pattern. We advance `fos_6h` (rank 1).

#### GAM

```{r mitkof-gam, cache=TRUE}
mitkof$grid_gam <- grid_search_cv(mitkof, pred_list_m, "gam")
print(head(mitkof$grid_gam))
```

```{r mitkof-gam-maps, fig.width=15, fig.height=6}
par(mfrow = c(1, 3))
for (i in 1:3) {
  predictors <- strsplit(mitkof$grid_gam$predictors[i], ", ")[[1]]
  gam_terms  <- ifelse(sapply(mitkof$test_df[predictors], is.numeric),
                       paste0("s(", predictors, ", k=4)"), predictors)
  formula    <- as.formula(paste("label ~", paste(gam_terms, collapse = " + ")))
  map_test_pred(gam_fit(formula, mitkof$df), mitkof)
}
```

GAM models outperform GLMs (higher mean AUC, lower SD). All top models use `fos_72h` and `area_72h`; they differ in curvature and aspect metrics. The model using `tangential_curv` and `aspect` (rank 3) produces the smoothest map. We advance that combination.

#### RF

```{r mitkof-rf, cache=TRUE}
mitkof$grid_rf <- grid_search_cv(mitkof, pred_list_m, "rf")
print(head(mitkof$grid_rf))
```

```{r mitkof-rf-maps, fig.width=15, fig.height=6}
par(mfrow = c(1, 3))
for (i in 1:3) {
  formula <- as.formula(paste("label ~",
    paste(strsplit(mitkof$grid_rf$predictors[i], ", ")[[1]], collapse = " + ")))
  map_test_pred(
    rf_fit(formula, mitkof$df, mitkof$grid_rf$ntrees[i], mitkof$grid_rf$min_nodes[i]),
    mitkof)
}
```

RF models have the lowest AUC and produce noisy, pixelated maps. All top models use `min.node.size = 10`, `area_72h`, `tangential_curv`, and `northness`. We advance rank 1 for the final comparison.

#### Best of the best — Mitkof

```{r mitkof-best, fig.width=15, fig.height=6}
par(mfrow = c(1, 3))

# GLM rank 1
formula <- as.formula(paste("label ~",
  paste(strsplit(mitkof$grid_glm$predictors[1], ", ")[[1]], collapse = " + ")))
map_test_pred(glm_fit(formula, mitkof$df), mitkof)

# GAM rank 3 (smoothest)
predictors <- strsplit(mitkof$grid_gam$predictors[3], ", ")[[1]]
gam_terms  <- paste0("s(", predictors, ", k=4)")
formula    <- as.formula(paste("label ~", paste(gam_terms, collapse = " + ")))
map_test_pred(gam_fit(formula, mitkof$df), mitkof)

# RF rank 1
formula <- as.formula(paste("label ~",
  paste(strsplit(mitkof$grid_rf$predictors[1], ", ")[[1]], collapse = " + ")))
map_test_pred(
  rf_fit(formula, mitkof$df, mitkof$grid_rf$ntrees[1], mitkof$grid_rf$min_nodes[1]),
  mitkof)
```

The GAM using `fos_72h`, `area_72h`, `tangential_curv`, and `aspect` produces the smoothest map and the best AUC. The RF map is too noisy for operational use. We select the GAM as the final Mitkof initiation model.

```{r mitkof-final-model}
mitkof$init_form <- as.formula(
  "label ~ s(fos_72h, k=4) + s(area_72h, k=4) + s(tangential_curv, k=4) + s(aspect, k=4)")
mitkof$init_type <- "gam"
mitkof$init_fit  <- gam_fit(mitkof$init_form, mitkof$df)
```

------------------------------------------------------------------------

### Wrangell

```{r wrangell-pred-list}
groups_w <- list(
  Slope     = c("gradient", "fos_72h", "fos_6h"),
  Area      = c("logaccum", "area_72h"),
  Curvature = c("mean_curv", "tangential_curv", "norm_curv", "profile_curv"),
  Aspect    = c("aspect", "northness", "eastness"),
  Lith      = "lith",
  Forest    = "forest"
)
pred_list_w <- expand.grid(groups_w, stringsAsFactors = FALSE) |>
  pmap(~ list(c(..1, ..2, ..3, ..4, ..5, ..6))) |>
  list_flatten()
```

#### GLM

```{r wrangell-glm, cache=TRUE}
wrangell$grid_glm <- grid_search_cv(wrangell, pred_list_w, "glm")
print(head(wrangell$grid_glm))
```

```{r wrangell-glm-maps, fig.width=15, fig.height=6}
par(mfrow = c(1, 3))
for (i in 1:3) {
  formula <- as.formula(paste("label ~",
    paste(strsplit(wrangell$grid_glm$predictors[i], ", ")[[1]], collapse = " + ")))
  map_test_pred(glm_fit(formula, wrangell$df), wrangell)
}
```

The best GLM models all use `gradient` and `area_72h`, differing only in curvature and aspect metrics. Maps are visually indistinguishable. We advance rank 1 (`tangential_curv`, `northness`).

#### GAM

```{r wrangell-gam, cache=TRUE}
wrangell$grid_gam <- grid_search_cv(wrangell, pred_list_w, "gam")
print(head(wrangell$grid_gam))
```

```{r wrangell-gam-maps, fig.width=15, fig.height=6}
par(mfrow = c(1, 3))
for (i in 1:3) {
  predictors <- strsplit(wrangell$grid_gam$predictors[i], ", ")[[1]]
  gam_terms  <- ifelse(sapply(wrangell$test_df[predictors], is.numeric),
                       paste0("s(", predictors, ", k=4)"), predictors)
  formula    <- as.formula(paste("label ~", paste(gam_terms, collapse = " + ")))
  map_test_pred(gam_fit(formula, wrangell$df), wrangell)
}
```

Top GAM models switch to `fos_72h` and all use `tangential_curv`. They differ in area and aspect metrics but produce indistinguishable maps. We advance rank 1 (`area_72h`, `northness`).

#### RF

```{r wrangell-rf, cache=TRUE}
wrangell$grid_rf <- grid_search_cv(wrangell, pred_list_w, "rf")
print(head(wrangell$grid_rf))
```

```{r wrangell-rf-maps, fig.width=15, fig.height=6}
par(mfrow = c(1, 3))
for (i in 1:3) {
  formula <- as.formula(paste("label ~",
    paste(strsplit(wrangell$grid_rf$predictors[i], ", ")[[1]], collapse = " + ")))
  map_test_pred(
    rf_fit(formula, wrangell$df, wrangell$grid_rf$ntrees[i], wrangell$grid_rf$min_nodes[i]),
    wrangell)
}
```

Top RF models use the same predictors (`fos_72h`, `logaccum`, `tangential_curv`, `northness`) with varying hyperparameters. Maps are noisy but less so than Mitkof. We advance rank 1.

#### Best of the best — Wrangell

```{r wrangell-best, fig.width=15, fig.height=6}
par(mfrow = c(1, 3))

# GLM rank 1
formula <- as.formula(paste("label ~",
  paste(strsplit(wrangell$grid_glm$predictors[1], ", ")[[1]], collapse = " + ")))
map_test_pred(glm_fit(formula, wrangell$df), wrangell)

# GAM rank 1
predictors <- strsplit(wrangell$grid_gam$predictors[1], ", ")[[1]]
gam_terms  <- ifelse(sapply(wrangell$test_df[predictors], is.numeric),
                     paste0("s(", predictors, ", k=4)"), predictors)
formula    <- as.formula(paste("label ~", paste(gam_terms, collapse = " + ")))
map_test_pred(gam_fit(formula, wrangell$df), wrangell)

# RF rank 1
formula <- as.formula(paste("label ~",
  paste(strsplit(wrangell$grid_rf$predictors[1], ", ")[[1]], collapse = " + ")))
map_test_pred(
  rf_fit(formula, wrangell$df, wrangell$grid_rf$ntrees[1], wrangell$grid_rf$min_nodes[1]),
  wrangell)
```

Both GLM and GAM are smooth; the RF is too noisy. The GAM has better AUC across the full training dataset, but the GLM's simplicity (no hyperparameters, directly interpretable coefficients) is appealing given that visual quality is comparable. We select the GLM as the final Wrangell initiation model.

```{r wrangell-final-model}
wrangell$init_form <- as.formula(
  "label ~ gradient + fos_72h + tangential_curv + northness + lith + forest")
wrangell$init_type <- "glm"
wrangell$init_fit  <- glm_fit(wrangell$init_form, wrangell$df)
```

# Save state

Save the updated island lists (excluding raster stacks) for use in the next notebook. Grid search results are retained for reference.

```{r save-state}
save_island(mitkof)
save_island(wrangell)
cat("State saved for Mitkof and Wrangell.\n")
cat("  mitkof$init_type:   ", mitkof$init_type,   "\n")
cat("  wrangell$init_type: ", wrangell$init_type, "\n")
```
