---
title: "inundation"
format: html
---

## Modeling runout width

Inundation can be calculated from cross-sectional area. There are a few ways to estimate cross-sectional area:

1.  estimate a flow volume and velocity for each node/pixel
2.  estimating volume and using empirical scaling relationships

It seems reasonable to me to take advantage of slope and valley geometry to estimate velocity. In any case, we don't have an easy way to model volume or velocity so estimates will be pretty crude. In the absence of volume estimates, we may just end up using a constant volume for every pixel.

```{r}
library(dplyr)
library(tidyr)
library(ggplot2)
library(terra)
source("./functions.R")

mitkof <- readRDS(
  "/Users/Jeff/Dropbox/Documents/TealWaters/Data/Tongass/mitkof_rasters/mitkof_state.rds")
wrangell <- readRDS(
  "/Users/Jeff/Dropbox/Documents/TealWaters/Data/Tongass/wrangell_rasters/wrangell_state.rds")

mitkof   <- load_rstack(mitkof)
wrangell <- load_rstack(wrangell)


covariates <- c("UpDist_m", "UpArea_m2", "gradient","log_grad",
                "tangential_curv","log_width",
                "Prof", "logaccum", "dem", "drop","start_grad","grad_roll", "tan_roll")
width_prep <- function(island){
  df <- island$nodes_df
  df <- df[df$Width_m>0,]
  df$log_width <- log10(df$Width_m)
  return(df)
}

mitkof$nodes_df <- width_prep(mitkof)
wrangell$nodes_df <- width_prep(wrangell)
```

```{r}
width_scatter <- function(island,form){
  nodes_df <- island$nodes_df
  long_df <- nodes_df %>%
    select(all_of(covariates)) %>%
    pivot_longer(-log_width, names_to = "variable", values_to = "value")
  
  cor_labels <- nodes_df %>%
    select(all_of(covariates)) %>%
    pivot_longer(-log_width, names_to = "variable", values_to = "value") %>%
    group_by(variable) %>%
    summarise(r = cor(value, log_width, use = "complete.obs"),
              label = paste0("r = ", round(r, 2)))
  
  # Compute binned averages (20 bins per covariate)
  binned_df <- long_df %>%
    group_by(variable) %>%
    mutate(bin = cut(value, breaks = 20)) %>%
    group_by(variable, bin) %>%
    summarise(value = mean(value, na.rm = TRUE),
              log_width = mean(log_width, na.rm = TRUE),
              .groups = "drop")
  
  ggplot(long_df, aes(x = value, y = log_width)) +
    geom_point(alpha = 0.2, size = 0.8) +
    geom_point(data = binned_df, aes(x = value, y = log_width),
               color = "red", size = 2) +
    geom_line(data = binned_df, aes(x = value, y = log_width),
              color = "red") +
    geom_text(data = cor_labels, aes(label = label),
              x = -Inf, y = Inf, hjust = -0.1, vjust = 1.5,
              color = "blue", size = 3) +
    facet_wrap(~variable, scales = "free_x") +
    theme_bw()
  
  width_model <- lm(form, data = nodes_df)
  print(summary(width_model))
  
  nodes_df$predicted <- 10^predict(width_model)
  
  p2 <- ggplot(nodes_df, aes(x = predicted, y = Width_m)) +
    geom_point(alpha = 0.1) +
    geom_abline(slope = 1, intercept = 0, color = "red", linetype = "dashed") +
    labs(x = "Predicted Width_m", y = "Measured Width_m",
         title = "Measured vs. Predicted") +
    theme_bw()
  print(p2)
  return(width_model)
}

mitkof$width_fit <- width_scatter(mitkof,log_width ~ logaccum + tangential_curv + log_grad+dem)
mitkof$width_se <- sigma(mitkof$width_fit)

wrangell$width_fit <- width_scatter(wrangell,log_width ~ logaccum + tangential_curv + log_grad + dem)
wrangell$width_se <- sigma(wrangell$width_fit)
```

```{r}
width_model <- lm(log_width ~ logaccum + tangential_curv + gradient+dem, data = nodes_df)
summary(width_model)

nodes_df$predicted <- 10^predict(width_model)

ggplot(nodes_df, aes(x = predicted, y = Width_m)) +
  geom_point(alpha = 0.1) +
  geom_abline(slope = 1, intercept = 0, color = "red", linetype = "dashed") +
  labs(x = "Predicted Width_m", y = "Measured Width_m",
       title = "Measured vs. Predicted") +
  theme_bw()
```

```{r}
width_map <- 10^predict(mitkof$test_stack,width_model)

se_pred <- sigma(width_model)  # residual standard error

# Predicted log width raster
log_width_map <- predict(mitkof$test_stack, width_model)

plot(width_map)
```

```{r}
hist(values(log_width_map))
```

## Modeling Inundation

To model inundation, we must expand runout probabilities across predicted widths. We can achieve this by expanding runout probabilities across a circular kernel defined by the predicted local width. We can incorporate uncertainty in the width by

```{r}
dist_m <- seq(0.1, 60, length.out = 500)
width  <- 30
se     <- 0.36

p_reach <- 1 - pnorm((log10(2 * dist_m) - log10(width)) / se)

plot(dist_m, p_reach, type = "l", lwd = 2,
     xlab = "Distance from center (m)", ylab = "P(reach)",
     main = "Log-normal width decay (width=30m, se=0.36)")
abline(v = width/2, lty = 2, col = "red")
legend("topright", legend = "width/2 = 15m", lty = 2, col = "red", bty = "n")
```

This is a computationally heavy task that R performs very slowly. We've retained the R version of the inundation function below in case useful, but ultimately we translated this function to C++ using Rcpp (the following code block).

```{r}
inundation_cox <- inundate(cox_runout,width_map,se_pred)
mitkof$ls_poly <- vect("/Users/Jeff/Dropbox/Documents/TealWaters/Data/Tongass/mitkof_shapefiles/LS_poly.shp")
mitkof$test_stack$ls_poly <- rasterize(mitkof$ls_poly,mitkof$test_stack$dem)
inun_prop_cox <- suscept_prop(mitkof$test_stack$ls_poly,inundation_cox)
plot(mitkof$test_stack$hs, col = gray.colors(256), legend = FALSE)
plot(inun_prop_cox, col = rev(hcl.colors(256, "RdYlGn")), add = TRUE, axes = FALSE, alpha = 0.35)
plot(mitkof$ls_poly, add = TRUE,legend = FALSE)
```

```{r}
inundation_test <- inundate(test_rast,width_map,se_pred)
inun_prop_test <- suscept_prop(mitkof$test_stack$ls_poly,inundation_test)
plot(mitkof$test_stack$hs, col = gray.colors(256), legend = FALSE)
plot(inun_prop_test, col = rev(hcl.colors(256, "RdYlGn")), add = TRUE, axes = FALSE, alpha = 0.35)
plot(mitkof$ls_poly, add = TRUE,legend = FALSE)
```

```{r}
inundation_cox_gam <- inundate(cox_gam_runout,width_map,se_pred)
inun_prop_cox_gam <- suscept_prop(mitkof$test_stack$ls_poly,inundation_cox_gam)
plot(mitkof$test_stack$hs, col = gray.colors(256), legend = FALSE)
plot(inun_prop_cox_gam, col = rev(hcl.colors(256, "RdYlGn")), add = TRUE, axes = FALSE, alpha = 0.35)
plot(mitkof$ls_poly, add = TRUE,legend = FALSE)
```

```{r}
inundation_logit <- inundate(logit_runout,width_map,se_pred)
inun_prop_logit <- suscept_prop(mitkof$test_stack$ls_poly,inundation_logit)
plot(mitkof$test_stack$hs, col = gray.colors(256), legend = FALSE)
plot(inun_prop_logit, col = rev(hcl.colors(256, "RdYlGn")), add = TRUE, axes = FALSE, alpha = 0.35)
plot(mitkof$ls_poly, add = TRUE,legend = FALSE)
```

# Applying to full datasets

```{r}
# runout <- rast("")

# width_map_full <- as.integer(10^predict(mitkof$rstack,width_model))
# width_vals <- as.integer(values(width_map_full))
# width_vals[!is.na(width_vals) & width_vals > 100] <- 100
# values(width_map_full) <- width_vals
# writeRaster(width_map_full,"/Users/Jeff/Dropbox/Documents/TealWaters/modeling_results/prediction_maps/mitkof_width.tif",datatype = "INT4S")
# gcrm(width_vals,width_map_full);gc()
width_map_full <- rast("/Users/Jeff/Dropbox/Documents/TealWaters/modeling_results/prediction_maps/mitkof_width.tif")
w <- 2*round((mean(nodes_df$Width_m,na.rm=TRUE)-1)/(2*res(width_map_full)[1]))+1
prob <- 1-1/w
width_map_smooth <- focal(
  width_map_full,
  w      = w,   # square window (scalar → w x w matrix of 1s)
  fun    = "max",
  na.policy = "omit"       # only compute where center cell is valid
)
writeRaster(width_map_smooth,"/Users/Jeff/Dropbox/Documents/TealWaters/modeling_results/prediction_maps/mitkof_width_smooth_max.tif")

inundation_full <- inundate(out_rast,width_map_smooth,se_pred)
```

```{r}
# mitkof$ls_poly <- vect("/Users/Jeff/Dropbox/Documents/TealWaters/Data/Tongass/mitkof_shapefiles/LS_poly.shp")
# mitkof$rstack$ls_poly <- rasterize(mitkof$ls_poly,mitkof$rstack$hs)
inun_prop_full <- suscept_prop(mitkof$rstack$ls_poly,inundation_smoothest)
writeRaster(inun_prop_full,"/Users/Jeff/Dropbox/Documents/TealWaters/modeling_results/prediction_maps/mitkof_inun_prop_smoothest_banks_masked.tif",overwrite=TRUE)
full_auc(mitkof$rstack$ls_poly,inun_prop_full)
```

```{r}
ls_poly <- vect("/Users/Jeff/Dropbox/Documents/TealWaters/Data/Tongass/mitkof_shapefiles/LS_poly.shp")
ls_poly_rast <- rasterize(ls_poly,dem)
runout_smooth <- rast("/Users/Jeff/Dropbox/Documents/TealWaters/modeling_results/prediction_maps/runout_smooth_temp.tif")
width_smooth <- rast("/Users/Jeff/Dropbox/Documents/TealWaters/modeling_results/prediction_maps/width_smooth_temp.tif")
inundate_full_smoothrunout <- inundate(runout_smooth,width_smooth,se = 0.35)
inun_prop_full <- suscept_prop(ls_poly_rast,inundate_full_smoothrunout)
writeRaster(inun_prop_full,"/Users/Jeff/Dropbox/Documents/TealWaters/modeling_results/prediction_maps/mitkof_inun_prop_smoothestrunout.tif")
full_auc(ls_poly_rast,inun_prop_full)
```

```{r}
cppFunction('
List smooth_flowpath_centered_cpp(NumericVector vals,
                                   IntegerVector width_vals,
                                   IntegerVector fdir_vals,
                                   IntegerVector sorted_idx,
                                   int nrows, int ncols,
                                   int half_window,
                                   double threshold) {
  int n = vals.size();
  
  // Build downstream lookup
  std::vector<int> downstream(n, -1);
  for (int i = 0; i < n; i++) {
    if (IntegerVector::is_na(fdir_vals[i])) continue;
    int fdir = fdir_vals[i];
    int row = i / ncols;
    int col = i % ncols;
    int j = -1;
    if      (fdir == 1)   { if (col < ncols-1)                  j = i + 1; }
    else if (fdir == 2)   { if (row < nrows-1 && col < ncols-1) j = i + ncols + 1; }
    else if (fdir == 4)   { if (row < nrows-1)                  j = i + ncols; }
    else if (fdir == 8)   { if (row < nrows-1 && col > 0)       j = i + ncols - 1; }
    else if (fdir == 16)  { if (col > 0)                        j = i - 1; }
    else if (fdir == 32)  { if (row > 0 && col > 0)             j = i - ncols - 1; }
    else if (fdir == 64)  { if (row > 0)                        j = i - ncols; }
    else if (fdir == 128) { if (row > 0 && col < ncols-1)       j = i - ncols + 1; }
    if (j >= 0 && j < n) downstream[i] = j;
  }
  
  // Build upstream lookup: highest runout probability donor
  std::vector<int> upstream(n, -1);
  for (int i = 0; i < n; i++) {
    int j = downstream[i];
    if (j < 0 || j >= n) continue;
    if (upstream[j] == -1) {
      upstream[j] = i;
    } else {
      int cur_best = upstream[j];
      bool i_better = !NumericVector::is_na(vals[i]) && 
                      (NumericVector::is_na(vals[cur_best]) || vals[i] > vals[cur_best]);
      if (i_better) upstream[j] = i;
    }
  }
  
  NumericVector result_vals  = clone(vals);
  IntegerVector result_width = clone(width_vals);
  
  for (int k = 0; k < sorted_idx.size(); k++) {
    int i = sorted_idx[k] - 1;
    if (NumericVector::is_na(vals[i]) || vals[i] < threshold) continue;
    
    double sum_v = 0.0, sum_w = 0.0;
    int count = 0;
    
    // Walk upstream
    int cur = i;
    for (int w = 0; w <= half_window; w++) {
      if (cur < 0 || cur >= n) break;
      if (!NumericVector::is_na(vals[cur])) {
        sum_v += vals[cur];
        sum_w += NumericVector::is_na(width_vals[cur]) ? 0.0 : width_vals[cur];
        count++;
      }
      cur = upstream[cur];
      if (cur < 0) break;
    }
    
    // Walk downstream (skip i)
    cur = downstream[i];
    for (int w = 0; w < half_window/2; w++) {
      if (cur < 0 || cur >= n) break;
      if (!NumericVector::is_na(vals[cur])) {
        sum_v += vals[cur];
        sum_w += NumericVector::is_na(width_vals[cur]) ? 0.0 : width_vals[cur];
        count++;
      }
      cur = downstream[cur];
    }
    
    if (count > 0) {
      result_vals[i]  = sum_v / count;
      result_width[i] = sum_w / count;
    }
  }
  
  return List::create(Named("runout") = result_vals,
                      Named("width")  = result_width);
}
')

smooth_flowpath <- function(runout, width, dem, flowdir, half_window = 10, threshold = 0.3) {
  rows <- nrow(dem)
  cols <- ncol(dem)
  sorted_idx <- as.integer(order(values(dem), decreasing = TRUE, na.last = NA))
  rm(dem);gc()
  vals       <- values(runout)[,1]
  width_vals <- as.integer(values(width)[,1])
  fdir_vals  <- as.integer(values(flowdir))
  
  out <- smooth_flowpath_centered_cpp(vals, width_vals, fdir_vals, sorted_idx,
                                      rows, cols,
                                      half_window, threshold)
  rm(vals, width_vals, fdir_vals, sorted_idx); gc()
  
  runout <- rast(runout)
  values(runout) <- out$runout
  writeRaster(runout, "/Users/Jeff/Dropbox/Documents/TealWaters/modeling_results/prediction_maps/runout_smooth_temp.tif", overwrite = TRUE)
  runout <- rast("/Users/Jeff/Dropbox/Documents/TealWaters/modeling_results/prediction_maps/runout_smooth_temp.tif")
  
  width <- rast(width)
  values(width) <- out$width
  writeRaster(width,  "/Users/Jeff/Dropbox/Documents/TealWaters/modeling_results/prediction_maps/width_smooth_temp.tif",  overwrite = TRUE)
  width  <- rast("/Users/Jeff/Dropbox/Documents/TealWaters/modeling_results/prediction_maps/width_smooth_temp.tif")
  rm(out); gc()
  
  return(list(runout = runout, width = width))
}
```

```{r}
library(terra)
library(Rcpp)
flowdir <- rast("/Users/Jeff/Dropbox/Documents/TealWaters/Data/Tongass/mitkof_rasters/flowdir.tif")
dem <- rast("/Users/Jeff/Dropbox/Documents/TealWaters/Data/Tongass/mitkof_rasters/dem_filled.tif")
width_map_full <- rast("/Users/Jeff/Dropbox/Documents/TealWaters/modeling_results/prediction_maps/mitkof_width.tif")
width_smooth <- rast("/Users/Jeff/Dropbox/Documents/TealWaters/modeling_results/prediction_maps/width_smooth_temp.tif")
runout_smooth <- rast("/Users/Jeff/Dropbox/Documents/TealWaters/modeling_results/prediction_maps/runout_smooth_temp.tif")
runout <- rast("/Users/Jeff/Dropbox/Documents/TealWaters/modeling_results/prediction_maps/mitkof_runout_cox.tif")
logaccum <- rast("/Users/Jeff/Dropbox/Documents/TealWaters/Data/Tongass/mitkof_rasters/logaccum.flt")
ls_prob <- rast("/Users/Jeff/Dropbox/Documents/TealWaters/modeling_results/prediction_maps/mitkof_gam_final3.tif")
# inundate_full <- inundate(runout,width_map_full,se = 0.35)
```

```{r}
# gc()
# inundate(runout, width_map_full, se = 0.35)
# gc()
smoothed <- smooth_flowpath(out_rast, width_map_full, dem, flowdir,
                             half_window = 17, threshold = 0.3)
gc()
inundation_smoothest <- inundate(smoothed$runout, smoothed$width, se = 0.35)
```

```{r}
stream_pts <- as.points(ifel(logaccum > 6, 1, NA), 
                         na.rm = TRUE)

# 2. Buffer the points by 50 m and dissolve into one polygon
stream_buf_poly <- aggregate(buffer(stream_pts, width = 50))

# 3. Rasterize the buffer polygon (1 inside, NA outside)
stream_buf_r <- rasterize(stream_buf_poly, logaccum, 
                           field = 1, background = NA)

# 4. Mask ls_prob: NA within the buffered stream zone
ls_prob_masked <- mask(ls_prob, stream_buf_r, maskvalues = 1)
```

```{r}
stream_r <- ifel(logaccum > 6, 1, NA)

# Expand by 50 m (the raster resolution is stored in res()); a 50 m buffer
# is achieved with a focal window of radius ceil(50 / res) pixels
pixel_res  <- res(logaccum)[1]          # e.g. 5 m
focal_r    <- ceiling(50 / pixel_res)                      # radius in pixels
win_size   <- 2 * focal_r + 1                              # odd window side length

# Any pixel within 50 m of a stream gets value 1; everything else stays NA
stream_buf <- focal(stream_r,
                    w         = win_size,
                    fun       = "max",
                    na.policy = "omit",
                    na.rm     = TRUE)

# Mask ls_prob: set to NA wherever the stream buffer is 1
ls_prob_masked <- mask(ls_prob, stream_buf, maskvalues = 1)
```
