---
title: "inundation"
format: html
---

## Modeling runout width

Inundation can be calculated from cross-sectional area. There are a few ways to estimate cross-sectional area:

1.  estimate a flow volume and velocity for each node/pixel
2.  estimating volume and using empirical scaling relationships

It seems reasonable to me to take advantage of slope and valley geometry to estimate velocity. In any case, we don't have an easy way to model volume or velocity so estimates will be pretty crude. In the absence of volume estimates, we may just end up using a constant volume for every pixel.

```{r}
library(dplyr)
library(tidyr)
library(ggplot2)
library(terra)

covariates <- c("UpDist_m", "UpArea_m2", "gradient","log_grad",
                "tangential_curv","log_width",
                "Prof", "logaccum", "dem", "drop","start_grad","grad_roll", "tan_roll")
nodes_df <- nodes_df[nodes_df$Width_m>0,]
nodes_df$log_width <- log10(nodes_df$Width_m)



long_df <- nodes_df %>%
  select(all_of(covariates)) %>%
  pivot_longer(-log_width, names_to = "variable", values_to = "value")

cor_labels <- nodes_df %>%
  select(all_of(covariates)) %>%
  pivot_longer(-log_width, names_to = "variable", values_to = "value") %>%
  group_by(variable) %>%
  summarise(r = cor(value, log_width, use = "complete.obs"),
            label = paste0("r = ", round(r, 2)))

# Compute binned averages (20 bins per covariate)
binned_df <- long_df %>%
  group_by(variable) %>%
  mutate(bin = cut(value, breaks = 20)) %>%
  group_by(variable, bin) %>%
  summarise(value = mean(value, na.rm = TRUE),
            log_width = mean(log_width, na.rm = TRUE),
            .groups = "drop")

ggplot(long_df, aes(x = value, y = log_width)) +
  geom_point(alpha = 0.2, size = 0.8) +
  geom_point(data = binned_df, aes(x = value, y = log_width),
             color = "red", size = 2) +
  geom_line(data = binned_df, aes(x = value, y = log_width),
            color = "red") +
  geom_text(data = cor_labels, aes(label = label),
            x = -Inf, y = Inf, hjust = -0.1, vjust = 1.5,
            color = "blue", size = 3) +
  facet_wrap(~variable, scales = "free_x") +
  theme_bw()
```

```{r}
width_model <- lm(log_width ~ logaccum + tangential_curv + gradient+dem, data = nodes_df)
summary(width_model)

nodes_df$predicted <- 10^predict(width_model)

ggplot(nodes_df, aes(x = predicted, y = Width_m)) +
  geom_point(alpha = 0.1) +
  geom_abline(slope = 1, intercept = 0, color = "red", linetype = "dashed") +
  labs(x = "Predicted Width_m", y = "Measured Width_m",
       title = "Measured vs. Predicted") +
  theme_bw()
```

```{r}
width_map <- 10^predict(mitkof$test_stack,width_model)

se_pred <- sigma(width_model)  # residual standard error

# Predicted log width raster
log_width_map <- predict(mitkof$test_stack, width_model)

plot(width_map)
```

```{r}
hist(values(log_width_map))
```

## Modeling Inundation

To model inundation, we must expand runout probabilities across predicted widths. We can achieve this by expanding runout probabilities across a circular kernel defined by the predicted local width. We can incorporate uncertainty in the width by

```{r}
dist_m <- seq(0.1, 60, length.out = 500)
width  <- 30
se     <- 0.36

p_reach <- 1 - pnorm((log10(2 * dist_m) - log10(width)) / se)

plot(dist_m, p_reach, type = "l", lwd = 2,
     xlab = "Distance from center (m)", ylab = "P(reach)",
     main = "Log-normal width decay (width=30m, se=0.36)")
abline(v = width/2, lty = 2, col = "red")
legend("topright", legend = "width/2 = 15m", lty = 2, col = "red", bty = "n")
```

This is a computationally heavy task that R performs very slowly. We've retained the R version of the inundation function below in case useful, but ultimately we translated this function to C++ using Rcpp (the following code block).

```{r}
# inundate <- function(runout,width,se){
#   inundation <- rast(runout)
#   
#   out_vals  <- values(runout)
#   width_vals <- values(width)
#   
#   nrows <- nrow(runout)
#   ncols <- ncol(runout)
#   res_m <- res(runout)[1]
#   rm(runout,width)
#   result    <- out_vals  # start with original values
#   
#   radius <- ceiling(10^(log10(width_vals) + 1 * se) / 2 / res_m )
#   
#   for (i in seq_len(length(out_vals))) {
#     if (is.na(out_vals[i]) || out_vals[i] == 0) next
#     r <- radius[i]
#     
#     row <- ((i - 1) %/% ncols) + 1
#     col <- ((i - 1) %%  ncols) + 1
#     
#     row_min <- max(1, row - r)
#     row_max <- min(nrows, row + r)
#     col_min <- max(1, col - r)
#     col_max <- min(ncols, col + r)
#     
#     neighbor_rows <- rep(row_min:row_max, each = (col_max - col_min + 1))
#     neighbor_cols <- rep(col_min:col_max, times = (row_max - row_min + 1))
#     neighbors     <- (neighbor_rows - 1) * ncols + neighbor_cols
#     
#     # Euclidean distance in meters (circular kernel)
#     dist_m <- sqrt(((neighbor_rows - row)^2 + (neighbor_cols - col)^2)) * res_m
#     
#     # Probability that flow reaches each neighbor given log-normal width uncertainty
#     p_reach <- 1 - pnorm((log10(2 * dist_m) - log10(width_vals[i])) / se)
#     
#     # Decayed probability
#     decayed_prob <- out_vals[i] * p_reach
#     
#     # Only update if higher than existing value
#     result[neighbors] <- pmax(result[neighbors], decayed_prob, na.rm = TRUE)
#   }
#   
#   # Write back to raster
#   values(inundation) <- result
#   return(inundation)
# }
```

```{r}
library(Rcpp)

cppFunction('
NumericVector inundate_cpp(NumericVector out_vals, 
                            IntegerVector width_vals,
                            IntegerVector radius,
                            int nrows, int ncols, 
                            double res_m, double se) {
  
  NumericVector result = clone(out_vals);
  int n = out_vals.size();
  
  for (int i = 0; i < n; i++) {
    if (NumericVector::is_na(out_vals[i]) || out_vals[i] == 0) continue;
    
    int r = radius[i];
    int row = i / ncols;
    int col = i % ncols;
    
    int row_min = std::max(0, row - r);
    int row_max = std::min(nrows - 1, row + r);
    int col_min = std::max(0, col - r);
    int col_max = std::min(ncols - 1, col + r);
    
    for (int nr = row_min; nr <= row_max; nr++) {
      for (int nc = col_min; nc <= col_max; nc++) {
        int j = nr * ncols + nc;
        
        double dist_m = sqrt(pow((double)(nr - row), 2) + 
                             pow((double)(nc - col), 2)) * res_m;
        
        if (dist_m == 0) dist_m = res_m * 0.5;
        
        double p_reach = 1 - R::pnorm((log10(2 * dist_m) - 
                                       log10(width_vals[i])) / se, 
                                       0, 1, 1, 0);
        double decayed = out_vals[i] * p_reach;
        
        if (decayed > result[j]) result[j] = decayed;
      }
    }
  }
  return result;
}
')

# Then your inundate function becomes:
inundate <- function(runout, width, se) {
  inundation <- rast(runout)
  out_vals   <- values(runout)[,1]
  width_vals <- as.integer(values(width)[,1])
  nrows      <- nrow(runout)
  ncols      <- ncol(runout)
  res_m      <- res(runout)[1]
  radius     <- round(10^(log10(width_vals) + 2 * se) / 2 / res_m)
  radius[is.na(radius)] <- 0L

  result <- inundate_cpp(out_vals, width_vals, as.integer(radius), 
                         nrows, ncols, res_m, se)
  values(inundation) <- result
  return(inundation)
}
```

```{r}
inundation_cox <- inundate(cox_runout,width_map,se_pred)
mitkof$ls_poly <- vect("/Users/Jeff/Dropbox/Documents/TealWaters/Data/Tongass/mitkof_shapefiles/LS_poly.shp")
mitkof$test_stack$ls_poly <- rasterize(mitkof$ls_poly,mitkof$test_stack$dem)
inun_prop_cox <- suscept_prop(mitkof$test_stack$ls_poly,inundation_cox)
plot(mitkof$test_stack$hs, col = gray.colors(256), legend = FALSE)
plot(inun_prop_cox, col = rev(hcl.colors(256, "RdYlGn")), add = TRUE, axes = FALSE, alpha = 0.35)
plot(mitkof$ls_poly, add = TRUE,legend = FALSE)
```

```{r}
inundation_test <- inundate(test_rast,width_map,se_pred)
inun_prop_test <- suscept_prop(mitkof$test_stack$ls_poly,inundation_test)
plot(mitkof$test_stack$hs, col = gray.colors(256), legend = FALSE)
plot(inun_prop_test, col = rev(hcl.colors(256, "RdYlGn")), add = TRUE, axes = FALSE, alpha = 0.35)
plot(mitkof$ls_poly, add = TRUE,legend = FALSE)
```

```{r}
inundation_cox_gam <- inundate(cox_gam_runout,width_map,se_pred)
inun_prop_cox_gam <- suscept_prop(mitkof$test_stack$ls_poly,inundation_cox_gam)
plot(mitkof$test_stack$hs, col = gray.colors(256), legend = FALSE)
plot(inun_prop_cox_gam, col = rev(hcl.colors(256, "RdYlGn")), add = TRUE, axes = FALSE, alpha = 0.35)
plot(mitkof$ls_poly, add = TRUE,legend = FALSE)
```

```{r}
inundation_logit <- inundate(logit_runout,width_map,se_pred)
inun_prop_logit <- suscept_prop(mitkof$test_stack$ls_poly,inundation_logit)
plot(mitkof$test_stack$hs, col = gray.colors(256), legend = FALSE)
plot(inun_prop_logit, col = rev(hcl.colors(256, "RdYlGn")), add = TRUE, axes = FALSE, alpha = 0.35)
plot(mitkof$ls_poly, add = TRUE,legend = FALSE)
```

# Applying to full datasets

```{r}
# runout <- rast("")

# width_map_full <- as.integer(10^predict(mitkof$rstack,width_model))
# width_vals <- as.integer(values(width_map_full))
# width_vals[!is.na(width_vals) & width_vals > 100] <- 100
# values(width_map_full) <- width_vals
# writeRaster(width_map_full,"/Users/Jeff/Dropbox/Documents/TealWaters/modeling_results/prediction_maps/mitkof_width.tif",datatype = "INT4S")
# gcrm(width_vals,width_map_full);gc()
width_map_full <- rast("/Users/Jeff/Dropbox/Documents/TealWaters/modeling_results/prediction_maps/mitkof_width.tif")
w <- 2*round((mean(nodes_df$Width_m,na.rm=TRUE)-1)/(2*res(width_map_full)[1]))+1
prob <- 1-1/w
width_map_smooth <- focal(
  width_map_full,
  w      = w,   # square window (scalar → w x w matrix of 1s)
  fun    = "max",
  na.policy = "omit"       # only compute where center cell is valid
)
writeRaster(width_map_smooth,"/Users/Jeff/Dropbox/Documents/TealWaters/modeling_results/prediction_maps/mitkof_width_smooth_max.tif")

inundation_full <- inundate(out_rast,width_map_smooth,se_pred)
```

```{r}
# mitkof$ls_poly <- vect("/Users/Jeff/Dropbox/Documents/TealWaters/Data/Tongass/mitkof_shapefiles/LS_poly.shp")
# mitkof$rstack$ls_poly <- rasterize(mitkof$ls_poly,mitkof$rstack$hs)
inun_prop_full <- suscept_prop(mitkof$rstack$ls_poly,inundation_smoothest)
writeRaster(inun_prop_full,"/Users/Jeff/Dropbox/Documents/TealWaters/modeling_results/prediction_maps/mitkof_inun_prop_smoothest_banks_masked.tif",overwrite=TRUE)
full_auc(mitkof$rstack$ls_poly,inun_prop_full)
```

```{r}
smooth_width <- function(width_map, runout, w, threshold = 0.3) {
  
  # 1. Sparse width map: only cells where runout exceeds threshold
  width_runout <- ifel(runout > threshold, width_map, NA)
  
  # 2. Smooth the sparse raster — focal ignores NA cells (non-runout),
  #    so only high-runout neighbors contribute to each smoothed value
  width_smoothed <- focal(
    width_runout,
    w         = w,
    fun       = "mean",
    na.policy = "omit",   # NA cells excluded from kernel
    na.rm     = TRUE
  )
  
  # 3. Overwrite full width map only where runout exceeds threshold
  width_out <- ifel(runout > threshold, width_smoothed, width_map)
  
  return(width_out)
}


smooth_runout <- function(runout, w, threshold = 0.3) {
  
  # 1. Sparse width map: only cells where runout exceeds threshold
  runout_high <- ifel(runout > threshold, runout, NA)
  
  # 2. Smooth the sparse raster — focal ignores NA cells (non-runout),
  #    so only high-runout neighbors contribute to each smoothed value
  runout_smooth <- focal(
    runout_high,
    w         = w,
    fun       = "mean",
    na.policy = "omit",   # NA cells excluded from kernel
    na.rm     = TRUE
  )
  
  # 3. Overwrite full width map only where runout exceeds threshold
  runout <- ifel(runout > threshold, runout_smooth, runout)
  
  return(runout)
}
```

```{r}
# cppFunction('
# NumericVector inundate_elev_cpp(NumericVector out_vals, 
#                                 NumericVector width_vals,
#                                 NumericVector elev_vals,
#                                 IntegerVector radius,
#                                 int nrows, int ncols, 
#                                 double res_m, double se,
#                                 double elev_thresh) {
#   
#   NumericVector result = clone(out_vals);
#   int n = out_vals.size();
#   
#   for (int i = 0; i < n; i++) {
#     if (NumericVector::is_na(out_vals[i]) || out_vals[i] == 0) continue;
#     if (NumericVector::is_na(elev_vals[i])) continue;
#     
#     int r = radius[i];
#     int row = i / ncols;
#     int col = i % ncols;
#     
#     int row_min = std::max(0, row - r);
#     int row_max = std::min(nrows - 1, row + r);
#     int col_min = std::max(0, col - r);
#     int col_max = std::min(ncols - 1, col + r);
#     
#     for (int nr = row_min; nr <= row_max; nr++) {
#       for (int nc = col_min; nc <= col_max; nc++) {
#         int j = nr * ncols + nc;
#         
#         if (NumericVector::is_na(elev_vals[j])) continue;
#         if (std::abs(elev_vals[j] - elev_vals[i]) > elev_thresh) continue;
#         
#         double dist_m = sqrt(pow((double)(nr - row), 2) + 
#                              pow((double)(nc - col), 2)) * res_m;
#         
#         if (dist_m == 0) dist_m = res_m * 0.5;
#         
#         double p_reach = 1 - R::pnorm((log10(2 * dist_m) - 
#                                        log10(width_vals[i])) / se, 
#                                        0, 1, 1, 0);
#         double decayed = out_vals[i] * p_reach;
#         
#         if (decayed > result[j]) result[j] = decayed;
#       }
#     }
#   }
#   return result;
# }
# ')
# 
# inundate <- function(runout, width, se, elev, elev_thresh = 2.0) {
#   inundation <- rast(runout)
#   out_vals   <- values(runout)[,1]
#   width_vals <- values(width)[,1]
#   elev_vals  <- values(elev)[,1]
#   nrows      <- nrow(runout)
#   ncols      <- ncol(runout)
#   res_m      <- res(runout)[1]
#   radius     <- ceiling(10^(log10(width_vals) + 1 * se) / 2 / res_m)
#   radius[is.na(radius)] <- 0L
# 
#   result <- inundate_elev_cpp(out_vals, width_vals, elev_vals,
#                               as.integer(radius), 
#                               nrows, ncols, res_m, se, elev_thresh)
#   values(inundation) <- result
#   return(inundation)
# }

```

```{r}
cppFunction('
NumericVector smooth_flowpath_cpp(NumericVector vals,
                                   IntegerVector fdir_vals,
                                   IntegerVector sorted_idx,
                                   int nrows, int ncols,
                                   int half_window,
                                   double threshold) {
  int n = vals.size();
  
  // Build downstream lookup
  std::vector<int> downstream(n, -1);
  for (int i = 0; i < n; i++) {
    if (IntegerVector::is_na(fdir_vals[i])) continue;
    int fdir = fdir_vals[i];
    int row = i / ncols;
    int col = i % ncols;
    int j = -1;
    if      (fdir == 1)   { if (col < ncols-1)                  j = i + 1; }
    else if (fdir == 2)   { if (row < nrows-1 && col < ncols-1) j = i + ncols + 1; }
    else if (fdir == 4)   { if (row < nrows-1)                  j = i + ncols; }
    else if (fdir == 8)   { if (row < nrows-1 && col > 0)       j = i + ncols - 1; }
    else if (fdir == 16)  { if (col > 0)                        j = i - 1; }
    else if (fdir == 32)  { if (row > 0 && col > 0)             j = i - ncols - 1; }
    else if (fdir == 64)  { if (row > 0)                        j = i - ncols; }
    else if (fdir == 128) { if (row > 0 && col < ncols-1)       j = i - ncols + 1; }
    if (j >= 0 && j < n) downstream[i] = j;
  }
  
  // For each cell, walk half_window steps downstream and half_window steps
  // upstream (via sorted_idx parent lookup) and average
  // Simple approximation: just walk downstream from each cell
  NumericVector result = clone(vals);
  
  for (int k = 0; k < sorted_idx.size(); k++) {
    int i = sorted_idx[k] - 1;
    if (NumericVector::is_na(vals[i]) || vals[i] < threshold) continue;
    
    // Collect window of values walking downstream
    double sum = 0.0;
    int count = 0;
    int cur = i;
    for (int w = 0; w <= half_window; w++) {
      if (cur < 0 || cur >= n) break;
      if (!NumericVector::is_na(vals[cur])) {
        sum += vals[cur];
        count++;
      }
      cur = downstream[cur];
      if (cur < 0) break;
    }
    if (count > 0) result[i] = sum / count;
  }
  return result;
}
')

cppFunction('
NumericVector smooth_flowpath_centered_cpp(NumericVector vals,
                                            IntegerVector fdir_vals,
                                            IntegerVector sorted_idx,
                                            int nrows, int ncols,
                                            int half_window,
                                            double threshold) {
  int n = vals.size();
  
  // Build downstream lookup (same as before)
  std::vector<int> downstream(n, -1);
  for (int i = 0; i < n; i++) {
    if (IntegerVector::is_na(fdir_vals[i])) continue;
    int fdir = fdir_vals[i];
    int row = i / ncols;
    int col = i % ncols;
    int j = -1;
    if      (fdir == 1)   { if (col < ncols-1)                  j = i + 1; }
    else if (fdir == 2)   { if (row < nrows-1 && col < ncols-1) j = i + ncols + 1; }
    else if (fdir == 4)   { if (row < nrows-1)                  j = i + ncols; }
    else if (fdir == 8)   { if (row < nrows-1 && col > 0)       j = i + ncols - 1; }
    else if (fdir == 16)  { if (col > 0)                        j = i - 1; }
    else if (fdir == 32)  { if (row > 0 && col > 0)             j = i - ncols - 1; }
    else if (fdir == 64)  { if (row > 0)                        j = i - ncols; }
    else if (fdir == 128) { if (row > 0 && col < ncols-1)       j = i - ncols + 1; }
    if (j >= 0 && j < n) downstream[i] = j;
  }
  
  // Build upstream lookup: for each cell, find neighbor that drains into it
  // with highest runout probability
  // A neighbor drains into cell i if downstream[neighbor] == i
  // fdir opposites: 1<->16, 2<->32, 4<->64, 8<->128
  std::vector<int> upstream(n, -1);
  for (int i = 0; i < n; i++) {
    int j = downstream[i];
    if (j < 0 || j >= n) continue;
    // i drains into j - check if i is the best upstream donor for j
    if (upstream[j] == -1) {
      upstream[j] = i;
    } else {
      // keep whichever donor has higher runout probability
      int cur_best = upstream[j];
      bool i_better = !NumericVector::is_na(vals[i]) && 
                      (NumericVector::is_na(vals[cur_best]) || vals[i] > vals[cur_best]);
      if (i_better) upstream[j] = i;
    }
  }
  
  NumericVector result = clone(vals);
  
  for (int k = 0; k < sorted_idx.size(); k++) {
    int i = sorted_idx[k] - 1;
    if (NumericVector::is_na(vals[i]) || vals[i] < threshold) continue;
    
    double sum = 0.0;
    int count = 0;
    
    // Walk upstream half_window steps
    int cur = i;
    for (int w = 0; w <= half_window; w++) {
      if (cur < 0 || cur >= n) break;
      if (!NumericVector::is_na(vals[cur])) {
        sum += vals[cur];
        count++;
      }
      cur = upstream[cur];
      if (cur < 0) break;
    }
    
    // Walk downstream half_window steps (skip i, already counted)
    cur = downstream[i];
    for (int w = 0; w < half_window/2; w++) {
      if (cur < 0 || cur >= n) break;
      if (!NumericVector::is_na(vals[cur])) {
        sum += vals[cur];
        count++;
      }
      cur = downstream[cur];
    }
    
    if (count > 0) result[i] = sum / count;
  }
  return result;
}
')

smooth_flowpath <- function(runout, dem, flowdir, half_window = 10, threshold = 0.3) {
  vals       <- values(runout)[,1]
  fdir_vals  <- as.integer(values(flowdir))
  sorted_idx <- as.integer(order(values(dem), decreasing = TRUE, na.last = NA))
  
  smoothed_vals <- smooth_flowpath_centered_cpp(vals, fdir_vals, sorted_idx,
                                        nrow(dem), ncol(dem),
                                        half_window, threshold)
  smoothed <- rast(runout)
  values(smoothed) <- smoothed_vals
  return(smoothed)
}

# Then:
runout_smooth_test <- smooth_flowpath(runout_test, dem_test, flowdir_test, half_window = 17)
```

```{r}
ls_poly <- vect("/Users/Jeff/Dropbox/Documents/TealWaters/Data/Tongass/mitkof_shapefiles/LS_poly.shp")
ls_poly_rast <- rasterize(ls_poly,dem)
runout_smooth <- rast("/Users/Jeff/Dropbox/Documents/TealWaters/modeling_results/prediction_maps/runout_smooth_temp.tif")
width_smooth <- rast("/Users/Jeff/Dropbox/Documents/TealWaters/modeling_results/prediction_maps/width_smooth_temp.tif")
inundate_full_smoothrunout <- inundate(runout_smooth,width_smooth,se = 0.35)
inun_prop_full <- suscept_prop(ls_poly_rast,inundate_full_smoothrunout)
writeRaster(inun_prop_full,"/Users/Jeff/Dropbox/Documents/TealWaters/modeling_results/prediction_maps/mitkof_inun_prop_smoothestrunout.tif")
full_auc(ls_poly_rast,inun_prop_full)
```

```{r}
cppFunction('
List smooth_flowpath_centered_cpp(NumericVector vals,
                                   IntegerVector width_vals,
                                   IntegerVector fdir_vals,
                                   IntegerVector sorted_idx,
                                   int nrows, int ncols,
                                   int half_window,
                                   double threshold) {
  int n = vals.size();
  
  // Build downstream lookup
  std::vector<int> downstream(n, -1);
  for (int i = 0; i < n; i++) {
    if (IntegerVector::is_na(fdir_vals[i])) continue;
    int fdir = fdir_vals[i];
    int row = i / ncols;
    int col = i % ncols;
    int j = -1;
    if      (fdir == 1)   { if (col < ncols-1)                  j = i + 1; }
    else if (fdir == 2)   { if (row < nrows-1 && col < ncols-1) j = i + ncols + 1; }
    else if (fdir == 4)   { if (row < nrows-1)                  j = i + ncols; }
    else if (fdir == 8)   { if (row < nrows-1 && col > 0)       j = i + ncols - 1; }
    else if (fdir == 16)  { if (col > 0)                        j = i - 1; }
    else if (fdir == 32)  { if (row > 0 && col > 0)             j = i - ncols - 1; }
    else if (fdir == 64)  { if (row > 0)                        j = i - ncols; }
    else if (fdir == 128) { if (row > 0 && col < ncols-1)       j = i - ncols + 1; }
    if (j >= 0 && j < n) downstream[i] = j;
  }
  
  // Build upstream lookup: highest runout probability donor
  std::vector<int> upstream(n, -1);
  for (int i = 0; i < n; i++) {
    int j = downstream[i];
    if (j < 0 || j >= n) continue;
    if (upstream[j] == -1) {
      upstream[j] = i;
    } else {
      int cur_best = upstream[j];
      bool i_better = !NumericVector::is_na(vals[i]) && 
                      (NumericVector::is_na(vals[cur_best]) || vals[i] > vals[cur_best]);
      if (i_better) upstream[j] = i;
    }
  }
  
  NumericVector result_vals  = clone(vals);
  IntegerVector result_width = clone(width_vals);
  
  for (int k = 0; k < sorted_idx.size(); k++) {
    int i = sorted_idx[k] - 1;
    if (NumericVector::is_na(vals[i]) || vals[i] < threshold) continue;
    
    double sum_v = 0.0, sum_w = 0.0;
    int count = 0;
    
    // Walk upstream
    int cur = i;
    for (int w = 0; w <= half_window; w++) {
      if (cur < 0 || cur >= n) break;
      if (!NumericVector::is_na(vals[cur])) {
        sum_v += vals[cur];
        sum_w += NumericVector::is_na(width_vals[cur]) ? 0.0 : width_vals[cur];
        count++;
      }
      cur = upstream[cur];
      if (cur < 0) break;
    }
    
    // Walk downstream (skip i)
    cur = downstream[i];
    for (int w = 0; w < half_window/2; w++) {
      if (cur < 0 || cur >= n) break;
      if (!NumericVector::is_na(vals[cur])) {
        sum_v += vals[cur];
        sum_w += NumericVector::is_na(width_vals[cur]) ? 0.0 : width_vals[cur];
        count++;
      }
      cur = downstream[cur];
    }
    
    if (count > 0) {
      result_vals[i]  = sum_v / count;
      result_width[i] = sum_w / count;
    }
  }
  
  return List::create(Named("runout") = result_vals,
                      Named("width")  = result_width);
}
')

smooth_flowpath <- function(runout, width, dem, flowdir, half_window = 10, threshold = 0.3) {
  rows <- nrow(dem)
  cols <- ncol(dem)
  sorted_idx <- as.integer(order(values(dem), decreasing = TRUE, na.last = NA))
  rm(dem);gc()
  vals       <- values(runout)[,1]
  width_vals <- as.integer(values(width)[,1])
  fdir_vals  <- as.integer(values(flowdir))
  
  out <- smooth_flowpath_centered_cpp(vals, width_vals, fdir_vals, sorted_idx,
                                      rows, cols,
                                      half_window, threshold)
  rm(vals, width_vals, fdir_vals, sorted_idx); gc()
  
  runout <- rast(runout)
  values(runout) <- out$runout
  writeRaster(runout, "/Users/Jeff/Dropbox/Documents/TealWaters/modeling_results/prediction_maps/runout_smooth_temp.tif", overwrite = TRUE)
  runout <- rast("/Users/Jeff/Dropbox/Documents/TealWaters/modeling_results/prediction_maps/runout_smooth_temp.tif")
  
  width <- rast(width)
  values(width) <- out$width
  writeRaster(width,  "/Users/Jeff/Dropbox/Documents/TealWaters/modeling_results/prediction_maps/width_smooth_temp.tif",  overwrite = TRUE)
  width  <- rast("/Users/Jeff/Dropbox/Documents/TealWaters/modeling_results/prediction_maps/width_smooth_temp.tif")
  rm(out); gc()
  
  return(list(runout = runout, width = width))
}
```

```{r}
library(terra)
library(Rcpp)
flowdir <- rast("/Users/Jeff/Dropbox/Documents/TealWaters/Data/Tongass/mitkof_rasters/flowdir.tif")
dem <- rast("/Users/Jeff/Dropbox/Documents/TealWaters/Data/Tongass/mitkof_rasters/dem_filled.tif")
width_map_full <- rast("/Users/Jeff/Dropbox/Documents/TealWaters/modeling_results/prediction_maps/mitkof_width.tif")
width_smooth <- rast("/Users/Jeff/Dropbox/Documents/TealWaters/modeling_results/prediction_maps/width_smooth_temp.tif")
runout_smooth <- rast("/Users/Jeff/Dropbox/Documents/TealWaters/modeling_results/prediction_maps/runout_smooth_temp.tif")
runout <- rast("/Users/Jeff/Dropbox/Documents/TealWaters/modeling_results/prediction_maps/mitkof_runout_cox.tif")
logaccum <- rast("/Users/Jeff/Dropbox/Documents/TealWaters/Data/Tongass/mitkof_rasters/logaccum.flt")
ls_prob <- rast("/Users/Jeff/Dropbox/Documents/TealWaters/modeling_results/prediction_maps/mitkof_gam_final3.tif")
# inundate_full <- inundate(runout,width_map_full,se = 0.35)
```

```{r}
# gc()
# inundate(runout, width_map_full, se = 0.35)
# gc()
smoothed <- smooth_flowpath(out_rast, width_map_full, dem, flowdir,
                             half_window = 17, threshold = 0.3)
gc()
inundation_smoothest <- inundate(smoothed$runout, smoothed$width, se = 0.35)
```

```{r}
stream_pts <- as.points(ifel(logaccum > 6, 1, NA), 
                         na.rm = TRUE)

# 2. Buffer the points by 50 m and dissolve into one polygon
stream_buf_poly <- aggregate(buffer(stream_pts, width = 50))

# 3. Rasterize the buffer polygon (1 inside, NA outside)
stream_buf_r <- rasterize(stream_buf_poly, logaccum, 
                           field = 1, background = NA)

# 4. Mask ls_prob: NA within the buffered stream zone
ls_prob_masked <- mask(ls_prob, stream_buf_r, maskvalues = 1)
```

```{r}
stream_r <- ifel(logaccum > 6, 1, NA)

# Expand by 50 m (the raster resolution is stored in res()); a 50 m buffer
# is achieved with a focal window of radius ceil(50 / res) pixels
pixel_res  <- res(logaccum)[1]          # e.g. 5 m
focal_r    <- ceiling(50 / pixel_res)                      # radius in pixels
win_size   <- 2 * focal_r + 1                              # odd window side length

# Any pixel within 50 m of a stream gets value 1; everything else stays NA
stream_buf <- focal(stream_r,
                    w         = win_size,
                    fun       = "max",
                    na.policy = "omit",
                    na.rm     = TRUE)

# Mask ls_prob: set to NA wherever the stream buffer is 1
ls_prob_masked <- mask(ls_prob, stream_buf, maskvalues = 1)
```

```{r}
library(Rcpp)

cppFunction('
NumericVector tpi_cpp(NumericVector elev_vals,
                      int nrows, int ncols,
                      double res, double D,
                      int n_pts = 8) {

  const double R   = D / 2.0;
  const double PI2 = 2.0 * M_PI;
  const int    n   = nrows * ncols;

  NumericVector out(n, NA_REAL);
  
  std::vector<int> drow(n_pts), dcol(n_pts);
  for (int k = 0; k < n_pts; k++) {
    double angle = PI2 * k / n_pts;
    drow[k] = (int)std::round(-(R * std::cos(angle)) / res);
    dcol[k] = (int)std::round( (R * std::sin(angle)) / res);
  }

  for (int i = 0; i < n; i++) {

    if (NumericVector::is_na(elev_vals[i])) continue;

    int row = i / ncols;
    int col = i % ncols;

    double sum_n = 0.0;
    int    cnt   = 0;

    for (int k = 0; k < n_pts; k++) {
      int nr = row + drow[k];
      int nc = col + dcol[k];
      if (nr < 0 || nr >= nrows || nc < 0 || nc >= ncols) continue;
      double val = elev_vals[nr * ncols + nc];
      if (NumericVector::is_na(val)) continue;
      sum_n += val;
      cnt++;
    }

    if (cnt > 0)
      out[i] = elev_vals[i] - (sum_n / cnt);
  }

  return out;
}
')

tpi_terra <- function(dem, D, n_pts = 8) {
  tpi_vals <- tpi_cpp(
    elev_vals = terra::values(dem)[, 1],
    nrows     = terra::nrow(dem),
    ncols     = terra::ncol(dem),
    res       = terra::res(dem)[1],
    D         = D,
    n_pts     = n_pts
  )
  out <- terra::rast(dem)
  terra::values(out) <- tpi_vals
  names(out) <- paste0("tpi_D", D)
  return(out)
}
```

```{r}
tpi <- tpi_terra(dem,D = 2500)
```
