---
title: "inundation"
format: html
---

## Modeling Inundation

The runout modeling described on the previous page provides a raster prediction of the probability that each pixel will be along the center axis of a landslide runout pathway. However, the hazard of landslide inundation extends across the landslide width, not just the centerline pixel. We therefore need some way of estimating the breadth of the landslide as it flows downslope. Most existing landslide indunation models require some estimate of landslide volume and then use empirical scaling relationships combined with cross-sectional topography to model inundation. This approach can work well when landslide volumes are known or can be reliably predicted. However, we do not have estimates of landslide volume.

Instead of relying on volume estimates, we have adopted a more direct statistical approach to modeling landslide width. We measured landslide width for every node along the centerlines of the mapped landslides. We can compare those width measurements against topographic and positional metrics at the same locations and build a statistical model based on those relationships.

```{r}
library(dplyr)
library(tidyr)
library(ggplot2)
library(terra)
source("./functions.R")

mitkof <- readRDS(
  "/Users/Jeff/Dropbox/Documents/TealWaters/Data/Tongass/mitkof_rasters/mitkof_state.rds")
wrangell <- readRDS(
  "/Users/Jeff/Dropbox/Documents/TealWaters/Data/Tongass/wrangell_rasters/wrangell_state.rds")

mitkof   <- load_rstack(mitkof)
wrangell <- load_rstack(wrangell)


covariates <- c("UpDist_m", "UpArea_m2", "gradient","log_grad",
                "tangential_curv","log_width",
                "Prof", "logaccum", "dem", "drop","start_grad","grad_roll", "tan_roll")
width_prep <- function(island){
  df <- island$nodes_df
  df <- df[df$Width_m>0,]
  df$log_width <- log10(df$Width_m)
  return(df)
}

mitkof$nodes_df <- width_prep(mitkof)
wrangell$nodes_df <- width_prep(wrangell)
```

```{r}
width_scatter <- function(island,form){
  nodes_df <- island$nodes_df
  long_df <- nodes_df %>%
    select(all_of(covariates)) %>%
    pivot_longer(-log_width, names_to = "variable", values_to = "value")
  
  cor_labels <- nodes_df %>%
    select(all_of(covariates)) %>%
    pivot_longer(-log_width, names_to = "variable", values_to = "value") %>%
    group_by(variable) %>%
    summarise(r = cor(value, log_width, use = "complete.obs"),
              label = paste0("r = ", round(r, 2)))
  
  # Compute binned averages (20 bins per covariate)
  binned_df <- long_df %>%
    group_by(variable) %>%
    mutate(bin = cut(value, breaks = 20)) %>%
    group_by(variable, bin) %>%
    summarise(value = mean(value, na.rm = TRUE),
              log_width = mean(log_width, na.rm = TRUE),
              .groups = "drop")
  
  p1 <- ggplot(long_df, aes(x = value, y = log_width)) +
    geom_point(alpha = 0.2, size = 0.8) +
    geom_point(data = binned_df, aes(x = value, y = log_width),
               color = "red", size = 2) +
    geom_line(data = binned_df, aes(x = value, y = log_width),
              color = "red") +
    geom_text(data = cor_labels, aes(label = label),
              x = -Inf, y = Inf, hjust = -0.1, vjust = 1.5,
              color = "blue", size = 3) +
    facet_wrap(~variable, scales = "free_x") +
    theme_bw()
  print(p1)
  
  width_model <- lm(form, data = nodes_df)
  print(summary(width_model))
  
  nodes_df$predicted <- 10^predict(width_model)
  
  p2 <- ggplot(nodes_df, aes(x = predicted, y = Width_m)) +
    geom_point(alpha = 0.1) +
    geom_abline(slope = 1, intercept = 0, color = "red", linetype = "dashed") +
    labs(x = "Predicted Width_m", y = "Measured Width_m",
         title = "Measured vs. Predicted") +
    theme_bw()
  print(p2)
  return(width_model)
}

mitkof$width_fit <- width_scatter(mitkof,log_width ~ logaccum + tangential_curv + log_grad+dem)
mitkof$width_se <- sigma(mitkof$width_fit)

wrangell$width_fit <- width_scatter(wrangell,log_width ~ logaccum + tangential_curv + log_grad + dem)
wrangell$width_se <- sigma(wrangell$width_fit)
```

```{r}
width_map <- 10^predict(mitkof$test_stack,mitkof$width_fit)

plot(width_map)
```

```{r}
hist(values(log_width_map))
```

## Modeling Inundation

To model inundation, we must expand runout probabilities across predicted widths. We can achieve this by expanding runout probabilities across a circular kernel defined by the predicted local width. We can incorporate uncertainty in the width by

```{r}
dist_m <- seq(0.1, 60, length.out = 500)
width  <- 30
se     <- 0.36

p_reach <- 1 - pnorm((log10(2 * dist_m) - log10(width)) / se)

plot(dist_m, p_reach, type = "l", lwd = 2,
     xlab = "Distance from center (m)", ylab = "P(reach)",
     main = "Log-normal width decay (width=30m, se=0.36)")
abline(v = width/2, lty = 2, col = "red")
legend("topright", legend = "width/2 = 15m", lty = 2, col = "red", bty = "n")
```

This is a computationally heavy task that R performs very slowly. We've retained the R version of the inundation function below in case useful, but ultimately we translated this function to C++ using Rcpp (the following code block).

```{r}
inundation_cox <- inundate(cox_runout,width_map,se_pred)
mitkof$ls_poly <- vect("/Users/Jeff/Dropbox/Documents/TealWaters/Data/Tongass/mitkof_shapefiles/LS_poly.shp")
mitkof$test_stack$ls_poly <- rasterize(mitkof$ls_poly,mitkof$test_stack$dem)
inun_prop_cox <- suscept_prop(mitkof$test_stack$ls_poly,inundation_cox)
plot(mitkof$test_stack$hs, col = gray.colors(256), legend = FALSE)
plot(inun_prop_cox, col = rev(hcl.colors(256, "RdYlGn")), add = TRUE, axes = FALSE, alpha = 0.35)
plot(mitkof$ls_poly, add = TRUE,legend = FALSE)
```

```{r}
inundation_test <- inundate(test_rast,width_map,se_pred)
inun_prop_test <- suscept_prop(mitkof$test_stack$ls_poly,inundation_test)
plot(mitkof$test_stack$hs, col = gray.colors(256), legend = FALSE)
plot(inun_prop_test, col = rev(hcl.colors(256, "RdYlGn")), add = TRUE, axes = FALSE, alpha = 0.35)
plot(mitkof$ls_poly, add = TRUE,legend = FALSE)
```

```{r}
inundation_cox_gam <- inundate(cox_gam_runout,width_map,se_pred)
inun_prop_cox_gam <- suscept_prop(mitkof$test_stack$ls_poly,inundation_cox_gam)
plot(mitkof$test_stack$hs, col = gray.colors(256), legend = FALSE)
plot(inun_prop_cox_gam, col = rev(hcl.colors(256, "RdYlGn")), add = TRUE, axes = FALSE, alpha = 0.35)
plot(mitkof$ls_poly, add = TRUE,legend = FALSE)
```

```{r}
inundation_logit <- inundate(logit_runout,width_map,se_pred)
inun_prop_logit <- suscept_prop(mitkof$test_stack$ls_poly,inundation_logit)
plot(mitkof$test_stack$hs, col = gray.colors(256), legend = FALSE)
plot(inun_prop_logit, col = rev(hcl.colors(256, "RdYlGn")), add = TRUE, axes = FALSE, alpha = 0.35)
plot(mitkof$ls_poly, add = TRUE,legend = FALSE)
```
