---
title: "Model Testing"
author: "Jeff Prancevic"
---

# Model exploration

Now that we have training datasets, we can dive into advanced modeling. We're considering several types of models:

-   **Logistic regression model (logit)** - A logit model fits a logistic curve to binary classification data according to independent predictor variables. A logit model describes the probability that a set of predictor values corresponds to a landslide `label = 1` or background `label = 0` value (see example below). It relies on monotonic relationships between predictors and outcome (i.e., that the probability of finding a landslide cell monotonically increases with increasing slope or with decreasing FoS). A logistic regression model will struggle to fit a non-monotonic relationship, for example, one where landslide peaks at some intermediate predictor value. For this reason, aspect may be a poor predictor in a logit model, while `northness` and `eastness` may be more useful. The only parameters to consider for a logit model is the selection of predictors.

    ```{r}
    glm_fit <- function(formula,train_df){
      model <- glm(formula, train_df, family = binomial())
      return(model)
    }

    predi <- function(model,df){  
      mclass <- class(model)[1]
      
      if(mclass=="ranger"){
        pred <- predict(model, data = df,
                          num.threads = parallel::detectCores() - 1)$predictions[, 2]
      }else{
        pred <- predict(model, newdata = df, type = "response")
      }
      return(pred)
    }
    ```

    ```{r}
    # load(file = "/Users/Jeff/Dropbox/Documents/TealWaters/Data/Tongass/landslide_modeling_data.RData")

    df <- wrangell$df[sample(nrow(wrangell$df), 100), ]
    df$label <- as.numeric(as.character(df$label))
    df <- df[order(df$gradient), ]

    prob   <- predi(glm_fit(label ~ gradient,df),df)

    plot(df$gradient, df$label, pch = 19, col = "gray40",
         xlab = "Topographic gradient", ylab = "P(Landslide)")
    lines(df$gradient, prob, col = "steelblue", lwd = 2)
    rm(prob,df)
    ```

-   **Generalized Additive Model (GAM)** - A GAM model is similar to a logit model, in that it relies on linear regression between predictor variables and a response (though not limited to a binary outcome). However, a GAM can fit any smooth relationship between predictor and response using splines. The user can specify the number of basis functions `k` to build the spline. Unlike a logit model, a GAM model can be tuned by adjusting `k` for each predictor variable.

    ```{r}
    gam_fit <- function(formula,train_df){
      model <- mgcv::gam(
        formula,
        data   = train_df,
        family = binomial,
        method = "REML"
      )
      return(model)
    }
    ```

-   **Random Forest modeling (RF)** - An RF model is a machine learning algorithm for classification. It allows for complex relationships between predictors and the response variables by building decision trees, which, in turn, discern classification by discretizing sets of predictor values. RFs have numerous hyperparameters that can be tuned. Here, we focus on `num.trees` the number of decision trees to build and `min.node.size` the minimum number of nodes (decision points) in each tree, as well as the combination of predictors.

    ```{r}
    library(ranger)
    rf_fit <- function(formula,train_df,ntrees=500,min_node=10){
      fit <- tryCatch(
        ranger(formula, data = train_df, num.trees = ntrees,
                       min.node.size = min_node, mtry = 1,
                       importance = "impurity",
                       num.threads = parallel::detectCores() - 1,
                       probability = TRUE),  # needed for prob predictions
        error = function(e) NULL
      )
      return(fit)
    }
    ```

# Evaluating Models

In this section we'll identify the best combination of predictors and the best model hyperparameters for assessing landslide susceptibility on both islands. With 16 covariates, there are 65,535 possible permutations of predictors. Combining that with 3 different model types and a quasi-infinite range of tunable hyperparameters, evaluating all possible models is not practical. Instead, we'll pare down the number of predictors and evaluate a range of reasonable hyperparameters. We will assess model performance in two ways:

1.  Quantitatively using average AUC values from our spatial 4-fold cross-validation analysis.
2.  Qualitatively by visually assessing prediction maps for our 1 km test slice. We're looking for predictions that are not noisy and do not contain artifacts, in order to provide clear guidance for forest management.

```{r}
## This function converts any susceptibility map with probability (or really any increasing quantitative measure) into a map of landslide proportions. In other words, the values of the resulting raster will display the percent of landslides cells with susceptibility values smaller than this cell. E.g., a value of 100 indicates that 100% of landslide cells have lower landslide susceptibility and this should be interpreted as an unstable pixel.
suscept_prop <- function(init,susceptibility){
  init_vals <- values(init)
  susc_vals <- values(susceptibility)
  
  ls_susc <- susc_vals[!is.na(init_vals)]
  ls_susc <- ls_susc[!is.na(ls_susc)]  # drop any NA in susc at those locations
  
  ecdf_fn <- ecdf(ls_susc)
  
  susc_remapped <- app(susceptibility, function(x) {
    # proportion of ls_susc <= x  →  invert to get proportion > x  →  scale to 0–100
    ecdf_fn(x) * 100
  })
  return(susc_remapped)
}

## Create a function to measure the AUC for full paired rasters of proportions and initiation zones
full_auc <- function(init,prop){
  init_mask <- !is.na(values(init))
  susc_vals <- values(prop)
  
  
  pos_s <- sort(susc_vals[init_mask & !is.na(susc_vals)])
  neg_s <- sort(susc_vals[!init_mask & !is.na(susc_vals)])
  
  thresholds <- seq(0, 100, length.out = 500)
  tpr <- 1 - findInterval(thresholds, pos_s) / length(pos_s)
  fpr <- 1 - findInterval(thresholds, neg_s) / length(neg_s)
  tpr[1] <- 1
  fpr[1] <- 1

  auc <- abs(sum(diff(fpr) * (head(tpr, -1) + tail(tpr, -1)) / 2))
  return(auc)
}

## Create a function to map the test prediction for any model
map_test_pred <- function(model,island,inset=TRUE){
  df <- island$test_df
  tstack <- island$test_stack
  # e <- ext(tstack)
  mclass <- class(model)[1]
  # df_train <- island$df[island$df$x >= e[1] & island$df$x <= e[2] &
  #                      island$df$y >= e[3] & island$df$y <= e[4], ]
  
  pred <- predi(model,df)
  if(mclass=="ranger"){
    preds <- model$forest$independent.variable.names
    mclass <- 'RF'
  }else{
    preds <- all.vars(formula(model))[-1]
  }
  
  suscept <- rast(tstack$mask)
  values(suscept) <- NA
  suscept[df$ind] <- as.vector(pred)
  
  prop <- suscept_prop(tstack$init_zone,suscept)
  auc <- full_auc(tstack$init_zone,prop)
  mars <- c(1, 1, 4, 1)
  plot(tstack$hs, col = gray.colors(256), legend = FALSE,
       main = paste(island$name,toupper(mclass),"Prediction Map\n",
                    "Preds:", paste(preds, collapse = ", "),"\n AUC:",round(auc, 3)),
       cex.main = 1, mar = mars)
  
  plot(prop, col = rev(hcl.colors(256, "RdYlGn")), add = TRUE, axes = FALSE, alpha = 0.35, mar = mars)
  
  init_polys <- as.polygons(tstack$init_zone, dissolve = TRUE)
  plot(init_polys,add = TRUE, axes = FALSE,legend = FALSE, mar = mars)
  
  if (isTRUE(inset)) {
    rect(island$inset$xlim[1], island$inset$ylim[1],
         island$inset$xlim[2], island$inset$ylim[2],
         border = "black", lwd = 2, lty = 2)
  
    inset_ext <- ext(island$inset$xlim[1], island$inset$xlim[2],
                     island$inset$ylim[1], island$inset$ylim[2])
    hs_crop   <- crop(tstack$hs,        inset_ext)
    prop_crop <- crop(prop,             inset_ext)
    zone_crop <- crop(tstack$init_zone, inset_ext)
  
    tmp <- tempfile(fileext = ".png")
    png(tmp, width = 300, height = 300)
    par(mar = c(0, 0, 0, 0))
    image(hs_crop,   col = gray.colors(256), axes = FALSE, xlab = "", ylab = "")
    image(prop_crop, col = adjustcolor(rev(hcl.colors(256, "RdYlGn")), alpha.f = 0.5), add = TRUE)
    # plot(as.polygons(zone_crop, dissolve = TRUE),lwd = 2, add = TRUE, axes = FALSE, legend = FALSE)
    plot(as.polygons(zone_crop, dissolve = TRUE), add = TRUE, 
     axes = FALSE, legend = FALSE, col = NA, border = "black", lwd = 3)
    dev.off()
  
    img <- png::readPNG(tmp)
  
    usr     <- par("usr")
    x_range <- usr[2] - usr[1]
    y_range <- usr[4] - usr[3]
    pad_x   <- x_range * 0.1
    pad_y   <- y_range * 0.05
    w       <- x_range * island$inset$size[1]
    h       <- y_range * island$inset$size[2]
    
  
    inset_pos <- switch(island$inset$loc,
      bottomright = c(usr[2] - w - pad_x, usr[3] + pad_y,usr[2] - pad_x, usr[3] + h + pad_y),
      bottomleft  = c(usr[1] + pad_x,     usr[3] + pad_y,     usr[1] + w + pad_x, usr[3] + h + pad_y),
      topright    = c(usr[2] - w - pad_x, usr[4] - h - pad_y, usr[2] - pad_x, usr[4] - pad_y),
      topleft     = c(usr[1] + pad_x,     usr[4] - h - pad_y, usr[1] + w + pad_x, usr[4] - pad_y)
    )
  
    rasterImage(img, inset_pos[1], inset_pos[2], inset_pos[3], inset_pos[4])
    rect(inset_pos[1], inset_pos[2], inset_pos[3], inset_pos[4], border = "black", lwd = 2)
  }

  invisible(NULL)
}
```

## Should we include categorical predictors?

Before we get into choosing model parameters, we wanted to assess whether to include polygon-based categorical parameters (`soils`, `lith`, and `forest`). Categorical shapefiles provide a coarse discretization of the landscape. We've experienced this in other projects and also in preliminary susceptibility maps in the Tongass. However, they have the potential to provide meaningful distinction in soil and vegetation properties that can improve susceptibility maps. Below, we evaluate model performance with and without `lithology` and `forest` polygons on both islands. We have elected to omit `soils` polygons, as the only distinction between soil classes is explicitly tied to topographic gradient, and thus already included in the model.

```{r fig.width=12, fig.height=11}
formula_shp <- label ~ fos_72h+ northness+ norm_curv+logaccum +lith+forest
formula <- label ~ fos_72h+ northness+ norm_curv+logaccum 

par(mfrow = c(2, 2))
island <- wrangell
map_test_pred(glm_fit(formula,island$df), island)
map_test_pred(glm_fit(formula_shp,island$df), island)
island <- mitkof
map_test_pred(glm_fit(formula,island$df), island)
map_test_pred(glm_fit(formula_shp,island$df), island)

```

The sample prediction maps above show the variable impact that these categorical variables can have on predictions. In Wrangell, lith and forest polygons improve the model performance for the sample slice, both quantitatively and visually, without creating dramatic polygon artifacts. The opposite is true for Mitkof: the polygons imprint sharp contrasts in landslide susceptibility that look unnatural, and the quantitative performance (as measured by AUC) decreases substantially when categorical predictors are included.

For the remaining evaluation, we will include `lith` and `forest` predictors for Wrangell, but we will exclude all categorical predictors for Mitkof.

## Checking covariates

All 13 quantitative covariates are derived from topographic measurements that we've classified into 4 groups: slope, curvature, area, and aspect. Including all of these measures is likely redundant. Intuitively, it likely makes sense to use only one covariate from each group in the final model. In this section, we will quickly assess the covariability and stability of covariates.

```{r}

library(car)
library(tidyverse)
library(broom)
library(ggcorrplot)

cov_check <- function(island){

  # ============================================================
# Group Coherence Diagnostics
# ============================================================
  vars <- c("gradient", "fos_72h", "fos_6h",
            "logaccum", "area_72h", "area_6h",
            "mean_curv", "tangential_curv", "norm_curv",
            "aspect", "northness", "eastness")
  
  groups <- list(
    Slope     = c("gradient", "fos_72h", "fos_6h"),
    Area      = c("logaccum", "area_72h", "area_6h"),
    Curvature = c("mean_curv", "tangential_curv", "norm_curv"),
    Aspect    = c("aspect", "northness", "eastness")
  )
  
  if(island$name=="Mitkof"){
    vars <- c(vars,"profile_curv")
    groups$Curvature<- c(groups$Curvature,"profile_curv")
  }
  
  train_df<-island$df
  
  group_lookup <- imap_dfr(groups, ~ tibble(term = .x, group = .y))
  
  # Scale predictors
  train_scaled <- train_df |>
    mutate(across(all_of(vars), ~ as.numeric(scale(.))))
  
  
  # ============================================================
  # PLOT 1: Correlation heatmap
  # ============================================================
  
  cor_matrix <- cor(train_scaled[, vars], use = "complete.obs")
  
  p1 <- ggcorrplot(
    cor_matrix,
    hc.order  = FALSE,
    type      = "lower",
    lab       = TRUE,
    lab_size  = 2.5,
    colors    = c("#4575b4", "white", "#d73027"),
    title     = "Feature Correlation"
  )
  
  # ============================================================
  # PLOT 3: Absolute marginal estimate bar chart
  # ============================================================
  
  marginal_coefs <- map_dfr(vars, function(v) {
    f <- as.formula(paste("label ~", v))
    m <- glm_fit(f, train_scaled)
    tidy(m) |>
      filter(term == v)
  })
  
  marginal_df <- marginal_coefs |>
    left_join(group_lookup, by = "term") |>
    mutate(term = fct_reorder(term, abs(estimate)))
  
  p2 <- ggplot(marginal_df, aes(x = term, y = abs(estimate), fill = group)) +
    geom_col() +
    coord_flip() +
    labs(title = "Scaled individual fit coefficient",
         x = NULL, y = "|Coefficient|", fill = "Group") +
    theme_minimal()
  
  
  # ============================================================
  # PLOT 2: VIF bar chart (x axis capped at 15)
  # ============================================================
  
  
  full_formula <- as.formula(paste("label ~", paste(vars, collapse = " + ")))
  joint_model  <- glm_fit(full_formula, train_scaled)
  
  vif_df <- tibble(term = names(vif(joint_model)), VIF = vif(joint_model)) |>
    left_join(group_lookup, by = "term") |>
    mutate(term = fct_reorder(term, desc(VIF)))

  p3 <- ggplot(vif_df, aes(x = term, y = VIF, fill = group)) +
    geom_col() +
    geom_hline(yintercept = 5,  linetype = "dashed", colour = "orange") +
    geom_hline(yintercept = 10, linetype = "dashed", colour = "red") +
    coord_flip(ylim = c(0, 15)) +
    labs(title = "Variance Inflation Factors",
         x = NULL, y = "VIF", fill = "Group") +
    theme_minimal()
  
  (p1 | (p2 + theme(legend.position = "none") | p3)) +
  plot_annotation(title = island$name)
}
```

### Mitkof

```{r fig.width=15, fig.height=6}
cov_check(mitkof)
```

In **Mitkof**, curvature and gradient covariates are all corelated with each other (left panel). This leads to high variance inflation factors for those variables (right panel), meaning that the relationships between those variables and landslide susceptibility become less clear when modeled alongside other covariates from their groups. For example, `gradient` can discern landslide prone areas on its own, but it struggles to further refine landslide susceptibility for areas that have already been modeled using another slope metric. Consequently, we will use one predictor each from these groups to ensure predictor stability. Based on the individual fit coefficients (middle panel), all predictors within these groups are similarly predictive of landslide susceptibility as other predictors within the groups. `fos_6h` and `tangential_curv` perform best but we will consider all.

We will exclude `area_6h` from the rest of the analysis. It is better correlated with slope metrics than area metrics, but it is not as strong of a predictor as other slope metrics. Among the two remaining area predictors (`logaccum` and `area_72h`), `area_72h` is a much stronger predictor on its own (center panel), and still has a relatively low VIF (right panel), making it a clear winner. `logaccum` will be excluded from the remaining analyses for Mitkof.

Within the aspect group, `northness` has the highest individual fit coefficient and the lowest VIF, giving it a clear advantage, particularly for the logit model. However, because `aspect` has a non-monotonic relationship with landslide susceptibility, it may perform better than expected in GAM and RF modeling. We will exclude `eastness` and include `aspect` and `northness`.

### Wrangell

```{r fig.width=15, fig.height=6}
cov_check(wrangell)
```

The analysis for Wrangell is very similar to Mitkof:

-   Curvature and slope covariates are all correlated and unstable when used together. It is not clear which predictor is best within these groups, so all will be considered.

-   `area_6h` is a second-rate measure of slope gradient and it will be excluded as a predictor.

For area, `area_72h` is a much better predictor than `logaccum` on its own but (unlike for Mitkof) it has a concerningly high VIF. We will evaluate the performance of both predictors.

For aspect predictors, raw `aspect` is a better predictor than for Mitkof. `eastness` is a better solo predictor than `northness` but it is also less stable (higher VIF). We will evaluate the performance of all three predictors.

## Grid Search

Now that we've identified which predictors are most reliable for both Wrangell and Mitkof, we will evaluate the performance of three types of models (logit, GAM, and RF) using all possible combinations of predictors, selecting one predictor from each group for each successive model run. For the logit and GAM models we will only vary the set of predictors. For the random forest model, we will also vary the number of trees we build and average over (300 and 500), as well as the minimum number of nodes within each tree (5 or 10).

For each model run, we will record the unique combination of predictors and hyperparameters and the associated performance statistics. We will use cross-validation to test 4 different fits of the model again 4 test datasets (based on our spatial blocking). We will calculate the average AUC value across the 4 test folds.

We take the best three predictor/parameter sets (as measured by AUC) and predict landslide susceptibility for our test slices. We will visually assess the top three model fits for each type of model and select the best to compare against the other models.

**Note for Dan and Jeff: the text below these maps may not match once the model is re-run with a different randomly sampled training dataset. I'll keep the text in, as it will be interesting to see if it matches future runs.**

```{r}
grid_search_cv <- function(island, predictor_list, model,
                           ntrees_list = c(300, 500), min_nodes_list = c(5, 10)) {
  
  df       <- island$df
  fold_ids <- island$fold_ids
  folds    <- unique(fold_ids)
  
  # --- Build parameter grid ---
  if (model == "rf") {
    param_grid <- expand.grid(
      pred_idx  = seq_along(predictor_list),
      ntrees    = ntrees_list,
      min_nodes = min_nodes_list
    )
  } else {
    param_grid <- data.frame(pred_idx = seq_along(predictor_list))
  }
  
  
  total   <- nrow(param_grid)
  counter <- 0L
  
  print(paste("Evaluating",model,"models for",island$name))

  # --- Evaluate each parameter set ---
  results <- pmap_dfr(param_grid, function(pred_idx, ntrees = NULL, min_nodes = NULL) {
    counter <<- counter + 1L
    cat(sprintf("\r[%d / %d] (%.1f%%)", counter, total, 100 * counter / total))
    flush.console()
    
    predictors <- predictor_list[[pred_idx]]
    formula    <- as.formula(paste("label ~", paste(predictors, collapse = " + ")))
    
    # k-fold cross-validation
    fold_aucs <- map_dbl(folds, function(k) {
      train <- df[fold_ids != k, ]
      test  <- df[fold_ids == k, ]
      
      model_fit <- if (model == "rf") {
        rf_fit(formula, train, ntrees = ntrees, min_node = min_nodes)
      } else if (model == "gam") {
        gam_terms <- ifelse(
          sapply(train[predictors], is.numeric),
          paste0("s(", predictors, ", k=4)"),
          predictors
        )
        formula <- as.formula(paste("label ~", paste(gam_terms, collapse = " + ")))
        gam_fit(formula, train)
      } else {
        glm_fit(formula, train)
      }
      
      probs  <- predi(model_fit, test)
      roc_obj <- pROC::roc(test$label, probs, quiet = TRUE)
      as.numeric(pROC::auc(roc_obj))
    })
    
    # Output row
    row <- tibble(
      island     = island$name,
      model      = model,
      predictors = paste(predictors, collapse = ", "),
      mean_auc   = mean(fold_aucs),
      sd_auc     = sd(fold_aucs)
    )
    if (model == "rf") {
      row$ntrees    <- ntrees
      row$min_nodes <- min_nodes
    }
    row
  })
  
  arrange(results, desc(mean_auc))
}
```

### Mitkof

```{r}
groups <- list(
    Slope     = c("gradient", "fos_72h", "fos_6h"),
    Area      = "area_72h",
    Curvature = c("mean_curv", "tangential_curv", "norm_curv","profile_curv"),
    Aspect    = c("aspect", "northness")
  )

pred_list <- expand.grid(groups,
  stringsAsFactors = FALSE)|>
  pmap(~ list(c(..1, ..2, ..3, ..4)))  |>
  list_flatten()
```

#### The best logit models

```{r}
mitkof$grid_glm <- grid_search_cv(mitkof,pred_list,"glm")
print(head(mitkof$grid_glm))
```

```{r fig.width=15, fig.height=6}
par(mfrow = c(1,3))
island <- mitkof

for (i in 1:3) {
  formula    <- as.formula(paste("label ~", paste(strsplit(island$grid_glm$predictors[i], ", ")[[1]], collapse = " + ")))
  map_test_pred(glm_fit(formula,island$df), island)
}
```

The top models all have essentially the same performance, as measured by AUC. The top three models vary only in the slope metric that they use. The two best models use the two factor of safety metrics (`fos_6h` and `fos_72h`) and are nearly indistinguishable. The model using gradient has a slightly different pattern. There is no obvious reason to not use the best model in this case, so we'll move `fos_6h` to the next round.

#### The best GAM models

```{r}
mitkof$grid_gam <- grid_search_cv(mitkof,pred_list,"gam")
print(head(mitkof$grid_gam))

```

```{r fig.width=15, fig.height=6}
par(mfrow = c(1,3))
island <- mitkof

for (i in 1:3) {
  predictors <- strsplit(island$grid_gam$predictors[i], ", ")[[1]]
  gam_terms <- ifelse(
    sapply(island$test_df[predictors], is.numeric),
    paste0("s(", predictors, ", k=4)"),
    predictors
  )
  formula <- as.formula(paste("label ~", paste(gam_terms, collapse = " + ")))
  map_test_pred(gam_fit(formula,island$df), island)
}
```

The GAM models performed slightly better than the logit models (larger mean AUC, smaller st. dev. of AUC). In this case the top 3 models all used `fos_72h` and `area_72h` but varied in which curvature and aspect measures they used. The model that uses `tangential_curv` and `aspect` (right panel, #3) produces predictions that are more smoothly varying. We will move model #3 forward.

#### The best RF models

```{r}
mitkof$grid_rf <- grid_search_cv(mitkof,pred_list,"rf")
print(head(mitkof$grid_rf))

```

```{r fig.width=15, fig.height=6}
par(mfrow = c(1,3))
island <- mitkof

for (i in 1:3) {
  formula    <- as.formula(paste("label ~", paste(strsplit(island$grid_rf$predictors[i], ", ")[[1]], collapse = " + ")))
  map_test_pred(rf_fit(formula,island$df,island$grid_rf$ntrees[i],island$grid_rf$min_nodes[i]), island)
}
```

The RF models performed worst of all three model types, with the lowest AUC values (across the entire training dataset) and the noisiest prediction maps. The AUC value for the test slide are actually quite good but the image is very pixelated, with high variation from pixel to pixel. All of the top three models used `min_nodes = 10`, and `area_72h`, `tangential_curv`, and `northness` as predictors. The model that used `gradient` (#1) is indistinguishable from those that used `fos_72h` (#2 and #3). We'll pass model #1 to the comparison but it's clearly far noiser than the others.

#### The best of the best for Mitkof

```{r fig.width=15, fig.height=6}
par(mfrow = c(1,3))
island <- mitkof

i<-1
formula    <- as.formula(paste("label ~", paste(strsplit(island$grid_glm$predictors[i], ", ")[[1]], collapse = " + ")))
map_test_pred(glm_fit(formula,island$df), island)

i<-3
formula    <- as.formula(paste("label ~", paste(paste0("s(", strsplit(island$grid_gam$predictors[i], ", ")[[1]], ", k=4)"), collapse = " + ")))
map_test_pred(gam_fit(formula,island$df), island)

i<-1
formula    <- as.formula(paste("label ~", paste(strsplit(island$grid_rf$predictors[i], ", ")[[1]], collapse = " + ")))
map_test_pred(rf_fit(formula,island$df,island$grid_rf$ntrees[i],island$grid_rf$min_nodes[i]), island)
```

Of the three models, the GAM model using `fos_72h`, `area_72h`, `tangential_curv`, and `aspect` is the smoothest. We will use that model for the final prediction map. We experimented with tuning the number of basis functions for each parameter below, but even though this increased the AUC slightly, the resulting map was noisier.

```{r}
mitkof$gam_form <- as.formula("label ~ s(fos_72h, k=4) + s(area_72h, k=4) + s(tangential_curv, k=4) + s(aspect, k=4)")

# island <- mitkof
# df       <- island$df
# fold_ids <- island$fold_ids
# folds    <- unique(fold_ids)
# fold_aucs <- map_dbl(folds, function(k) {
#   train <- df[fold_ids != k, ]
#   test  <- df[fold_ids == k, ]
#   
#   model_fit <- gam_fit(island$gam_form, train)
# 
#   probs  <- predi(model_fit, test)
#   roc_obj <- pROC::roc(test$label, probs, quiet = TRUE)
#   as.numeric(pROC::auc(roc_obj))
# })
# mean(fold_aucs)
# 
# par(mfrow = c(1,2))
# map_test_pred(gam_fit(island$gam_form,island$df), island)
# i<-3
# formula    <- as.formula(paste("label ~", paste(paste0("s(", strsplit(island$grid_gam$predictors[i], ", ")[[1]], ", k=4)"), collapse = " + ")))
# map_test_pred(gam_fit(formula,island$df), island)
```

### Wrangell

```{r}
groups <- list(
    Slope     = c("gradient", "fos_72h", "fos_6h"),
    Area      = c("logaccum", "area_72h"),
    Curvature = c("mean_curv", "tangential_curv", "norm_curv","profile_curv"),
    Aspect    = c("aspect", "northness", "eastness"),
    Lith = "lith",
    Forest = "forest"
  )

pred_list <- expand.grid(groups,
  stringsAsFactors = FALSE)|>
  pmap(~ list(c(..1, ..2, ..3, ..4, ..5, ..6)))  |>
  list_flatten()
```

#### Best logit models

```{r}
wrangell$grid_glm <- grid_search_cv(wrangell,pred_list,"glm")
print(head(wrangell$grid_glm))

```

```{r fig.width=15, fig.height=6}
par(mfrow = c(1,3))
island <- wrangell

for (i in 1:3) {
  formula    <- as.formula(paste("label ~", paste(strsplit(island$grid_glm$predictors[i], ", ")[[1]], collapse = " + ")))
  map_test_pred(glm_fit(formula,island$df), island)
}
```

The best logit maps all used `gradient` and `area_72h`, and varied only in curvature and aspect metrics. The results are visually indistinguishable, so we will pass the #1 map (using `tangential_curv` and `northness`) to the next step.

#### Best GAM models

```{r}
wrangell$grid_gam <- grid_search_cv(wrangell,pred_list,"gam")
print(head(wrangell$grid_gam))

```

```{r fig.width=15, fig.height=6}
par(mfrow = c(1,3))
island <- wrangell

for (i in 1:3) {
  predictors <- strsplit(island$grid_gam$predictors[i], ", ")[[1]]
  gam_terms <- ifelse(
    sapply(island$test_df[predictors], is.numeric),
    paste0("s(", predictors, ", k=4)"),
    predictors
  )
  formula <- as.formula(paste("label ~", paste(gam_terms, collapse = " + ")))
  map_test_pred(gam_fit(formula,island$df), island)
}
```

For the top GAM models, the best slope metric switched to `fos_72h` and all of the top models use `tangential_curv`. They vary in area and aspect metrics, but again, resulting maps have no meaningful visual distinction. We will pass the best model (#1), which again uses `area_72h` and `northness`, to the final evaluation.

#### Best RF models

```{r}
wrangell$grid_rf <- grid_search_cv(wrangell,pred_list,"rf")
print(head(wrangell$grid_rf))

```

```{r fig.width=15, fig.height=6}
par(mfrow = c(1,3))
island <- wrangell

for (i in 1:3) {
  formula    <- as.formula(paste("label ~", paste(strsplit(island$grid_rf$predictors[i], ", ")[[1]], collapse = " + ")))
  map_test_pred(rf_fit(formula,island$df,island$grid_rf$ntrees[i],island$grid_rf$min_nodes[i]), island)
}
```

The top three RF models all use the same predictors (`fos_72h`, `logaccum`, `tangential_curv`, and `northness`) but vary in their hyperparameters. The resulting models are nearly indistinguishable and we will pass the #1 model to the final evaluation. All models are noisy and pixelated, although not as bad as the RF models on Mitkof.

#### Best of the best

Based on the above results

```{r fig.width=15, fig.height=6}
par(mfrow = c(1,3))
island <- wrangell

i<-1
formula    <- as.formula(paste("label ~", paste(strsplit(island$grid_glm$predictors[i], ", ")[[1]], collapse = " + ")))
map_test_pred(glm_fit(formula,island$df), island)

i<-1
predictors <- strsplit(island$grid_gam$predictors[i], ", ")[[1]]
gam_terms <- ifelse(
  sapply(island$test_df[predictors], is.numeric),
  paste0("s(", predictors, ", k=4)"),
  predictors
)
formula <- as.formula(paste("label ~", paste(gam_terms, collapse = " + ")))
map_test_pred(gam_fit(formula,island$df), island)

i<-1
formula    <- as.formula(paste("label ~", paste(strsplit(island$grid_rf$predictors[i], ", ")[[1]], collapse = " + ")))
map_test_pred(rf_fit(formula,island$df,island$grid_rf$ntrees[i],island$grid_rf$min_nodes[i]), island)
```

Both the logit and GAM models are smoother than the RF model but it's difficult selecting between the two. The GAM model had better AUC metrics across the entire training dataset, but the logit model is attractive for its simplicity. We will use the logit model, which requires no additional hyperparameters.

```{r}
wrangell$glm_form <- as.formula("label ~ gradient + fos_72h + tangential_curv + northness + lith + forest")
```
