---
title: "prediction_maps"
format: html
---

# Making prediction maps

In order to make prediction maps with the structure modeling approaches we're using, we need to feed tabular data into the model. Rather than doing direct raster math, we will extract predictor values for each pixel, and then feed tables of predictor values into the model. Predictions are mapped back onto a raster with the same resolution as the raster inputs.

## Prediction metric: Proportions vs. Probabilities

As we're testing several types of models with various training points, we want to standardize model outputs according to landslide observations (the `init` raster layer). To do that, we recast predictions according to what portion of the observed landslide initiation pixels are captured by thresholding the landscape according to any output value. The advantage of this approach is that we can directly compare any arbitrary landslide susceptibility metric and recast it according to how it describes the observed landslide pattern.

In contrast, the outputs of the various models on the two different islands, typically mapped as "probabilities", are not necessarily directly comparable. Moreover, the term "probability" could be misleading in the context of landslide susceptibility mapping. A 60% probability prediction does not mean that there's a 60% chance of a landslide initiating at a pixel over some unspecified timescale. Rather, based on the topographic metrics, the model is estimating that there's a 60% chance that the pixel should be classified as one of the observed landslide pixels. Recasting these probabilities as proportions makes them intuitively meaningful.

The code block below ingests maps of predictors (rasters and shapefiles), runs those through any model, converts the output to ***proportions***, and then saves it to a geotiff that can be loaded into GIS software.

```{r}

full_pred_map <- function(island,model,outfile,shape_dir=NULL){
  
  if(class(model)[1]=="ranger"){
    preds <- model$forest$independent.variable.names
  }else{
    preds <- all.vars(formula(model))[-1]
  }
  
  b_inds <- names(island$rstack) %in% preds
  
  cat('indexing mask pixels... \n')
  # Don't store all_coords upfront - compute per chunk
  all_cells  <- which(!is.na(values(island$rstack$mask)))
  
  cat('initializing susceptibility map... \n')
  # Write to disk incrementally using a temp file approach
  susceptibility        <- rast(island$rstack$mask)
  values(susceptibility) <- NA
  writeRaster(susceptibility, outfile, overwrite=FALSE)
  susceptibility <- rast(outfile)  # open as file-backed raster
  
  chunk_size <- 5e5
  n_chunks   <- ceiling(length(all_cells) / chunk_size)
  
  cat_vars = c("soils","lith","forest")
  if(any(cat_vars %in% preds)){
    cat("loading shapefiles... \n")
    soils <- vect(file.path(shape_dir, "soils.shp"))
    soil_vals <- unique(island$df$soils)
    soils$SMU[!soils$SMU %in% soil_vals] <- 'U'
    
    lith <- vect(file.path(shape_dir, "lith.shp"))
    lith_vals <- unique(island$df$lith)
    lith$STATE_UNIT[!lith$STATE_UNIT %in% lith_vals] <- 'U'
    
    forest <- vect(file.path(shape_dir, "cover.shp"))
    for_vals <- unique(island$df$forest)
    forest$ftype[!forest$ftype %in% for_vals] <- 'U'
  }
  
  cat('making susceptibility predictions... \n')
  for (i in 1:n_chunks) {
    cat(sprintf("\r[%d / %d] (%.1f%%)", i, n_chunks, 100 * i / n_chunks))
    flush.console()
  
    idx    <- ((i-1)*chunk_size + 1):min(i*chunk_size, length(all_cells))
    
    # Compute coordinates per chunk rather than storing all upfront
    coords <- xyFromCell(island$rstack$mask, all_cells[idx])
    coords <- vect(coords, crs = crs(island$rstack$mask))
    
    chunk_df <- as.data.frame(terra::extract(island$rstack[[b_inds]],coords,method="simple"))
    if(any(cat_vars %in% preds)){
      lith_chunk <- terra::extract(lith, coords)
      chunk_df$lith <- lith_chunk$STATE_UNIT
      soils_chunk <- terra::extract(soils, coords)
      chunk_df$soils <- soils_chunk$FOR_SMU
      forest_chunk <- terra::extract(forest, coords)
      forest_chunk <- forest_chunk[!duplicated(forest_chunk$id.y), ]
      chunk_df$forest <- forest_chunk$ftype
    }
    
    chunk_pred <- predi(model,chunk_df)
    
    # Write directly to file-backed raster
    susceptibility[all_cells[idx]] <- as.vector(chunk_pred)
  
    rm(coords, chunk_df, chunk_pred); gc()
  }
  
  cat('converting probabilities to proportions... \n')
  susc_prop <- suscept_prop(island$rstack$init_zone,susceptibility)
  rm(susceptibility); gc()
  
  cat(paste('writing output file:',outfile,'\n'))
  writeRaster(susc_prop, outfile, overwrite=TRUE)
  
  cat('calculating AUC values for all pixels (background and landslide) \n')
  auc <- full_auc(island$rstack$init_zone,susc_prop)
  print(paste('AUC for full dataset =',auc))
  return(auc)
}
```

```{r}
mitkof$gam_fit <- gam_fit(mitkof$gam_form,mitkof$df)
full_pred_map(mitkof,mitkof$gam_fit,"/Users/Jeff/Dropbox/Documents/TealWaters/modeling_results/prediction_maps/mitkof_gam_final.tif")
```

```{r}
wrangell$glm_fit <- glm_fit(wrangell$glm_form,wrangell$df)
w_auc <- full_pred_map(wrangell,wrangell$glm_fit,"/Users/Jeff/Dropbox/Documents/TealWaters/modeling_results/prediction_maps/wrangell_glm_final.tif",shape_dir=w_shape_dir)
```
