---
title: "prediction_maps"
format: html
---

# Make prediction maps

## Fit desired model

```{r}
library(ranger)

# predictors <- c("fos_72h","gradient","mean_curv","tangential_curv")
predictors <- c("gradient","mean_curv","tangential_curv","soils","forest","lith")

formula <- as.formula(paste("label ~", paste(predictors, collapse = " + ")))

# # Fit final model on full training data
# fit_rf_final <- randomForest(formula, data = df_all, ntree = 500,
#                      nodesize = 10, mtry = 1,
#                      importance = TRUE)

# refit with ranger (same parameters)
fit_rf_final <- ranger(formula, data = mitkof$df, num.trees = 500,
                       min.node.size = 10, mtry = 1,
                       importance = "impurity",
                       num.threads = parallel::detectCores() - 1,
                       probability = TRUE)  # needed for prob predictions

```

## Load Mitkof data

```{r}

raster_dir <- "/Users/Jeff/Dropbox/Documents/TealWaters/Data/Tongass/mitkof_rasters/"
mitkof$mask <- rast(paste0(raster_dir,"mask.flt"))

```

```{r}

# Don't store all_coords upfront - compute per chunk
all_cells  <- which(!is.na(values(mitkof$mask)))

# Write to disk incrementally using a temp file approach
susceptibility        <- rast(mitkof$mask)
values(susceptibility) <- NA
outfile <- "/Users/Jeff/Dropbox/Documents/TealWaters/modeling_results/prediction_maps/RF_suscept_mitkof_selftrain_wshapes.tif"
writeRaster(susceptibility, outfile, overwrite=TRUE)
susceptibility <- rast(outfile)  # open as file-backed raster

chunk_size <- 5e5
n_chunks   <- ceiling(length(all_cells) / chunk_size)

for (i in 1:n_chunks) {
  cat("Processing chunk", i, "of", n_chunks, "\n")

  idx    <- ((i-1)*chunk_size + 1):min(i*chunk_size, length(all_cells))
  
  # Compute coordinates per chunk rather than storing all upfront
  coords <- xyFromCell(mitkof$mask, all_cells[idx])
  coords <- vect(coords, crs = crs(mitkof$mask))
  
  chunk_df <- as.data.frame(terra::extract(mitkof$rstack, coords,method="simple"))
  lith_chunk <- terra::extract(lith, coords)
  chunk_df$lith <- lith_chunk$STATE_UNIT
  soils_chunk <- terra::extract(soils, coords)
  chunk_df$soils <- soils_chunk$FOR_SMU
  forest_chunk <- terra::extract(forest, coords)
  chunk_df$forest <- forest_chunk$fprod
  
  chunk_pred <- predict(fit_rf_final, data = chunk_df,
                      num.threads = parallel::detectCores() - 1)$predictions[, 2]
  
  # Write directly to file-backed raster
  susceptibility[all_cells[idx]] <- chunk_pred

  rm(coords, chunk_df, chunk_pred); gc()
}

writeRaster(susceptibility, outfile, overwrite=TRUE)
```

```{r}
init <- rast(paste0(raster_dir,'init.flt'))

init_vals <- values(init)
susc_vals <- values(susceptibility)

ls_susc <- susc_vals[!is.na(init_vals)]
ls_susc <- ls_susc[!is.na(ls_susc)]  # drop any NA in susc at those locations

# Step 2: Build empirical CDF of ls_susc and remap the full susceptibility raster
# ecdf(x)(x) gives the proportion of ls_susc values <= x
# Inverting: 100 means all ls_susc values are LARGER than this pixel's value
#            0   means no  ls_susc values are larger (this pixel is at the top)
#            50  = median of ls_susc

ecdf_fn <- ecdf(ls_susc)

susc_remapped <- app(susceptibility, function(x) {
  # proportion of ls_susc <= x  →  invert to get proportion > x  →  scale to 0–100
  ecdf_fn(x) * 100
})

writeRaster(susc_remapped,"/Users/Jeff/Dropbox/Documents/TealWaters/modeling_results/prediction_maps/RF_suscept_mitkof_selftrain_wshapes_prop.tif")
```

```{r}
# Extract susceptibility values with init as mask
init_mask <- !is.na(values(init))
susc_vals <- values(susc_remapped)

pos <- susc_vals[init_mask & !is.na(susc_vals)]
neg <- susc_vals[!init_mask & !is.na(susc_vals)]

# ROC
thresholds <- seq(0, 100, length.out = 500)
tpr <- sapply(thresholds, function(t) mean(pos >= t))
fpr <- sapply(thresholds, function(t) mean(neg >= t))

auc <- abs(sum(diff(fpr) * (head(tpr, -1) + tail(tpr, -1)) / 2))

plot(fpr, tpr, type = "l", col = "steelblue", lwd = 2,
     xlab = "Fraction of negatives above threshold",
     ylab = "Fraction of positives above threshold",
     main = sprintf("ROC Curve (AUC = %.4f)", auc))
abline(0, 1, lty = 2, col = "grey60")
```

## Logistic full-scale modeling

### Grid search for best combo of predictors

```{r}
gam_grid_search_kfold <- function(df,
                                  fold_ids,
                                  predictor_list = NULL,
                                  min_ch = 1,
                                  max_ch = NULL) {
  all_predictors <- setdiff(names(df), c("label", "x", "y","lith","soils","forest","dataset","ID"))
  if (is.null(max_ch)) max_ch <- length(all_predictors)
  if (is.null(predictor_list)) {
    predictor_list <- map(
      seq(min_ch, max_ch),
      ~ combn(all_predictors, .x, simplify = FALSE)
    ) |> list_flatten()
  } else {
    predictor_list <- keep(predictor_list, ~ length(.x) >= min_ch & length(.x) <= max_ch)
  }

  grid <- expand.grid(
    predictor_idx = seq_along(predictor_list)
  )

  total   <- nrow(grid)
  counter <- 0L
  k       <- length(unique(fold_ids))

  results <- pmap(grid, function(predictor_idx) {
    counter <<- counter + 1L
    cat(sprintf("\r[%d / %d] (%.1f%%)", counter, total, 100 * counter / total))
    flush.console()

    preds   <- predictor_list[[predictor_idx]]
    p       <- length(preds)
    formula <- as.formula(paste("label ~", paste(preds, collapse = " + ")))

    # --- inner CV loop across folds ---
    fold_aucs <- map_dbl(seq_len(k), function(fold_i) {
      train_df <- df[fold_ids != fold_i, ]
      test_df  <- df[fold_ids == fold_i, ]

      fit <- tryCatch(
        # glm(formula, data = train_df, family = binomial()),
        mgcv::gam(
          formula,
          data   = train_df,
          family = binomial,
          method = "REML"
        ),
        error   = function(e) NULL,
        warning = function(w) NULL  # catches convergence warnings
      )
      if (is.null(fit)) return(NA_real_)

      pred   <- predict(fit, newdata = test_df, type = "response")
      roc_glm <- roc(test_df$label, pred, quiet = TRUE)
      as.numeric(auc(roc_glm))
    })

    tibble(
      predictors   = paste(preds, collapse = " + "),
      n_predictors = p,
      mean_auc     = mean(fold_aucs, na.rm = TRUE),
      sd_auc       = sd(fold_aucs,   na.rm = TRUE)
    )
  }) |> list_rbind()

  results |> arrange(desc(mean_auc))
}
```

```{r}
group1 <- c("fos_72h", "fos_6h", "gradient")
group2 <- c("logaccum", "area_72h", "area_6h")
group3 <- c("aspect", "northness", "eastness")
group4 <- c("mean_curv", "norm_curv", "profile_curv", "tangential_curv")

predictor_list <- expand.grid(
  g1 = group1,
  g2 = group2,
  g3 = group3,
  g4 = group4,
  stringsAsFactors = FALSE
) |>
  pmap(~ list(c(..1, ..2, ..3, ..4)))  |>
  list_flatten()

wrangell_grid_gam_2 <- gam_grid_search_kfold(wrangell$df,sb_w$folds_ids,predictor_list = predictor_list)

```

```{r}
form_mitkof_gam <- label ~ s(fos_72h, k=5) + s(northness, k=3) + s(norm_curv, k=3)+ s(area_72h, k=10)
form_wrangell_gam <- label ~ s(fos_72h, k=10) + s(eastness, k=5) + s(tangential_curv, k=10)+ s(logaccum, k=5)

fold_aucs <- map_dbl(seq_len(4), function(fold_i) {
      train_df <- wrangell$df[sb$folds_ids != fold_i, ]
      test_df  <- wrangell$df[sb$folds_ids == fold_i, ]

      fit <- tryCatch(
        mgcv::gam(
          form_wrangell_gam,
          data   = train_df,
          family = binomial,
          method = "REML"
        ),
        error   = function(e) NULL,
        warning = function(w) NULL  # catches convergence warnings
      )
      if (is.null(fit)) return(NA_real_)

      pred   <- predict(fit, newdata = test_df, type = "response")
      roc_glm <- roc(test_df$label, pred, quiet = TRUE)
      as.numeric(auc(roc_glm))
    })
mean(fold_aucs)
```

```{r}

```

### Full-scale modeling

```{r}
formula <- as.formula(paste("label ~", mitkof_grid$predictors[1]))

fit_glm_final <- glm(formula, data = mitkof$df, family = binomial())

# Don't store all_coords upfront - compute per chunk
all_cells  <- which(!is.na(values(mask_r)))

# Write to disk incrementally using a temp file approach
susceptibility        <- rast(mask_r)
values(susceptibility) <- NA
outfile <- "/Users/Jeff/Dropbox/Documents/TealWaters/modeling_results/prediction_maps/logit_suscept_mitkof_selftrain.tif"
writeRaster(susceptibility, outfile, overwrite=TRUE)
susceptibility <- rast(outfile)  # open as file-backed raster

chunk_size <- 5e5
n_chunks   <- ceiling(length(all_cells) / chunk_size)

for (i in 1:n_chunks) {
  cat("Processing chunk", i, "of", n_chunks, "\n")

  idx    <- ((i-1)*chunk_size + 1):min(i*chunk_size, length(all_cells))
  
  # Compute coordinates per chunk rather than storing all upfront
  coords <- xyFromCell(mask_r, all_cells[idx])
  coords <- vect(coords, crs = crs(mask_r))
  
  chunk_df <- as.data.frame(terra::extract(mitkof$rstack, coords,method="simple"))
  
  chunk_pred <- predict(fit_glm_final, newdata = chunk_df,type = "response")
  
  # Write directly to file-backed raster
  susceptibility[all_cells[idx]] <- chunk_pred

  rm(coords, chunk_df, chunk_pred); gc()
}

writeRaster(susceptibility, outfile, overwrite=TRUE)
```

```{r}
init <- rast(paste0(raster_dir,'init.flt'))

init_vals <- values(init)
susc_vals <- values(susceptibility)

ls_susc <- susc_vals[!is.na(init_vals)]
ls_susc <- ls_susc[!is.na(ls_susc)]  # drop any NA in susc at those locations

# Step 2: Build empirical CDF of ls_susc and remap the full susceptibility raster
# ecdf(x)(x) gives the proportion of ls_susc values <= x
# Inverting: 100 means all ls_susc values are LARGER than this pixel's value
#            0   means no  ls_susc values are larger (this pixel is at the top)
#            50  = median of ls_susc

ecdf_fn <- ecdf(ls_susc)

susc_remapped <- app(susceptibility, function(x) {
  # proportion of ls_susc <= x  →  invert to get proportion > x  →  scale to 0–100
  ecdf_fn(x) * 100
})

writeRaster(susc_remapped,"/Users/Jeff/Dropbox/Documents/TealWaters/modeling_results/prediction_maps/GLM_suscept_mitkof_selftrain_prop.tif")
```

```{r}
# Extract susceptibility values with init as mask
init_mask <- !is.na(values(init))
susc_vals <- values(susc_remapped)

pos <- susc_vals[init_mask & !is.na(susc_vals)]
neg <- susc_vals[!init_mask & !is.na(susc_vals)]

# ROC
thresholds <- seq(0, 100, length.out = 500)
tpr <- sapply(thresholds, function(t) mean(pos >= t))
fpr <- sapply(thresholds, function(t) mean(neg >= t))

auc <- abs(sum(diff(fpr) * (head(tpr, -1) + tail(tpr, -1)) / 2))

plot(fpr, tpr, type = "l", col = "steelblue", lwd = 2,
     xlab = "Fraction of negatives above threshold",
     ylab = "Fraction of positives above threshold",
     main = sprintf("ROC Curve (AUC = %.4f)", auc))
abline(0, 1, lty = 2, col = "grey60")
```

```{r}
mitkof_scaled <- mitkof$df |> mutate(across(where(is.numeric) & !c(label, x, y), scale))
fit_glm <- glm(formula, data = mitkof_scaled, family = binomial())
summary(fit_glm)
```

```{r}

# formula <- as.formula(paste("label ~", mitkof_grid_gam$predictors[36]))
raster_dir <- "/Users/Jeff/Dropbox/Documents/TealWaters/Data/Tongass/wrangell_rasters/"
wrangell$mask <- rast(paste0(raster_dir,"mask.flt"))

fit_gam_final <- mgcv::gam(
          form_wrangell_gam,
          data   = wrangell$df,
          family = binomial,
          method = "REML"
)

preds <- all.vars(form_wrangell_gam)
preds <- preds[preds != "label"]  # remove response variable

# Don't store all_coords upfront - compute per chunk
all_cells  <- which(!is.na(values(wrangell$mask)))

# Write to disk incrementally using a temp file approach
susceptibility        <- rast(wrangell$mask)
values(susceptibility) <- NA
outfile <- "/Users/Jeff/Dropbox/Documents/TealWaters/modeling_results/prediction_maps/gam_suscept_wrangell_best.tif"
writeRaster(susceptibility, outfile, overwrite=TRUE)
susceptibility <- rast(outfile)  # open as file-backed raster

chunk_size <- 5e5
n_chunks   <- ceiling(length(all_cells) / chunk_size)

for (i in 1:n_chunks) {
  cat("Processing chunk", i, "of", n_chunks, "\n")

  idx    <- ((i-1)*chunk_size + 1):min(i*chunk_size, length(all_cells))
  
  # Compute coordinates per chunk rather than storing all upfront
  coords <- xyFromCell(wrangell$mask, all_cells[idx])
  coords <- vect(coords, crs = crs(wrangell$mask))
  
  chunk_df <- as.data.frame(terra::extract(wrangell$rstack[[preds]], coords,method="simple"))
  
  chunk_pred <- predict(fit_gam_final, newdata = chunk_df,type = "response")
  
  # Write directly to file-backed raster
  susceptibility[all_cells[idx]] <- matrix(chunk_pred, ncol = 1)

  rm(coords, chunk_df, chunk_pred); gc()
}

init <- rast(paste0(raster_dir,'init.flt'))

init_vals <- values(init)
susc_vals <- values(susceptibility)

ls_susc <- susc_vals[!is.na(init_vals)]
ls_susc <- ls_susc[!is.na(ls_susc)]  # drop any NA in susc at those locations


ecdf_fn <- ecdf(ls_susc)

susc_remapped <- app(susceptibility, function(x) {
  # proportion of ls_susc <= x  →  invert to get proportion > x  →  scale to 0–100
  ecdf_fn(x) * 100
})

writeRaster(susc_remapped,"/Users/Jeff/Dropbox/Documents/TealWaters/modeling_results/prediction_maps/gam_suscept_wrangell_best_prop.tif")

# Extract susceptibility values with init as mask
init_mask <- !is.na(values(init))
susc_vals <- values(susc_remapped)


pos_s <- sort(susc_vals[init_mask & !is.na(susc_vals)], decreasing = TRUE)
neg_s <- sort(susc_vals[!init_mask & !is.na(susc_vals)], decreasing = TRUE)

thresholds <- seq(0, 100, length.out = 500)
tpr <- 1 - findInterval(thresholds, rev(pos_s)) / length(pos_s)
fpr <- 1 - findInterval(thresholds, rev(neg_s)) / length(neg_s)


auc <- abs(sum(diff(fpr) * (head(tpr, -1) + tail(tpr, -1)) / 2))

plot(fpr, tpr, type = "l", col = "steelblue", lwd = 2,
     xlab = "Fraction of negatives above threshold",
     ylab = "Fraction of positives above threshold",
     main = sprintf("ROC Curve (AUC = %.4f)", auc))
abline(0, 1, lty = 2, col = "grey60")
```

## Explanation of model selection

We have worked through several modeling approaches to partition the landscape according to landslide susceptibility. We have concluded that both logistic regression and generalized additive modeling

```{r}

```
