---
title: "Landslide Runout Analysis"
format: html
---

# Evaluating runout pathways and building runout models

## Loading the node datasets

```{r}
library(terra)

nodes<-vect('/Users/Jeff/Dropbox/Documents/TealWaters/Data/Tongass/Mitkof_shapefiles/centerNodes.shp')
nodes$logaccum <- log10(nodes$UpArea_m2)
names(nodes)[names(nodes) == "Grad"] <- "gradient"
names(nodes)[names(nodes) == "Tan"] <- "tangential_curv"

nodes <- extract(mitkof$rstack$dem, nodes, bind = TRUE, method = "simple")
nodes <- nodes[!is.na(nodes$dem), ]

# Get elevation at the start point (min UpDist_m) for each polygon
start_elev <- tapply(nodes$dem[nodes$UpDist_m == 0], 
                     nodes$Polygon[nodes$UpDist_m == 0], 
                     mean)  # mean in case of ties

# Calculate drop with respect to start elevation
nodes$drop <- start_elev[as.character(nodes$Polygon)] - nodes$dem

# runout <- tapply(nodes$UpDist_m, nodes$Polygon, max, na.rm = TRUE)

```

## Plotting runout nodes

```{r fig.width=8, fig.height=8}
# Get gradient at start point (UpDist_m == 0) for each polygon
start_grad <- tapply(nodes$gradient[nodes$UpDist_m == 0],
                     nodes$Polygon[nodes$UpDist_m == 0],
                     mean)  # mean in case of ties

# Get the endpoint (max UpDist_m) for each polygon
endpoints <- do.call(rbind, lapply(split(as.data.frame(nodes), nodes$Polygon), function(df) {
  df[which.max(df$UpDist_m), ]
}))
# Map start gradient onto endpoints
endpoints$start_grad <- start_grad[as.character(endpoints$Polygon)]

# Color palette for gradient values
# Normalize start_grad to 0-1 range based on chosen limits
lo <- quantile(endpoints$start_grad, 0.05, na.rm = TRUE)
hi <- quantile(endpoints$start_grad, 0.95, na.rm = TRUE)

# Clamp values to [lo, hi] then normalize
clamped <- pmin(pmax(endpoints$start_grad, lo), hi)
normalized <- (clamped - lo) / (hi - lo)

# Map to colors
grad_colors <- colorRampPalette(c("blue", "white", "red"))(100)
endpoint_cols <- grad_colors[ceiling(normalized * 99) + 1]

# Plot
plot(nodes$UpDist_m, -nodes$drop,
     pch = 16, cex = 0.4, col = "black",
     xlab = "Runout Distance (m)", ylab = "Drop (m)")

points(endpoints$UpDist_m, -endpoints$drop,
       pch = 21, cex = 1.5, col = "black", bg = endpoint_cols)
legend("topright",
       legend = c(round(hi, 3), round((lo + hi) / 2, 3), round(lo, 3)),
       pt.bg  = grad_colors[c(100, 50, 1)],
       pch    = 21,
       pt.cex = 1.5,
       title  = "Gradient at initiation",
       bty    = "n")
```

```{r}

```

## Fitting runout models

```{r}
plot_survival_paths <- function(model, df) {
  
  # Compute cumulative survival
  if (inherits(model, "coxph")) {
    m_name <- 'Survival Analysis'
    df$hazard <- predict(model, type = "expected")
  } else if (inherits(model, "glm")) {
    m_name <- 'Logistic Regression'
    df$hazard <- predict(model, type = "response")
  }
  
  df <- do.call(rbind, lapply(split(df, df$Polygon), function(d) {
      d <- d[order(d$UpDist_m), ]
      d$surv_prob <- exp(-cumsum(d$hazard))
      d
    }))
  
  # Get endpoint survival probabilities
  endpoints <- do.call(rbind, lapply(split(df, df$Polygon), function(d) {
    d[nrow(d), ]
  }))
  print(paste("runout nodes with low probability (bad!):", sum(df$surv_prob <= 0.1, na.rm = TRUE)))
  print(paste("endpoints with low probability (good!):", sum(endpoints$surv_prob <= 0.1, na.rm = TRUE)))
  
  # Set up side-by-side layout
  layout(matrix(c(1, 2), nrow = 1), widths = c(3, 1))
  par(mar = c(5, 4, 4, 0))
  
  # Main survival plot
  plot(NULL,
       xlim = range(df$UpDist_m, na.rm = TRUE),
       ylim = c(0, 1),
       xlab = "Runout distance (m)", ylab = "P(flow passes node)",
       main = m_name)
  
  invisible(lapply(split(df, df$Polygon), function(d) {
    d <- d[order(d$UpDist_m), ]
    lines(d$UpDist_m, d$surv_prob, col = "black", lwd = 0.5)
  }))
  points(endpoints$UpDist_m, endpoints$surv_prob,
           pch = 21, cex = 1.2, col = "black", bg = "yellow")
  
  # Histogram panel
  h <- hist(endpoints$surv_prob, breaks = 20, plot = FALSE)
  par(mar = c(5, 0, 4, 2))
  barplot(h$counts,
          horiz      = TRUE,
          space      = 0,
          xlim       = c(0, 30),
          ylim       = c(0, length(h$breaks) - 1),
          col        = "yellow",
          border     = "black",
          xlab       = "Endpoint \n probabilities",
          main       = "",
          axes       = TRUE)
    
    layout(1)  # reset layout
  }
```

### "Local" survival model

I first want to test our ability to model hazard and local survival based only on local conditions. The model would be slightly easier to apply because we can make one hazard map for the entire study area and then quick track survival along runout paths. However, this will still require marching downslope from every potential initiation zone to quantify runout probabilities. \*\* could try to add 'prof' \*\*

```{r}
library(survival)

# Sort to ensure correct ordering
nodes_df <- as.data.frame(nodes)
nodes_df <- nodes_df[order(nodes_df$Polygon, nodes_df$UpDist_m), ]


# Create tstart/tstop and event indicator
nodes_df <- do.call(rbind, lapply(split(nodes_df, nodes_df$Polygon), function(df) {
  df$tstart <- df$UpDist_m
  df$tstop  <- c(df$UpDist_m[-1], df$UpDist_m[nrow(df)] + 1)  # shift by one row
  df$event  <- c(rep(0, nrow(df) - 1), 1)  # death at last node
  df
}))
nodes_df <- nodes_df[nodes_df$tstart<nodes_df$tstop,]

# Fit time-varying CoxPH
cox_model <- coxph(Surv(tstart, tstop, event) ~ gradient + tangential_curv + logaccum,
                   data = nodes_df,
                   na.action = na.omit)

lp_train <- predict(cox_model,type = "lp")
y <- cox_model$y
event_times <- sort(unique(y[, "stop"][y[, "status"] == 1]))
h0 <- sapply(event_times, function(t) {
  in_risk <- (y[, "start"] < t) & (y[, "stop"] >= t)
  sum(y[, "status"] == 1 & y[, "stop"] == t) / sum(exp(lp_train[in_risk]))
})
mitkof$runout_h0 <- mean(h0)


summary(cox_model)
plot_survival_paths(cox_model, nodes_df)
```

```{r}
suvival <- Surv(tstart, tstop, event)

```

### "Local" survival model with splines

```{r}
cox_gam_model <- coxph(Surv(tstart, tstop, event) ~ pspline(gradient) + 
                                                 pspline(tangential_curv) + 
                                                 pspline(logaccum),
                   data = nodes_df,
                   na.action = na.omit)

lp_train <- predict(cox_gam_model,type = "lp")
y <- cox_model$y
event_times <- sort(unique(y[, "stop"][y[, "status"] == 1]))
h0 <- sapply(event_times, function(t) {
  in_risk <- (y[, "start"] < t) & (y[, "stop"] >= t)
  sum(y[, "status"] == 1 & y[, "stop"] == t) / sum(exp(lp_train[in_risk]))
})
mitkof$runout_h0_gam <- mean(h0,na.rm = TRUE)
# summary(cox_gam_model)
plot_survival_paths(cox_gam_model, nodes_df)
```

### Survival model with some "memory"

Above, we modeled runout based only on local conditions. In reality, the upslope conditions will play an important role in determining the size and velocity of the flow, and therefore its likelihood of stopping soon. In this more sophisticated model, we'll track the conditions (gradient and curvature) of upslope nodes, which will help determine if its been gaining or losing momentum. We'll also track the starting gradient of the slide, which might help determine the initial size of the slide.

```{r}
library(zoo)  # for rollmean

nodes_df$start_grad <- start_grad[as.character(nodes_df$Polygon)]

nodes_df <- do.call(rbind, lapply(split(nodes_df, nodes_df$Polygon), function(df) {
  df$grad_roll <- sapply(seq_len(nrow(df)), function(i) {
    if (i == 1) df$gradient[1]
    else mean(df$gradient[max(1, i-4):(i)], na.rm = TRUE)
  })
  df$tan_roll <- sapply(seq_len(nrow(df)), function(i) {
    if (i == 1) df$tangential_curv[1]
    else mean(df$tangential_curv[max(1, i-15):(i)], na.rm = TRUE)
  })
  df
}))

nodes_df <- nodes_df[nodes_df$UpDist_m != 0, ]
# Fit time-varying CoxPH
cox_model_mem <- coxph(Surv(tstart, tstop, event) ~ grad_roll + tan_roll + start_grad + logaccum,
                   data = nodes_df,
                   na.action = na.exclude)

summary(cox_model_mem)
plot_survival_paths(cox_model_mem, nodes_df)
```

### Logistic regression stopping model

```{r}
# Create binary event column (1 = deposition at this node, 0 = flow continues)
nodes_df$event <- as.integer(nodes_df$event)  # should already exist
nodes_df$log_grad <- log10(nodes_df$gradient)
# mitkof$test_stack$log_grad <- log10(mitkof$test_stack$gradient)
mitkof$rstack$log_grad <- log10(mitkof$rstack$gradient)

# Fit logistic regression
logit_model <- glm(event ~ log_grad + tangential_curv + logaccum+dem,
                   data = nodes_df,
                   family = binomial)

summary(logit_model)
plot_survival_paths(logit_model, nodes_df)

# Compare concordance (AUC) between the two
library(pROC)
roc_cox   <- roc(nodes_df$event, predict(cox_model, type = "risk"))
roc_logit <- roc(nodes_df$event, predict(logit_model, type = "response"))

auc(roc_cox)
auc(roc_logit)
```

The model that considers upslope conditions boosted model concordance from 0.758 to 0.772.

## Hazard maps for test slices

make the hazard map using the local variables (Grad, Tan, log10accum)

```{r}
## Create a function to map the test prediction for any model
map_test_hazard <- function(model,island,inset=TRUE){
  stack <- island$test_stack
    # Compute cumulative survival
  if (inherits(model, "coxph")) {
    m_name <- 'Survival Analysis'
    lp <- predict(stack, model, type = "lp")
    hazard <- 1 - exp(-island$runout_h0 * exp(lp))
  } else if (inherits(model, "glm")) {
    m_name <- 'Logistic Regression'
    print('m_name is logit')
    hazard <- predict(stack, model, type = "response")
  }
  
  mars <- c(1, 1, 4, 1)
  plot(stack$hs, col = gray.colors(256), legend = FALSE,
       main = paste(island$name,m_name,"Hazard Map\n"),
       cex.main = 1, mar = mars)
  
  # plot(hazard, col = hcl.colors(256, "RdYlGn"), range=c(0,.06), add = TRUE, axes = FALSE, alpha = 0.35, mar = mars)
  surv = 1-hazard
  plot(surv, col = hcl.colors(256, "RdYlGn"), range=c(0.9,1), add = TRUE, axes = FALSE, alpha = 0.35, mar = mars)
  
  if (isTRUE(inset)) {
    rect(island$inset$xlim[1], island$inset$ylim[1],
         island$inset$xlim[2], island$inset$ylim[2],
         border = "black", lwd = 2, lty = 2)
  
    inset_ext <- ext(island$inset$xlim[1], island$inset$xlim[2],
                     island$inset$ylim[1], island$inset$ylim[2])
    hs_crop   <- crop(stack$hs,        inset_ext)
    hazard_crop <- crop(hazard,             inset_ext)
  
    tmp <- tempfile(fileext = ".png")
    png(tmp, width = 300, height = 300)
    par(mar = c(0, 0, 0, 0))
    image(hs_crop,   col = gray.colors(256), axes = FALSE, xlab = "", ylab = "")
    image(hazard_crop, col = adjustcolor(hcl.colors(256, "RdYlGn"), alpha.f = 0.5), add = TRUE)
    # plot(as.polygons(zone_crop, dissolve = TRUE),lwd = 2, add = TRUE, axes = FALSE, legend = FALSE
    dev.off()
  
    img <- png::readPNG(tmp)
  
    usr     <- par("usr")
    x_range <- usr[2] - usr[1]
    y_range <- usr[4] - usr[3]
    pad_x   <- x_range * 0.1
    pad_y   <- y_range * 0.05
    w       <- x_range * island$inset$size[1]
    h       <- y_range * island$inset$size[2]
    
  
    inset_pos <- switch(island$inset$loc,
      bottomright = c(usr[2] - w - pad_x, usr[3] + pad_y,usr[2] - pad_x, usr[3] + h + pad_y),
      bottomleft  = c(usr[1] + pad_x,     usr[3] + pad_y,     usr[1] + w + pad_x, usr[3] + h + pad_y),
      topright    = c(usr[2] - w - pad_x, usr[4] - h - pad_y, usr[2] - pad_x, usr[4] - pad_y),
      topleft     = c(usr[1] + pad_x,     usr[4] - h - pad_y, usr[1] + w + pad_x, usr[4] - pad_y)
    )
  
    rasterImage(img, inset_pos[1], inset_pos[2], inset_pos[3], inset_pos[4])
    rect(inset_pos[1], inset_pos[2], inset_pos[3], inset_pos[4], border = "black", lwd = 2)
  }

  invisible(NULL)
}
```

```{r}
# Logit hazard map - works cleanly
mitkof$test_stack$log_grad <- log10(mitkof$test_stack$gradient)
hazard_logit <- predict(mitkof$test_stack, logit_model, type = "response")
map_test_hazard(cox_model,mitkof,inset = FALSE)


# Cox hazard map - predict risk score only
# hazard_cox <- predict(mitkof$test_stack, cox_model, type = "risk")
```

# Runout predictions

Use a modified accumulatio

The chance of a flow leaving cell must be equal to the probability that a flow in the cell leaves the cell. The probability that there's a flow in the cell is the combined probability of flow entering the cell from upslope (could be multiple cells, btw) and the probability of a local initiation. 

flow_prob\[cell\] = (1 - (1-ls_prob\[cell\])\*(1-out_prob\[upslope_cells\]))

out_prob\[cell\] = surv\[cell\]\*flow_prob\[cell\]

alternatively, you could apply the survival only to the flow from upslope, since I'm not sure it makes sense that a flow can initiate and deposit in the same cell:

out_prob\[cell\] = (1 - (1-ls_prob\[cell\])\*(1-surv\[cell\]\*out_prob\[upslope_cells\]))

```{r}
# library(flowdem)
# 
# down_cell <- function(dem,flowdir){
#   # Precompute downstream index for every cell
#   nrows <- nrow(dem)
#   ncols <- ncol(dem)
#   n <- ncell(dem)
#   fdir_vals <- values(flowdir)
#   downstream <- integer(n)
#   
#   idx <- which(fdir_vals == 1);   downstream[idx] <- idx + 1
#   idx <- which(fdir_vals == 2);   downstream[idx] <- idx + ncols + 1
#   idx <- which(fdir_vals == 4);   downstream[idx] <- idx + ncols
#   idx <- which(fdir_vals == 8);   downstream[idx] <- idx + ncols - 1
#   idx <- which(fdir_vals == 16);  downstream[idx] <- idx - 1
#   idx <- which(fdir_vals == 32);  downstream[idx] <- idx - ncols - 1
#   idx <- which(fdir_vals == 64);  downstream[idx] <- idx - ncols
#   idx <- which(fdir_vals == 128); downstream[idx] <- idx - ncols + 1
#   
#   # Clip boundaries
#   downstream[downstream < 1 | downstream > n] <- NA
#   
#   rm(idx, fdir_vals); gc()
#   return(downstream)
# }
# 
# 
# runout_test <- function(model,island){
#   stack <- island$test_stack
#   df <- island$test_df
#   pred <- predi(island$gam_fit,df)
#   
#   ls_prob <- rast(stack$mask)
#   values(ls_prob) <- 0
#   ls_prob[df$ind] <- as.vector(pred)/50
#   
#   if (inherits(model, "coxph")) {
#     m_name <- 'Survival Analysis'
#     lp <- predict(stack, model, type = "lp")
#     hazard <- 1 - exp(-island$runout_h0 * exp(lp))
#   } else if (inherits(model, "glm")) {
#     m_name <- 'Logistic Regression'
#     hazard <- predict(stack, model, type = "response")
#   }
#   
#   flowdir <- terrain(stack$dem, v = "flowdir")
#   downstream <- down_cell(stack$dem,flowdir)
#   
#   # Initialize
#   flow_prob <- values(ls_prob)/10   # start with initiation probability
#   out_prob  <- numeric(ncell(stack$dem))
#   surv_vals <- 1 - values(hazard)  # probability of NOT stopping
#   
#   elev_vals <- values(stack$dem)
#   sorted_idx <- order(elev_vals, decreasing = TRUE, na.last = NA)
#   
#   
#   for (i in sorted_idx) {
#     flow_prob[i] <- surv_vals[i] * flow_prob[i]
#     j <- downstream[i]
#     if (!is.na(j)) {
#       flow_prob[j] <- 1 - (1 - flow_prob[j]) * (1 - flow_prob[i])
#     }
#   }
#   out_rast <- rast(stack$dem)
#   values(out_rast) <- flow_prob
#   return(out_rast)
# }
# 
# cox_runout <- runout_test(cox_model,mitkof)
```

```{r}
cppFunction('
NumericVector route_flow_cpp(IntegerVector flow_prob_int,
                              IntegerVector surv_vals_int,
                              IntegerVector fdir_vals,
                              IntegerVector sorted_idx,
                              int nrows, int ncols) {
  int n = flow_prob_int.size();
  std::vector<float> result(n);
  std::vector<float> surv(n);
  
  for (int i = 0; i < n; i++) {
    result[i] = flow_prob_int[i] == NA_INTEGER ? std::numeric_limits<float>::quiet_NaN() : flow_prob_int[i] / 10000000.0f;
    surv[i]   = surv_vals_int[i] == NA_INTEGER ? std::numeric_limits<float>::quiet_NaN() : surv_vals_int[i] / 10000000.0f;
  }
  
  for (int k = 0; k < sorted_idx.size(); k++) {
    int i = sorted_idx[k] - 1;
    
    if (std::isnan(result[i]) || std::isnan(surv[i])) continue;
    
    result[i] = surv[i] * result[i];
    
    int fdir = fdir_vals[i];
    int row = i / ncols;
    int col = i % ncols;
    int j = -1;
    
    if      (fdir == 1)   { if (col < ncols-1)                  j = i + 1; }
    else if (fdir == 2)   { if (row < nrows-1 && col < ncols-1) j = i + ncols + 1; }
    else if (fdir == 4)   { if (row < nrows-1)                  j = i + ncols; }
    else if (fdir == 8)   { if (row < nrows-1 && col > 0)       j = i + ncols - 1; }
    else if (fdir == 16)  { if (col > 0)                        j = i - 1; }
    else if (fdir == 32)  { if (row > 0 && col > 0)             j = i - ncols - 1; }
    else if (fdir == 64)  { if (row > 0)                        j = i - ncols; }
    else if (fdir == 128) { if (row > 0 && col < ncols-1)       j = i - ncols + 1; }
    
    if (j >= 0 && j < n && !std::isnan(result[j])) {
      result[j] = 1.0f - (1.0f - result[j]) * (1.0f - result[i]);
    }
  }
  
  NumericVector out(n);
  for (int i = 0; i < n; i++) {
    out[i] = std::isnan(result[i]) ? NA_REAL : (double)result[i];
  }
  return out;
}
')


runout <- function(init_pred, island, stack, hazard_mod){
  
  # make ls hazard map that is zero outside the mask (on gentler terrain) and na off island
  ls_prob <- ifel(is.na(stack$dem), NA, ifel(is.na(init_pred), 0, init_pred / 500))
  rm(init_pred); gc()
  flow_prob <- as.integer(values(ls_prob) * 10000000L)
  rm(ls_prob); gc()
  
  if (inherits(hazard_mod, "coxph")) {
    # Check if model contains splines
    has_spline <- any(sapply(names(hazard_mod$pterms), function(x) x != ""))
    
    if (has_spline) {
      # Chunk-based prediction to avoid empty chunk error with pspline
      pred_vars <- all.vars(hazard_mod$formula)
      pred_vars <- pred_vars[pred_vars %in% names(stack)]
      
      n         <- ncell(stack$dem)
      nchunks   <- 200
      chunk_size <- ceiling(n / nchunks)
      lp_vals   <- rep(NA_real_, n)
      
      for (i in seq_len(nchunks)) {
        idx_start <- (i - 1) * chunk_size + 1
        idx_end   <- min(i * chunk_size, n)
        chunk_n   <- idx_end - idx_start + 1
        
        # df_chunk <- as.data.frame(
        #   lapply(pred_vars, function(v) readValues(stack[[v]], idx_start, chunk_n)),
        #   col.names = pred_vars
        # )
        cells <- idx_start:idx_end
        df_chunk <- as.data.frame(
          lapply(pred_vars, function(v) stack[[v]][cells]),
          col.names = pred_vars
        )
        
        valid <- complete.cases(df_chunk)
        if (sum(valid) == 0) next
        
        lp_vals[idx_start:idx_end][valid] <- predict(hazard_mod, 
                                                      newdata = df_chunk[valid, ], 
                                                      type = "lp")
        rm(df_chunk); gc()
      }
      
      lp <- rast(stack$dem)
      values(lp) <- lp_vals
      rm(lp_vals); gc()
      
    } else {
      lp <- predict(stack, hazard_mod, type = "lp")
    }
    
    hazard <- 1 - exp(-island$runout_h0 * exp(lp))
    rm(lp); gc()
    
  } else if (inherits(hazard_mod, "glm")) {
    hazard <- predict(stack, hazard_mod, type = "response")
  }
  
  surv_vals <- as.integer((1 - values(hazard)) * 10000000L)
  rm(hazard); gc()
  
  fdir_vals  <- as.integer(values(stack$flowdir))
  sorted_idx <- as.integer(order(values(stack$dem), decreasing = TRUE, na.last = NA))
  gc()
  
  out_prob <- route_flow_cpp(
    flow_prob_int = flow_prob,
    surv_vals_int = surv_vals,
    fdir_vals     = fdir_vals,
    sorted_idx    = sorted_idx,
    nrows         = nrow(stack$dem),
    ncols         = ncol(stack$dem)
  )
  rm(flow_prob, surv_vals, fdir_vals, sorted_idx); gc()
  
  out_rast <- rast(stack$dem)
  values(out_rast) <- out_prob
  return(out_rast)
}
```

## Modeling runout for test slices

```{r}
# logit_runout <- runout_test(logit_model,mitkof)
# # plot(logit_runout)
# 
# cox_runout <- runout_test(cox_model,mitkof)
# plot(cox_runout)
# 
# cox_gam_runout <- runout_test(cox_gam_model,mitkof)
# plot(cox_gam_runout)


ls_prob <- predict(mitkof$test_stack,mitkof$gam_fit,type = "response")
ls_prob <- mask(ls_prob,mitkof$test_stack$mask)

logit_runout <- runout(ls_prob,mitkof,mitkof$test_stack,logit_model)
plot(logit_runout)

cox_runout <- runout(ls_prob,mitkof,mitkof$test_stack,cox_model)
plot(cox_runout)

cox_gam_runout <- runout(ls_prob,mitkof,mitkof$test_stack,cox_gam_model)
plot(cox_gam_runout)
```

# Applying to the full datasets

```{r}
pred <- rast("/Users/Jeff/Dropbox/Documents/TealWaters/modeling_results/prediction_maps/mitkof_gam_final2.tif")
out_rast <- runout(pred,mitkof,mitkof$rstack,cox_gam_model)

```

```{r}

# pred <- rast("/Users/Jeff/Dropbox/Documents/TealWaters/modeling_results/prediction_maps/mitkof_gam_final2.tif")
# 
# # make ls hazard map that is zero outside the mask (on gentler terrain) and na off island
# ls_prob <- ifel(is.na(mitkof$rstack$dem), NA,ifel(is.na(pred), 0, pred / 50))
# rm(pred); gc()
# flow_prob <- as.integer(values(ls_prob) * 10000000L)
# rm(ls_prob); gc()

lp <- predict(mitkof$rstack, cox_model, type = "lp")
hazard <- 1 - exp(-mitkof$runout_h0 * exp(lp))
rm(lp)
surv_vals <- as.integer((1 - values(hazard)) * 10000000L)
rm(hazard); gc()

fdir_vals <- as.integer(values(mitkof$rstack$flowdir))

sorted_idx <- as.integer(order(values(mitkof$rstack$dem), decreasing = TRUE, na.last = NA))

out_prob <- route_flow_cpp(
  flow_prob_int = flow_prob,
  surv_vals_int = surv_vals,
  fdir_vals     = fdir_vals,
  sorted_idx    = as.integer(sorted_idx),
  nrows         = nrow(mitkof$rstack$dem),
  ncols         = ncol(mitkof$rstack$dem)
)

rm(flow_prob, surv_vals, fdir_vals, sorted_idx); gc()

out_rast <- rast(mitkof$rstack$dem)
values(out_rast) <- out_prob

# writeRaster(out_rast,"/Users/Jeff/Dropbox/Documents/TealWaters/modeling_results/prediction_maps/mitkof_runout_cox.tif")
cox_runout <- runout(ls_prob,mitkof,mitkof$test_stack,cox_model)
```
