---
title: "Characterizing Initiation Locations"
author: "Dan"
date: today
format:
  html:
    code-fold: true
execute:
  warning: false
editor: visual
bibliography: references.bib
---

# Sample points

The issue now is, how to select the sample of points? How many, and how should they be divided between the initiation zones and everywhere else? Given that the initiation zones may cover a fraction of a percent of the study area, it is not immediately obvious how to divvy up the points. An assumption inherent in these classification methods is that the value at any sample point is independent of the value at any other point. Because these methods are ultimately determining proportions (e.g., what proportion of sampled points over a gradient interval of 0.7 to 0.8 fell within initiation zones), if some points are essentially sampling the same location (e.g., the same landslide scar), the calculated proportions will be biased towards values at those locations. How do we ensure that the sampled points provide independent values? Finally, how do our choices about sample selection affect the modeled probabilities?

## Independence

How far apart do the sample points need to be to ensure that each provides a measure of terrain attributes independent of the values at any other point? We can look at the degree of autocorrelation for topographic attributes using, e.g., a semivariogram, as we do below, but what does that actually tell us?

Consider a long, planar slope. The elevation changes in a regular fashion along the fall line, but not at all along a contour. The gradient (measured along the fall line) is about the same everywhere on the slope, so the degree of correlation is high. The gradient will decrease at the ridge crest and the base of the slope, so the range on the semivariogram should be about the length of the hillslope. Yet, we may find multiple, separate landslides initiating within lesser distances of each other on the slope. What terrain attributes might be affecting landslide initiation? If we exclude the ridge and base of the slope from calculation of the semivariance, we may find a smaller range (and sill) indicative of minor variations in gradient along the slope. Is this the appropriate separation distance to use?

Given our reliance on remotely sensed data, we cannot fully assess the factors that influence landslide initiation and it is probable that some of these factors are not visible with the data we have. These invisible-to-use factors may vary over length scales less than the variations visible in our remotely sensed data, even with high-resolution lidar-derived topography. The size distribution of the landslide scars themselves may offer some insight. Here are boxplots and cumulative frequency distributions of the estimated width of the initiation zones of the 244 landslide polygons used from the Tongass National Forest landslide inventory on Wrangell Island. Width is estimated both as the square root of the initiation-zone surface area and from the average width of the entire landslide polygon.

Both measures suggest that about 90% of all initiation zones are 50 m or less in width. Let's look at some potential predictors. Program samplePoints will read the specified rasters of predictors, find the range of predictor values that fall within the initiation-zone polygons, and create a mask file to exclude areas outside those ranges. This is similar to excluding the ridge top and valley floor with looking at gradient values in the planar-slope example described above. We are examining only those areas with the range of predictor values where landslide initiations were observed. The code chunks below will run samplePoints to create a mask raster and then use the cv_spatial_autocor function from the blockCV package to calculate semivariograms for each predictor raster. The semivariogram will indicate the range of spatial autocorrelation for each predictor.

```{r}
library(blockCV)
library(automap)
library(sf)
library(terra)
library(stringr)
library(TerrainWorksUtils)
library(ggplot2)
library(patchwork)
library(data.table)

dataFolder <- "c:/work/data/wrangell/"

rasterList <- read.csv(paste0(dataFolder, "R4rasterList.csv"), )
R4rasters <- vector("list",nrow(rasterList)-1)

for (i in 1:nrow(rasterList)) {

  if (str_detect(rasterList[i,1], "MASK")) {
    outMask <- rasterList[i,2]
  } else {
    r <- vector("list", 6)
    for (j in 1:6) {
      r[[j]] <- trimws(rasterList[i,j])
    }
    R4rasters[[i]] <- r
  }
}

for (i in 1:length(R4rasters)) {
  r <- terra::rast(paste0(R4rasters[[i]][2],".flt"))
  names(r) <- R4rasters[[i]][1]
  if (i == 1) {
    rstack <- r
  } else {
    rstack <- c(rstack, r)
  }
}

rasterList <- read.csv(paste0(dataFolder, "I4rasterList.csv"))
I4rasters <- vector("list", nrow(rasterList))

for (i in 1:nrow(rasterList)) {
  r <- vector("list",3)
  r[[1]] <- trimws(rasterList[i,1])
  r[[2]] <- trimws(rasterList[i,2])
  r[[3]] <- trimws(rasterList[i,3])
  I4rasters[[i]] <- r
}

# Run sample points to build a raster masking zones outside the range of predictor values
# associated with landslide initiation zones. Edit input variables as needed
work <- "c:/work/data/wrangell/"
inRaster <- paste0(dataFolder,"init") # initiation zones, raster created by LS_poly above
areaPerSample <- 10000. # within initiation zones, one point every areaPerSample square meters
buffer_in <- 10. # buffer around sample points inside initiation zones, in meters
buffer_out <- 3. # buffer around sample points outside initiation zones
margin <- 1. # margin around initiation zones to preclude sample points close to the edge, in meters
ratio <- 3000. # ratio of sample points outside initiation zones to those within
nbins <- 100 # number of bins for building histograms

minPatch <- 100000.
inPoints <- paste0(dataFolder, "inpoints1000")
outPoints <- paste0(dataFolder, "outpoints") # used below for spatstat
outInit <- paste0(dataFolder, "initMod")
outInitPoints <- paste0(dataFolder, "initPoints")
table <- paste0(dataFolder, "table")
scratchDir <- "c:/work/scratch"
executableDir <- "c:/work/sandbox/landslideutilities/projects/samplepoints/x64/release"

returnCode <- TerrainWorksUtils::samplePoints(
  inRaster,
  areaPerSample,
  buffer_in,
  buffer_out,
  margin,
  ratio,
  nbins,
  R4rasters,
  I4rasters,
  minPatch,
  inPoints,
  outPoints,
  outMask,
  outInit,
  outInitPoints,
  table,
  scratchDir,
  executableDir)

if (returnCode != 0) {
  stop("Error in samplePoints, return code: ", returnCode)
}

pcum_in1 <- vector(mode = "list", length = length(R4rasters))
pcum_out1 <- vector(mode = "list", length = length(R4rasters))
pdif1 <- vector(mode = "list", length = length(R4rasters))
pbar1 <- vector(mode = "list", length = length(R4rasters))
pbar_in1 <- vector(mode = "list", length = length(R4rasters))
pbar_out1 <- vector(mode = "list", length = length(R4rasters))
density <- vector(mode = "list", length = length(R4rasters))
dencum <- vector(mode = "list", length = length(R4rasters))
rocplot <- vector(mode = "list", length = length(R4rasters))
auc <- data.table(name=character(), val=numeric())
roc_dt <- data.table()
xrange <- data.table(start=numeric(),end=numeric())

for (i in 1:length(R4rasters)) {
  name <- R4rasters[[i]][[1]]
  
  d_in <- as.data.table(read.csv(paste0(table, "_", name, "_in.csv")))
  
  d_out <- as.data.table(read.csv(paste0(table, "_", name, "_out.csv")))
 
  sum_in <- d_in[, sum(Number)]
  sum_out <- d_out[, sum(Number)]
  
  pbar1[[i]] <- ggplot() +
    geom_bar(data=d_in, 
             aes(x = Mid, y = Number/sum_in, fill = "Inside"), 
             stat = "identity",
             alpha = 0.6) +
    geom_bar(data=d_out,
             aes(x = Mid, y = Number/sum_out, fill = "Outside"),
             stat = "identity",
             alpha = 0.6) +
    labs(x = name,
         y = "Proportion") +
    theme(legend.title = element_blank(),
          legend.position = "inside",
          legend.position.inside = c(0.9, 0.8)) 
  
  pdif1[[i]] <- ggplot() +
    geom_line(data=d_in, 
              aes(x=Upper, y = CumulativeProp, color = "Inside"),
              linewidth = 1.2) +
    geom_line(data=d_out,
              aes(x = Upper, y = CumulativeProp, color = "Outside"),
              linewidth = 1.2) +
    labs(x = name,
         y = "Cumulative Proportion") +
    theme(legend.title = element_blank(),
          legend.position = "inside",
          legend.position.inside = c(0.85, 0.2))
  
# Calculate density as area of initiation zone per unit area of the study area (the masked area)
  den <- d_in[, .(Upper, Number)]
  den <- den[, Nout := d_out[, Number]]
  den <- den[Nout > 0, ]
  den <- den[, density := Number/Nout]

  xrange <- rbind(xrange, list(min(den$Upper),max(den$Upper)))
  
# Generate an ROC curve
  roc <- den[, .(density, Number, Nout)]
  roc <- roc[density > 0,]
  roc <- roc[order(-density)]
  roc <- roc[, tpr := rev(cumsum(Number))] # true positive rate
  tprmax <- max(roc$tpr)
  roc <- roc[, tpr := tpr/tprmax]
  roc <- roc[, fpr := rev(cumsum(Nout))] # false positive rate
  fprmax <- max(roc$fpr)
  roc <- roc[, fpr := fpr/fprmax]
  roc <- roc[, Index := as.factor(name)]
  roc_dt <- rbind(roc_dt,roc)

# Calculate area under the ROC curve  
  s = 0
  for (j in 1:(nrow(roc)-1)) {
    s_tpr <- (roc[j]$tpr + roc[j+1]$tpr) / 2
    s_fpr <- roc[j]$fpr - roc[j+1]$fpr
    s <- s + (s_tpr * s_fpr)
  }
  thisname <- R4rasters[[i]][1]
  auc <- rbind(auc, list(thisname, s))
  
  density[[i]] <- ggplot(data=den) + 
    geom_point(aes(x=Upper, y=density), shape = 1, size = 2.8) +
    geom_smooth(aes(x=Upper,y=density), method="loess",se=FALSE, span=0.2, color="black", linewidth = 1.) +
    labs(x = name, y = "Density") +
    coord_cartesian(xlim=c(xrange[i]$start,xrange[i]$end))

  den <- den[order(density)]
  den <- den[, cumulative := cumsum(density)]  
  denmax <- max(den$cumulative)  
  den <- den[, cumulative := cumulative/denmax]
  
  dencum[[i]] <- ggplot(den, aes(x=Upper, y=cumulative)) +
    geom_point(shape = 20, size = 1.3) + 
    geom_line(color = "black", linewidth = 0.5) +
    labs(title = name,
         x = "Density", 
         y = "Proportion of area")
  
  rocplot[[i]] <- ggplot(roc, aes(x=fpr, y=tpr)) +
    geom_line(color = "black", linewidth = 0.8) +
    geom_point(shape = 1, size = 4) +
    geom_segment(x=0,y=0,xend=1,yend=1, color="gray", linewidth=1.) +
    coord_cartesian(x=c(0,1)) +
    labs(title = name,
         subtitle = paste0("AUC = ", round(s, 3)),
         x = "False Positive Rate (Proportion of masked area)",
         y = "True Positive Rate (Proportion of initiation-zone area)")

}
```

```{r}
# Build a semivariogram for each predictor.
var <- cv_spatial_autocor(r=rstack, num_sample=100000)
var$range_table
```

Gradient was calculated over a diameter of 30 meters, thus smoothing variations over length scales less than that. The range of the semivariogram for this measure of gradient is about 300 meters, which is around the midpoint of hillslope lengths included within the mask raster. The two flow accumulation rasters, PCA_5 and PCA_72, indicate the approximate contributing area for each DEM cell for a 5 and 72 hour storm, assuming a saturated hydraulic conductivity of 1 m/sec and ignoring infiltration time. PCA_5 is highly correlated with gradient and has a similar semivariogram range. PCA_72 has flow concentrations within channels and exhibits much larger variability than PCA_5, indicated by its larger range and sill. The factor of safety values (FoS) are derived from the gradient and flow accumulation values. Despite the large differences between PCA_5 and PCA_72, the FoS rasters based on these exhibit similar patterns of variability with nearly equal ranges and sills.

What do these range values imply for our strategy of sampling independent points? They are all several times larger than the typical scale of a landslide initiation zone. Even if we use the \~300-m range found for gradient as the minimum point spacing, that precludes more than one sample point per landslide initiation zone. Ultimately the goal is to have a representative sample that accurately reflects the range and frequency distribution of predictor values across both the observed initiation zones and the masked area spanning the range of values observed within the initiation zones. Independence precludes bias in our representation of those frequency distributions, but bias will still be an issue if we have too few points.

## Number

How many randomly placed sample points do we need to obtain an accurate and complete representation of the frequency distribution of predictor values? That depends on how uniform the frequency distributions are. Long-tailed distributions require more points to ensure that rare conditions get samples. We can compare frequency distributions of sample-point sets to those of the predictor rasters. This is straight forward for single predictors, but becomes more complicated when looking at combinations of predictors.

Representative sample, curse of dimensionality.

### Single predictors

Let's start with one sample point per initiation zone and a point spacing of 300m elsewhere. We can generate point samples with program samplePoints.

```{r}
# Edit input variables as needed
inRaster <- "c:/work/data/wrangell/init" # initiation zones, raster created by LS_poly
areaPerSample <- 282743. # within initiation zones, one point every areaPerSample square meters
buffer_in <- 300. # buffer around sample points, in meters
buffer_out <- 300. # buffer around sample points outside initiation zones
margin <- 1. # margin around initiation zones to preclude sample points close to the edge, in meters
ratio <- 3. # ratio of sample points outside initiation zones to those within
nbins <- 100 # number of bins for building histograms

minPatch <- 100000.
inPoints <- "c:/work/data/wrangell/inpoints_1"
outPoints <- "c:/work/data/wrangell/outpoints_1"
outMask <- "c:/work/data/wrangell/mask"
outInit <- "c:/work/data/wrangell/initMod"
outInitPoints <- ""
table <- "c:/work/data/wrangell/table1"
scratchDir <- "c:/work/scratch"
executableDir <- "c:/work/sandbox/landslideutilities/projects/samplepoints/x64/release"

returnCode <- TerrainWorksUtils::samplePoints(
  inRaster,
  areaPerSample,
  buffer_in,
  buffer_out,
  margin,
  ratio,
  nbins,
  R4rasters,
  I4rasters,
  minPatch,
  inPoints,
  outPoints,
  outMask,
  outInit,
  outInitPoints,
  table,
  scratchDir,
  executableDir)
```

I found building cumulative distributions with R to be rather slow, so program samplePoints also outputs a csv file with a histogram for each of the predictors. We can use these to compare the frequency distributions of the sample points to those of the predictor rasters. The files have five columns:

1.  Lower, the lower bound of the interval

2.  Upper, the upper bound of the interval

3.  Mid, the midpoint of the interval

4.  Number, the number of DEM cells or sample points included in the interval

5.  CumulativeProp, the cumulative proportion of cells or samples

There are three things of interest, examined in the follows plots:

1.  How different, or similar, are the frequency distributions for areas inside and outside of the initiation zones? This difference is characterized using the landslide density as a function of the predictor value, here calculated in terms of initiation-zone-area/total-area (limited only to the masked region). Density is calculated for each interval used to build the histograms.

2.  How well do the frequency distributions for the sample points match those of the area sampled?

3.  How well does the variation in landslide density indicated by the sample match the actual landslide density measured using all DEM grid points?

```{r}

pcum_in1 <- vector(mode = "list", length = length(R4rasters))
pcum_out1 <- vector(mode = "list", length = length(R4rasters))
pbar_in1 <- vector(mode = "list", length = length(R4rasters))
pbar_out1 <- vector(mode = "list", length = length(R4rasters))
densample1 <- vector(mode = "list", length = length(R4rasters))
dencum <- vector(mode = "list", length = length(R4rasters))
denscum1 <- vector(mode = "list", length = length(R4rasters))

for (i in 1:length(R4rasters)) {
#for (i in 2:3) {
  name <- R4rasters[[i]][[1]]
  
  d_in <- as.data.table(read.csv(paste0(table, "_", name, "_in.csv")))
  
  d_out <- as.data.table(read.csv(paste0(table, "_", name, "_out.csv")))
  
  d_sampleIn <- as.data.table(read.csv(paste0(table, 
                                              "_", 
                                              name, 
                                              "_inSample.csv")))
  
  d_sampleOut <- as.data.table(read.csv(paste0(table,
                                               "_",
                                               name,
                                               "_outSample.csv")))
 
  sum_in <- d_in[, sum(Number)]
  sum_out <- d_out[, sum(Number)]
  sum_sampleIn <- d_sampleIn[, sum(Number)]
  sum_sampleOut <- d_sampleOut[, sum(Number)]
  
  pbar_in1[[i]] <- ggplot() +
    geom_bar(data=d_sampleIn, 
             aes(x=Mid, y=Number/sum_sampleIn, fill="Sample Points"), 
             stat="identity", 
             alpha=0.6) +
    geom_bar(data=d_in, 
             aes(x=Mid, y=Number/sum_in, fill="Total Area"), 
             stat="identity", 
             alpha=0.3) +
    labs(x = name,
         y = "Proportion") +
    theme(legend.title = element_blank(),
          legend.position = "inside",
          legend.position.inside = c(0.9, 0.8))
  
  pbar_out1[[i]] <- ggplot() +
    geom_bar(data=d_sampleOut, 
             aes(x=Mid, y=Number/sum_sampleOut, fill="Sample Points"), 
             stat="identity", 
             alpha=0.6) +
    geom_bar(data=d_out, 
             aes(x=Mid, y=Number/sum_out, fill="Total Area"), 
             stat="identity", 
             alpha=0.6) +
    labs(x = name,
         y = "Proportion") +
    theme(legend.title = element_blank(),
          legend.position = "inside",
          legend.position.inside = c(0.9, 0.8))
  
  pcum_in1[[i]] <- ggplot() + 
    geom_line(data = d_in, 
              aes(x = Upper, y = CumulativeProp, color = "Total Area"),
              linewidth = 1.2) +
    geom_line(data=d_sampleIn, 
              aes(x=Upper, y=CumulativeProp, color="Sample Points"), 
              linewidth=1.2) +
    labs(x = name,
         y = "Cumulative Proportion") +
    theme(legend.title = element_blank(),
          legend.position = "inside",
          legend.position.inside = c(0.9,0.2))
  
  pcum_out1[[i]] <- ggplot() +
    geom_line(data=d_out, 
              aes(x=Upper, y=CumulativeProp, color="Total Area"), 
              linewidth=1.2) +
    geom_line(data=d_sampleOut, 
              aes(x=Upper, y=CumulativeProp, color="Sample Points"), 
              linewidth=1.2) +
    labs(x = name,
         y = "Cumulative Proportion") +
    theme(legend.title = element_blank(),
          legend.position = "inside",
          legend.position.inside = c(0.9, 0.2))
  
  den <- d_in[, .(Upper, Number)]
  den <- den[, Nout := d_out[, Number]]
  den <- den[Nout > 0, ]
  den <- den[, density := Number/Nout]

  # Calculate density based on the sample points
  dens <- d_sampleIn[, .(Upper, Number)]
  dens <- dens[, Nout := d_sampleOut[, Number]]
  dens <- dens[, Nout := Nout + Number]
  dens <- dens[Nout > 0,]
  dens <- dens[, density := Number/Nout]

  densample1[[i]] <- ggplot(data=dens) +
    geom_point(aes(x=Upper, y=density), shape = 1, size = 2.5) +
    geom_smooth(aes(x=Upper,y=density), method="loess",se=FALSE,span=0.2,color="black",linewidth=1.) +
    labs(x = name, y = "Density") +
    coord_cartesian(xlim=c(xrange[i]$start,xrange[i]$end))
  
  den <- den[, cumulative := cumsum(density)]  
  denmax <- max(den$cumulative)  
  den <- den[, cumulative := cumulative/denmax]
  
  dens <- dens[, cumulative := cumsum(density)]
  denmax <- max(dens$cumulative)
  dens <- dens[, cumulative := cumulative/denmax]  
  
  denscum1[[i]] <- ggplot() +
    geom_point(data=den, aes(x=Upper, y=cumulative, color = "Total Area"), shape = 20, size = 1.8) +
    geom_point(data=dens, aes(x=Upper, y=cumulative, color = "Sample"), shape = 20, size = 1.8) + 
    geom_line(data=den, aes(x=Upper, y=cumulative, color = "Total Area"), linewidth = 0.5) +
    geom_line(data=dens, aes(x=Upper, y=cumulative, color = "Sample"), linewidth = 0.5) +
    labs(title = name,
         x = "Density", 
         y = "Proportion of area") +
    theme(legend.title = element_blank(),
          legend.position = "inside",
          legend.position.inside = c(0.8,0.2))

}
```

The first set of plots below contrasts the frequency distribution of each predictor inside initiation zones to the distribution over the entire masked study area. Differences in the two distributions are highlighted by variation in landslide density.

The next set of plots look at how well the sampled points characterize the entire frequency distribution of predictor values, both inside the initiation zones and over the entire study area.

```{r}
#| label: inside_1
#| 
for (i in 1:length(R4rasters)) {
  thisplot <- pbar_in1[[i]] / pcum_in1[[i]]
  thisplot <- thisplot + plot_annotation(
    title = paste0("Comparison of sample points and total area inside initiation zones for ", R4rasters[[i]][[1]]),
    subtitle = "One sample point per landslide, n = 236",
    caption = "Bar plots show the proportion of area in each interval, cumulative plots show the cumulative proportion of area")
  plot(thisplot)
}
```

```{r}

#| label: outside_1
for (i in 1:length(R4rasters)) {
  thisplot <- pbar_out1[[i]] / pcum_out1[[i]]
  thisplot <- thisplot + plot_annotation(
    title = paste0("Comparison of sample points and total area outside initiation zones for ", R4rasters[[i]][[1]]),
    subtitle = "Three times more samples outside than inside, n = 708",
    caption = "Bar plots show the proportion of area in each interval, cumulative plots show the cumulative proportion of area")
  plot(thisplot)
}
```

Landslide density is the metric used to rank zones in terms of susceptibility to landslide initiation. These plots show how density changes over the range of each of the predictors examined here. I have plotted the density values for all predictors using the same vertical scale, so it is clear that gradient provides the greatest ability to distinguish high-density from low-density zones.

```{r}
#| label: density_1
for (i in 1:length(R4rasters)) {
  thisplot1 <- density[[i]] / densample1[[i]]
  thisplot <- thisplot1 | denscum1[[i]]
  thisplot <- thisplot + plot_annotation(
    title = paste0("Landslide density for ", R4rasters[[i]][[1]]),
    subtitle = "Total study area (upper left) vs sampled points (lower left) and cumulative distributions (right)",
    caption = "Density calculated as the ratio of the number of points inside initiation zones to points outside for each histogram interval")
  plot(thisplot)
}
```

We want to accurately characterize variations in density with our sample, which these clearly do not.

Let's increase the number of sample points inside initiation zones. The previous examples had one point per initiation zone, regardless of its size. In the next example, we seek one point per every 314 m<sup>2</sup> of initiation zone area and require a 10-m buffer around points inside the initiation zones. The spacing of points outside the initiation zones is reduced to a minimum of 100 m and we still require three times as many points outside the initiation zones as inside.

```{r}
# Edit input variables as needed
inRaster <- "c:/work/data/wrangell/init" # initiation zones, raster created by LS_poly
areaPerSample <- 315. # within initiation zones, one point every areaPerSample square meters
buffer_in <- 10. # buffer around sample points, in meters
buffer_out <- 100.
margin <- 1. # margin around initiation zones to preclude sample points close to the edge, in meters
ratio <- 3. # ratio of sample points outside initiation zones to those within
nbins <- 100 # number of bins for building histograms

minPatch <- 100000.
inPoints <- "c:/work/data/wrangell/inpoints_2"
outPoints <- "c:/work/data/wrangell/outpoints_2"
outMask <- "c:/work/data/wrangell/mask"
outInit <- "c:/work/data/wrangell/initMod"
table <- "c:/work/data/wrangell/table2"
scratchDir <- "c:/work/scratch"
executableDir <- "c:/work/sandbox/landslideutilities/projects/samplepoints/x64/release"

returnCode <- TerrainWorksUtils::samplePoints(
  inRaster,
  areaPerSample,
  buffer_in,
  buffer_out,
  margin,
  ratio,
  nbins,
  R4rasters,
  I4rasters,
  minPatch,
  inPoints,
  outPoints,
  outMask,
  outInit,
  outInitPoints,
  table,
  scratchDir,
  executableDir)


pcum_in2 <- vector(mode = "list", length = length(R4rasters))
pcum_out2 <- vector(mode = "list", length = length(R4rasters))
pbar_in2 <- vector(mode = "list", length = length(R4rasters))
pbar_out2 <- vector(mode = "list", length = length(R4rasters))
densample2 <- vector(mode = "list", length = length(R4rasters))
denscum2 <- vector(mode = "list", length = length(R4rasters))

for (i in 1:length(R4rasters)) {
  name <- R4rasters[[i]][[1]]
  
  d_in <- as.data.table(read.csv(paste0(table, "_", name, "_in.csv")))
  
  d_out <- as.data.table(read.csv(paste0(table, "_", name, "_out.csv")))  
  
  d_sampleIn <- as.data.table(read.csv(paste0(table,"_",name,"_inSample.csv")))
  
  d_sampleOut <- as.data.table(read.csv(paste0(table,"_",name,"_outSample.csv")))
 
  sum_sampleIn <- d_sampleIn[, sum(Number)]
  sum_sampleOut <- d_sampleOut[, sum(Number)]
  
  pbar_in2[[i]] <- ggplot() +
    geom_bar(data=d_sampleIn, 
             aes(x=Mid, y=Number/sum_sampleIn, fill="Sample Points"), 
             stat="identity", 
             alpha=0.6) +
    geom_bar(data=d_in, 
             aes(x=Mid, y=Number/sum_in, fill="Total Area"), 
             stat="identity", 
             alpha=0.3) +
    labs(x = name,
         y = "Proportion") +
    theme(legend.title = element_blank(),
          legend.position = "inside",
          legend.position.inside = c(0.9, 0.8))
  
  pbar_out2[[i]] <- ggplot() +
    geom_bar(data=d_sampleOut, 
             aes(x=Mid, y=Number/sum_sampleOut, fill="Sample Points"), 
             stat="identity", 
             alpha=0.6) +
    geom_bar(data=d_out, 
             aes(x=Mid, y=Number/sum_out, fill="Total Area"), 
             stat="identity", 
             alpha=0.6) +
    labs(x = name,
         y = "Proportion") +
    theme(legend.title = element_blank(),
          legend.position = "inside",
          legend.position.inside = c(0.9, 0.8))
  
  pcum_in2[[i]] <- ggplot() + 
    geom_line(data = d_in, 
              aes(x = Upper, y = CumulativeProp, color = "Total Area"),
              linewidth = 1.2) +
    geom_line(data=d_sampleIn, 
              aes(x=Upper, y=CumulativeProp, color="Sample Points"), 
              linewidth=1.2) +
    labs(x = name,
         y = "Cumulative Proportion") +
    theme(legend.title = element_blank(),
          legend.position = "inside",
          legend.position.inside = c(0.9,0.2))
  
  pcum_out2[[i]] <- ggplot() +
    geom_line(data=d_out, 
              aes(x=Upper, y=CumulativeProp, color="Total Area"), 
              linewidth=1.2) +
    geom_line(data=d_sampleOut, 
              aes(x=Upper, y=CumulativeProp, color="Sample Points"), 
              linewidth=1.2) +
    labs(x = name,
         y = "Cumulative Proportion") +
    theme(legend.title = element_blank(),
          legend.position = "inside",
          legend.position.inside = c(0.9, 0.2))
  
  den <- d_in[, .(Upper, Number)]
  den <- den[, Nout := d_out[, Number]]
  den <- den[Nout > 0, ]
  den <- den[, density := Number/Nout]
  den <- den[, cumulative := cumsum(density)]  
  denmax <- max(den$cumulative)  
  den <- den[, cumulative := cumulative/denmax]
  
  dens <- d_sampleIn[, .(Upper, Number)]
  dens <- dens[, Nout := d_sampleOut[, Number]]
  dens <- dens[, Nout := Nout + Number]
  dens <- dens[Nout > 0,]
  dens <- dens[, density := Number/Nout]
  dens <- dens[, cumulative := cumsum(density)]
  denmax <- max(dens$cumulative)
  dens <- dens[, cumulative := cumulative/denmax]
  
  densample2[[i]] <- ggplot(dens, aes(x=Upper, y=density)) + 
    geom_point(shape = 1, size = 2.5) +
    geom_smooth(aes(x=Upper,y=density), method="loess",se=FALSE,span=0.2,color="black",linewidth=1.) +
    labs(x = name, y = "Density") +
    coord_cartesian(xlim=c(xrange[i]$start,xrange[i]$end))
  
  denscum2[[i]] <- ggplot() +
    geom_point(data=den, aes(x=Upper, y=cumulative, color = "Total Area"), shape = 20, size = 1.8) +
    geom_point(data=dens, aes(x=Upper, y=cumulative, color = "Sample"), shape = 20, size = 1.8) + 
    geom_line(data=den, aes(x=Upper, y=cumulative, color = "Total Area"), linewidth = 0.5) +
    geom_line(data=dens, aes(x=Upper, y=cumulative, color = "Sample"), linewidth = 0.5) +
    labs(title = name,
         x = "Density", 
         y = "Proportion of area") +
    theme(legend.title = element_blank(),
          legend.position = "inside",
          legend.position.inside = c(0.8,0.2))

}
```

```{r}
#| label: inside_2
for (i in 1:length(R4rasters)) {
  thisplot <- pbar_in2[[i]] / pcum_in2[[i]]
  thisplot <- thisplot + plot_annotation(
    title = paste0("Comparison of sample points and total area inside initiation zones for ", R4rasters[[i]][[1]]),
    subtitle = "One sample per 315 square meters, 10-m buffer around points, n = 501",
    caption = "Bar plots show the proportion of area in each interval, cumulative plots show the cumulative proportion of area")
  plot(thisplot)
}
```

```{r}
#| label: outside_2
for (i in 1:length(R4rasters)) {
  thisplot <- pbar_out2[[i]] / pcum_out2[[i]]
  thisplot <- thisplot + plot_annotation(
    title = paste0("Comparison of sample points and total area outside initiation zones for ", R4rasters[[i]][[1]]),
    subtitle = "Three times more samples outside than inside, n = 1503",
    caption = "Bar plots show the proportion of area in each interval, cumulative plots show the cumulative proportion of area")
  plot(thisplot)
}
```

```{r}
#| label: density_2
for (i in 1:length(R4rasters)) {
  thisplot1 <- density[[i]] / densample2[[i]]
  thisplot <- thisplot1 | denscum2[[i]]
  thisplot <- thisplot + plot_annotation(
    title = paste0("Landslide density for ", R4rasters[[i]][[1]]),
    subtitle = "Total study area (upper left) vs sampled points (lower left) and cumulative distributions (right)",
    caption = "Density calculated as the ratio of the number of points inside initiation zones to points outside for each histogram interval")
  plot(thisplot)
}
```

Increase the number of points inside the initiation zones again.

```{r}
# Edit input variables as needed
inRaster <- "c:/work/data/wrangell/init" # initiation zones, raster created by LS_poly
areaPerSample <- 15. # within initiation zones, one point every areaPerSample square meters
buffer_in <- 2.5 # buffer around sample points, in meters
buffer_out <- 25.
margin <- 1. # margin around initiation zones to preclude sample points close to the edge, in meters
ratio <- 15. # ratio of sample points outside initiation zones to those within
nbins <- 100 # number of bins for building histograms

minPatch <- 100000.
inPoints <- "c:/work/data/wrangell/inpoints_3"
outPoints <- "c:/work/data/wrangell/outpoints_3"
outMask <- "c:/work/data/wrangell/mask"
outInit <- "c:/work/data/wrangell/initMod"
table <- "c:/work/data/wrangell/table3"
scratchDir <- "c:/work/scratch"
executableDir <- "c:/work/sandbox/landslideutilities/projects/samplepoints/x64/release"

returnCode <- TerrainWorksUtils::samplePoints(
  inRaster,
  areaPerSample,
  buffer_in,
  buffer_out,
  margin,
  ratio,
  nbins,
  R4rasters,
  I4rasters,
  minPatch,
  inPoints,
  outPoints,
  outMask,
  outInit,
  outInitPoints,
  table,
  scratchDir,
  executableDir)
```

```{r}

pcum_in3 <- vector(mode = "list", length = length(R4rasters))
pcum_out3 <- vector(mode = "list", length = length(R4rasters))
pbar_in3 <- vector(mode = "list", length = length(R4rasters))
pbar_out3 <- vector(mode = "list", length = length(R4rasters))
densample3 <- vector(mode = "list", length = length(R4rasters))
denscum3 <- vector(mode = "list", length = length(R4rasters))
denshold <- vector(mode = "list", length = length(R4rasters))

for (i in 1:length(R4rasters)) {
  name <- R4rasters[[i]][[1]]
  
  d_in <- as.data.table(read.csv(paste0(table, "_", name, "_in.csv")))
  
  d_out <- as.data.table(read.csv(paste0(table, "_", name, "_out.csv")))  
  
  d_sampleIn <- as.data.table(read.csv(paste0(table, 
                                              "_", 
                                              name, 
                                              "_inSample.csv")))
  
  d_sampleOut <- as.data.table(read.csv(paste0(table,
                                               "_",
                                               name,
                                               "_outSample.csv")))
 
  sum_sampleIn <- d_sampleIn[, sum(Number)]
  sum_sampleOut <- d_sampleOut[, sum(Number)]
  
  pbar_in3[[i]] <- ggplot() +
    geom_bar(data = d_sampleIn, 
             aes(x=Mid, y=Number/sum_sampleIn, fill="Sample Points"), 
             stat="identity", 
             alpha=0.6) +
    geom_bar(data = d_in, 
             aes(x=Mid, y=Number/sum_in, fill="Total Area"), 
             stat="identity", 
             alpha=0.3) +
    labs(x = name,
         y = "Proportion") +
    theme(legend.title = element_blank(),
          legend.position = "inside",
          legend.position.inside = c(0.9, 0.8))
  
  pbar_out3[[i]] <- ggplot() +
    geom_bar(data=d_sampleOut, 
             aes(x=Mid, y=Number/sum_sampleOut, fill="Sample Points"), 
             stat="identity", 
             alpha=0.6) +
    geom_bar(data=d_out, 
             aes(x=Mid, y=Number/sum_out, fill="Total Area"), 
             stat="identity", 
             alpha=0.6) +
    labs(x = name,
         y = "Proportion") +
    theme(legend.title = element_blank(),
          legend.position = "inside",
          legend.position.inside = c(0.9, 0.8))
  
  pcum_in3[[i]] <- ggplot() + 
    geom_line(data = d_in, 
              aes(x = Upper, y = CumulativeProp, color = "Total Area"),
              linewidth = 1.2) +
    geom_line(data=d_sampleIn, 
              aes(x=Upper, y=CumulativeProp, color="Sample Points"), 
              linewidth=1.2) +
    labs(x = name,
         y = "Cumulative Proportion") +
    theme(legend.title = element_blank(),
          legend.position = "inside",
          legend.position.inside = c(0.9,0.2))
  
  pcum_out3[[i]] <- ggplot() +
    geom_line(data=d_out, 
              aes(x=Upper, y=CumulativeProp, color="Total Area"), 
              linewidth=1.2) +
    geom_line(data=d_sampleOut, 
              aes(x=Upper, y=CumulativeProp, color="Sample Points"), 
              linewidth=1.2) +
    labs(x = name,
         y = "Cumulative Proportion") +
    theme(legend.title = element_blank(),
          legend.position = "inside",
          legend.position.inside = c(0.9, 0.2))

  den <- d_in[, .(Upper, Number)]
  den <- den[, Nout := d_out[, Number]]
  den <- den[Nout > 0, ]
  den <- den[, density := Number/Nout]
  den <- den[, cumulative := cumsum(density)]  
  denmax <- max(den$cumulative)  
  den <- den[, cumulative := cumulative/denmax]
  
  dens <- d_sampleIn[, .(Upper, Number)]
  dens <- dens[, Nout := d_sampleOut[, Number]]
  dens <- dens[, Nout := Nout + Number]
  dens <- dens[Nout > 0,]
  dens <- dens[, density := Number/Nout]
  dens <- dens[, cumulative := cumsum(density)]
  denmax <- max(dens$cumulative)
  dens <- dens[, cumulative := cumulative/denmax]
  dens <- dens[, lnodds := log(density/(1-density))]
  denshold[[i]] <- dens
  
  densample3[[i]] <- ggplot(dens, aes(x=Upper, y=density)) + 
    geom_point(shape = 1, size = 2.5) +
    geom_smooth(aes(x=Upper,y=density), 
                method="loess",
                se=FALSE,
                span=0.2,
                color="black",
                linewidth=1.) +
    labs(x = name, y = "Density") +
    coord_cartesian(xlim=c(xrange[i]$start,xrange[i]$end))
  
  denscum3[[i]] <- ggplot() +
    geom_point(data=den, aes(x=Upper, y=cumulative, color = "Total Area"), 
               shape = 20, 
               size = 1.8) +
    geom_point(data=dens, aes(x=Upper, y=cumulative, color = "Sample"), 
               shape = 20, size = 1.8) + 
    geom_line(data=den, aes(x=Upper, y=cumulative, color = "Total Area"), 
              linewidth = 0.5) +
    geom_line(data=dens, aes(x=Upper, y=cumulative, color = "Sample"), 
              linewidth = 0.5) +
    labs(title = name,
         x = "Density", 
         y = "Proportion of area") +
    theme(legend.title = element_blank(),
          legend.position = "inside",
          legend.position.inside = c(0.8,0.2))
  
}
```

```{r}
#| label: inside_3
for (i in 1:length(R4rasters)) {
  thisplot <- pbar_in3[[i]] / pcum_in3[[i]]
  thisplot <- thisplot + plot_annotation(
    title = paste0("Comparison of sample points and total area inside initiation zones for ", R4rasters[[i]][[1]]),
    subtitle = "One sample per 79 square meters, 5-m buffer around points, n = 1825",
    caption = "Bar plots show the proportion of area in each interval, cumulative plots show the cumulative proportion of area")
  plot(thisplot)
}
```

```{r}
#| label: outside_3
for (i in 1:length(R4rasters)) {
  thisplot <- pbar_out3[[i]] / pcum_out3[[i]]
  thisplot <- thisplot + plot_annotation(
    title = paste0("Comparison of sample points and total area outside initiation zones for ", R4rasters[[i]][[1]]),
    subtitle = "Three times more samples outside than inside, n = 5475",
    caption = "Bar plots show the proportion of area in each interval, cumulative plots show the cumulative proportion of area")
  plot(thisplot)
}
```

```{r}

#| label: density_3
for (i in 1:length(R4rasters)) {
  thisplot1 <- density[[i]] / densample3[[i]]
  thisplot <- thisplot1 | denscum3[[i]]
  thisplot <- thisplot + plot_annotation(
    title = paste0("Landslide density for ", R4rasters[[i]][[1]]),
    subtitle = "Total study area (top) vs sampled points (bottom)",
    caption = "Density calculated as the ratio of the number of points inside initiation zones to points outside for each histogram interval")
  plot(thisplot)
}
```

Each increase in the number of sample points produces an improvement in the match of the sample and total-area distributions, particularly for the initiation zones. In these examples, the initiation zones have one third as many points as the total analysis area, but they also cover only 1/1000th of the total analysis area, so the number of points per unit area in the initiation zones is about 300 times higher.

I will use the sample points from the third example to train and validate a machine-learning model to predict landslide initiation probability. I'm using the R package [mlr3](https://mlr3.mlr-org.com/), which simplifies implementation of a machine-learning workflow.

```{r}
library(mgcv)
library(stringr)

# The csv file with the sample points was created with the last run of program SamplePoints.
data_wide <- as.data.table(read.csv(paste0(table, "_wide.csv")))
data_wide <- data_wide[, Class := str_trim(Class)]
data_wide <- data_wide[, Class := as.factor(Class)]
data_wide <- data_wide[, Landform := as.factor(Landform)]
data_wide <- data_wide[, Soils := as.factor(Soils)]
data_wide <- data_wide[, Record := NULL]
save(data_wide, file="data_wide.Rdata")
```
